---
title: "Beaver DOC: analysis plan for the revised submission"
subtitle: "The research questions, the variables that answer them, and the figures that go into the manuscript"
author: "Leonardo Capitani"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-tools: true
    embed-resources: true
    fig-width: 9
    fig-align: center
bibliography: references.bib
csl: ecology-letters.csl
execute:
  echo: true
  message: false
  warning: false
editor: visual
---

```{r setup}
#| include: false
library(tidyverse); library(scales); library(patchwork); library(ggdist); library(knitr); library(here)
invisible(Sys.setlocale("LC_CTYPE", "en_US.UTF-8"))
set.seed(7)
# One colour per meaning, used in every figure: DOC added (source) = skyblue, DOC removed (sink) = coral,
# negligible / no-evidence = lightgrey. Never used for anything else.
PAL <- c(add = "skyblue", remove = "coral", none = "lightgrey")
dir.create(here("results", "figures", "plan"), recursive = TRUE, showWarnings = FALSE)
```

::: {.callout-tip appearance="simple"}
## Purpose

This document fixes the story of the manuscript: three research questions, the variables that answer them, and the figures that go into the main text and into Appendix S1. Everything not listed here is left out. Figure numbers in this document (Figure 1, 2, …) are its own; the manuscript figure each one becomes is written in bold above it.
:::

# 1. The variables

The subject of every effect statement is **beaver-engineering**, the physical modification beavers create, because the upstream–downstream contrast measures the whole beaver-engineered stream. The unit of analysis is the stream.

```{r}
#| label: tbl-variables
#| tbl-cap: "The derived variables used throughout this document, referenced from every figure caption below."
#| echo: false
kable(tribble(
  ~Symbol, ~Name, ~Definition, ~`Meaning (unit)`, ~`Question it answers`,
  "$\\text{DOC}_{up,s}$, $\\text{DOC}_{down,s}$", "upstream DOC, downstream DOC", "measured",
  "DOC concentration entering / leaving the beaver-engineered stream in season $s$ ($w$ = winter, $s$ = summer) **(mg L$^{-1}$)**",
  "What is the DOC concentration before and after the beaver-engineered stream in each season?",
  "$\\Delta\\text{DOC}_s$", "local response", "$\\text{DOC}_{down,s} - \\text{DOC}_{up,s}$",
  "the change in DOC that beaver-engineering produces in season $s$; positive = enrichment, negative = depletion **(mg L$^{-1}$)**",
  "Does beaver-engineering increase or decrease DOC, and by how much, in each season?",
  "$S$", "inherited seasonal contrast", "$\\text{DOC}_{up,summer} - \\text{DOC}_{up,winter}$",
  "the winter-to-summer change of upstream DOC, i.e. what the upstream river corridor delivers to the stream **(mg L$^{-1}$)**",
  "How much does DOC change between winter and summer in the river before it reaches the beaver-engineered stream?",
  "$R_s$", "relative response", "$\\Delta\\text{DOC}_s / \\lvert S \\rvert$",
  "how much DOC beaver-engineering adds ($R_s > 0$) or removes ($R_s < 0$) in season $s$, as a fraction of the inherited seasonal contrast **(dimensionless)**",
  "How much does beaver-engineering add or remove DOC, relative to the inherited seasonal contrast?",
  "$B$", "beaver seasonal modification", "$\\Delta\\text{DOC}_{summer} - \\Delta\\text{DOC}_{winter}$",
  "how much beaver-engineering changes the winter-to-summer DOC contrast **(mg L$^{-1}$)**",
  "Does beaver-engineering change the seasonal DOC contrast, beyond its effect within each season?",
  "$G$", "seasonal gain", "$B / S$",
  "$G > 0$: beaver-engineering **reinforces** the inherited seasonal contrast; $-1 < G < 0$: **dampens** it; $G < -1$: **reverses** it **(dimensionless)**",
  "Does beaver-engineering reinforce, dampen or reverse the inherited seasonal contrast?",
  "$A_\\text{beaver}$, $A_\\text{floodplain}$", "areas", "mapped",
  "ponded footprint of the beaver-engineered stream; contributing catchment floodplain area **(km$^2$)**",
  "How large is the beaver-engineered footprint, and how large is the floodplain that generates $S$?",
  "$F$", "footprint contrast", "$A_\\text{floodplain} / A_\\text{beaver}$",
  "how much larger the floodplain area is than the beaver-engineered footprint **(dimensionless)**",
  "How small is the beaver-engineered footprint relative to the floodplain that generates $S$?"))
```

**Manuscript placement: Appendix S1 (Figure S10).**

![**Derivation of the variables.** Top: the same causal structure operates in winter and summer; downstream DOC inherits upstream DOC and is modified by local hydrological and biological processes. Bottom (dashed): $\Delta\text{DOC}_s$, $S$, $R_s$ and $F$ are calculations from the concentrations, not causal nodes. Solid arrows are causal pathways; the green arrow is the persistence of the hydrologic template between seasons. Variable definitions: @tbl-variables.](../images/DAG_variable_derivation.png){#fig-dag}

# 2. Research questions and hypotheses

**Q1 — Direction.** *Does beaver-engineering consistently increase or decrease DOC downstream compared with upstream?*
Hypothesis: if a stream is impounded by beaver dams, then the local response $\Delta\text{DOC}_s$ will not have a consistent sign across streams, because impoundment simultaneously starts DOC-producing processes — exudation and senescence of phytoplankton and macrophytes, leaching of litter and inundated soils [@søndergaard1985; @sell1992; @aitkenhead1999] — and DOC-removing processes — microbial mineralization, photodegradation and sedimentation during the longer residence time [@johnston2017; @nummi2018] — whose balance depends on the stream; syntheses accordingly report enrichment, no change and depletion [@larsen2021; @grudzinski2022]. We further expect the balance to tilt toward enrichment in summer, when production exceeds removal [@mulholland1997].

**Q2 — Drivers.** *Which abiotic and biotic drivers set the direction of the local response?*
Hypothesis: if the number of dams increases, then water residence time increases because each dam impounds water and slows flow [@westbrook2006; @majerova2015]; if water residence time and solar radiation increase, then phytoplankton and macrophyte abundance increase because nutrients are retained and light is available for photosynthesis [@reynolds2006; @finger2007; @feuchtmayr2009]; if primary producer abundance and soil litter cover increase, then $\Delta\text{DOC}_s$ increases because exudation, senescence and litter leaching release DOC [@reitsema2018; @gessner1999]; if upstream DOC increases, then $\Delta\text{DOC}_s$ decreases because a larger incoming load is proportionally retained or mineralized within the pond [@larson2007; @wohl2013]. The thirteen path-specific hypotheses are in Appendix S1, Table S1.

**Q3 — Magnitude and seasonal modification.** *How much does beaver-engineering add or remove DOC relative to the inherited seasonal contrast, and does it reinforce, dampen or reverse that contrast?*
Hypothesis, magnitude: if the season is summer, then the relative response $R_{summer}$ will be positive and a substantial fraction of the inherited seasonal contrast, because higher light and temperature raise autochthonous DOC production inside the pond independently of the upstream corridor [@mulholland1997; @battin2023]; if the season is winter, then $R_{winter}$ will be close to zero because primary production is minimal.
Hypothesis, modification: if the summer response exceeds the winter response ($B > 0$ where $S > 0$), then the seasonal gain $G$ will be positive and beaver-engineering will reinforce the inherited seasonal contrast; because the balance of DOC-producing and DOC-removing processes is stream-specific (Q1), we expect $G$ to be heterogeneous across streams, with reinforcement, dampening and reversal all present.
Because $A_\text{beaver}$ is a very small part of $A_\text{floodplain}$ (footprint contrast $F \gg 1$), both responses arise on a footprint far smaller than the corridor that generates $S$.

**Q4 — Intensity.** *Does the local response scale with the intensity of beaver-engineering — the number of dams and the water residence time — or is it organised by something else?*
Hypothesis: if the number of dams and the water residence time increase, then $R_{summer}$ and $G$ will become more positive, because longer impoundment retains nutrients and creates habitat for the primary producers that add DOC in summer [@westbrook2006; @majerova2015; @feuchtmayr2009]. Because Q2 found only weak paths from the dams through water residence time, we expect at most a weak ordering of the streams along these gradients. Q4 is descriptive and belongs in Appendix S1; it answers the question a reader asks after Figure 2b: "are the streams on the right the ones with more dams?"

The division of labour between the two quantities is fixed: **$R_s$ answers "how much, in season $s$" (Figure 2a, b); $G$ answers "what does that do to the seasonal contrast itself" (Figure 2c)**.

| Question | Answered by | This document | Manuscript |
|---|---|---|---|
| Q1 Direction | $\Delta\text{DOC}_s$ | @fig-q1 | Figure 1a (main text) |
| Q2 Drivers | structural causal model | @fig-q2 | Figure 1b (main text) |
| Q3 Magnitude and seasonal modification | $R_s$ (figure), $G$ (text), $F$ as context | @fig-R | Figure 2 (main text) |
| Q4 Intensity | $R_{summer}$ and $G$ against the number of dams and the water residence time (Bayesian slope) | @fig-intensity | Figure S12 (Appendix S1) |
| Spatial coherence (Discussion support) | $S$, $B$, downstream seasonal contrast $T$ | @fig-spatial | Figure S11 (Appendix S1) |
| Model validation (Methods support) | leave-one-out and hold-out prediction of downstream DOC and $\Delta\text{DOC}$ | @fig-holdout | Figure S13 (Appendix S1) |
| Methods | sampling design; $S$, $\Delta\text{DOC}_s$, $R_s$, $F$ | @fig-map | Figure 3 (main text) |
| Derivation of the variables | — | @fig-dag | Figure S10 (Appendix S1) |

# 3. Q1 — Direction

**Manuscript placement: main text, Figure 1a (existing, unchanged).**

![**Direction of the local response.** Posterior predictive distribution of $\Delta\text{DOC}_s$ for a single sampling of a beaver-engineered stream, in winter and summer, from the measurement-error model on the raw upstream and downstream concentrations. Coral: posterior predictive probability of a sink response ($\Delta\text{DOC}_s < -0.5$ mg L$^{-1}$); grey: negligible response (ROPE $\pm 0.5$ mg L$^{-1}$); skyblue: source response ($\Delta\text{DOC}_s > 0.5$ mg L$^{-1}$). Point: median; bars: 66% and 95% intervals. $\Delta\text{DOC}_s$ is defined in @tbl-variables.](../overleaf/figures/media/Figure_2a.png){#fig-q1}

```{r q1-data}
# raw data: one row per sampling; pair each stream's winter and summer sampling
m <- read.csv(here("data", "processed", "m12.csv"))
w <- m |> filter(season == "winter") |> select(site, beaver_territory, up_w = DOC_input, dn_w = DOC_out,
                                                 Afp = A_floodplain_km2, Abv = A_beaver_km2)
s <- m |> filter(season == "summer") |> select(site, beaver_territory, up_s = DOC_input, dn_s = DOC_out)
# the 158 streams with both seasons complete; local responses and footprint contrast
d <- inner_join(w, s, by = c("site", "beaver_territory")) |> filter(complete.cases(up_w, dn_w, up_s, dn_s)) |>
  mutate(dw = dn_w - up_w,      # local response in winter
         ds = dn_s - up_s,      # local response in summer
         F  = Afp / Abv)        # footprint contrast
# shares of enrichment by season, and streams that switch direction between seasons
sw <- d |> summarise(n = n(), winter_enrich = mean(dw > 0), summer_enrich = mean(ds > 0),
                     dep_to_enr = sum(dw < 0 & ds > 0), enr_to_dep = sum(dw > 0 & ds < 0))
```

::: {.callout-note}
## Key message of @fig-q1

@fig-q1 shows, for each season, the distribution of the local response one would measure by visiting one beaver-engineered stream once (the posterior predictive distribution of $\Delta\text{DOC}_s$ for a single sampling). Averaged over all streams the local response is negligible: +0.14 mg L$^{-1}$ in summer and 0.00 mg L$^{-1}$ in winter, both inside the ROPE of $\pm$0.5 mg L$^{-1}$. A single sampling, however, is highly variable. In winter, 77% of samplings are negligible, 11% show a source response and 12% a sink response; in summer, 75% are negligible, 15% source and 9% sink. The important message: beaver-engineering does **not** change DOC in a consistent direction — a single visit most likely finds no ecologically relevant change, and when it does, in summer a source response is more likely than a sink response.

Sentence to add to the Results: of the `r sw$n` streams sampled in both seasons, `r sprintf("%.0f%%", 100*sw$winter_enrich)` show enrichment in winter and `r sprintf("%.0f%%", 100*sw$summer_enrich)` in summer; `r sw$dep_to_enr` streams switch from winter depletion to summer enrichment, `r sw$enr_to_dep` the other way.
:::

# 4. Q2 — Drivers

**Manuscript placement: main text, Figure 1b (redrawn: the outcome is downstream DOC).**

The chunk below draws the structural causal model as it will appear in the manuscript. The layout is fixed by hand (node positions and edge routing); the numbers are the posterior mean effects of model `b1_downstream`, and each edge's line type encodes the posterior probability of its direction.

```{r fig-q2}
#| fig-height: 5.6
#| fig-cap: "**Drivers of downstream DOC.** Structural causal model with downstream DOC as the outcome (model `b1_downstream`) and the posterior mean standardized effects; the upstream-DOC path is the inheritance slope (0.95, mg per mg). Skyblue: positive and coral: negative effects with posterior probability above 0.90; dashed grey: no-evidence."
# ---- node positions (x, y on a 0-10 canvas) --------------------------------
nodes <- data.frame(
  name = c("Max dam\nheight", "Number of\ndams", "Channel\ngradient", "Water\nresidence time",
           "Solar\nradiation", "Macrophytes", "Phytoplankton", "Soil litter\ncover",
           "Upstream\nDOC", "Downstream\nDOC"),
  x = c(0.9, 0.9, 2.9, 3.2, 4.6, 6.6, 6.2, 6.2, 8.8, 8.8),
  y = c(6.6, 4.6, 2.0, 5.0, 9.3, 7.2, 5.0, 2.9, 8.8, 5.4))
rownames(nodes) <- c("dh", "nd", "gr", "wrt", "sol", "mac", "pla", "lit", "up", "out")
# ---- edges: posterior mean effect (label), class by P > 0.90, curvature ----
edges <- read.csv(text = 'from,to,label,class,curv,lx,ly
dh,wrt,,none,0,0,0
nd,wrt,0.06,pos,0,-0.1,0.35
gr,wrt,-0.08,neg,0,0.45,0
wrt,pla,0.11,pos,0,0,0.30
wrt,mac,,none,0,0,0
wrt,out,,none,0.18,0,0
sol,mac,0.14,pos,-0.15,0.55,0.55
sol,pla,0.19,pos,0.30,-0.85,0.10
sol,lit,,none,0.30,0,0
mac,out,0.08,pos,0,0,0.32
pla,out,0.14,pos,0,0,0.30
lit,out,0.07,pos,0,-0.2,0.32
up,out,0.95,pos,0,0.45,0')
# shorten each arrow so it starts/ends at the box border, not the centre
a <- merge(edges, nodes, by.x = "from", by.y = "row.names")
a <- merge(a, nodes, by.x = "to", by.y = "row.names", suffixes = c("", "2"))
dx <- a$x2 - a$x; dy <- a$y2 - a$y; L <- sqrt(dx^2 + dy^2)
a$x0 <- a$x + dx / L * 0.85; a$y0 <- a$y + dy / L * 0.55
a$x1 <- a$x2 - dx / L * 0.90; a$y1 <- a$y2 - dy / L * 0.55
a$xm <- (a$x0 + a$x1) / 2 + a$lx; a$ym <- (a$y0 + a$y1) / 2 + a$ly
pal2 <- c(pos = "#3B9AD1", neg = "#E8603C", none = "grey55")   # skyblue / coral / grey
lty2 <- c(pos = "solid",   neg = "solid",   none = "dashed")
arr <- arrow(length = unit(0.16, "cm"), type = "closed")
pq2 <- ggplot()
st <- subset(a, curv == 0)   # straight and curved arrows drawn separately
pq2 <- pq2 + geom_segment(data = st, aes(x = x0, y = y0, xend = x1, yend = y1),
                          colour = pal2[st$class], linetype = lty2[st$class], linewidth = 0.7, arrow = arr)
for (i in which(a$curv != 0))
  pq2 <- pq2 + geom_curve(data = a[i, ], aes(x = x0, y = y0, xend = x1, yend = y1),
                          curvature = a$curv[i], colour = pal2[a$class[i]],
                          linetype = lty2[a$class[i]], linewidth = 0.7, arrow = arr)
pq2 <- pq2 +
  geom_text(data = subset(a, label != ""), aes(x = xm, y = ym, label = label), size = 4.2) +
  geom_label(data = nodes[rownames(nodes) != "out", ], aes(x = x, y = y, label = name),
             size = 4.1, label.padding = unit(0.3, "lines"), label.r = unit(0.05, "lines")) +
  geom_label(data = nodes["out", ], aes(x = x, y = y, label = name), size = 4.4,
             fontface = "bold", label.padding = unit(0.45, "lines"),
             label.r = unit(0.9, "lines"), linewidth = 0.9, colour = "#C0208E") +
  annotate("text", x = 0.15, y = 9.8, label = "(b)", size = 6.5, hjust = 0) +
  xlim(0, 10) + ylim(1.2, 10) + theme_void()
ggsave(here("images", "Figure_1b_downstream.png"), pq2, width = 9, height = 5.6, dpi = 300, bg = "white")
pq2
```

::: {.callout-note}
## Key message of @fig-q2

Downstream DOC is mostly the upstream state carried forward: the inheritance slope is 0.95 [0.92, 0.98] mg per mg, by far the largest path in the model. Around that inherited background, beaver-engineering adds a local modification through the primary producers and the litter: phytoplankton (+0.14) and macrophytes (+0.08), fed by solar radiation (+0.19 and +0.14), and soil litter cover (+0.07). Water residence time and the number/height of the dams act only weakly, and only indirectly through the producers. Beaver-engineering does not retain incoming DOC — it inherits it, then adds to it. Supporting material stays in Appendix S1 (Table S1, Figures S1–S9).
:::

**Parameterization.** The structural causal model is reported with **downstream DOC as the outcome**: upstream DOC enters as the inherited incoming state with a freely estimated inheritance slope, and $\Delta\text{DOC}_s$ is computed from the posterior afterwards as $\text{DOC}_{down,s} - \text{DOC}_{up,s}$, so upstream DOC is never both a predictor and part of the outcome. This reading replaces the older wording of Q2, which described the upstream-DOC path as upstream DOC *causing greater retention*; downstream DOC instead **inherits** the upstream state (slope 0.95 mg per mg), and the hydrological/biological pathways are local modifications around that inherited background, not retention of it. This model (`b1_downstream`, `notebook/scm_rebuild.qmd`) is the $\Delta$DOC model in a different parameterization: every mediator effect agrees to the second decimal, and the upstream-DOC path is re-read as an inheritance slope of **0.95 [0.92, 0.98]** — downstream DOC inherits 95% of the upstream concentration, mg per mg. Figure 1b is redrawn accordingly (@fig-q2); the Methods gain one sentence stating the parameterization and the inheritance slope.

**Robustness check (Josh's suggestion, adopted): does fixing the inheritance slope at 1 change anything else?** `b1_downstream` estimates the slope freely (prior Normal(1, 0.3), posterior 0.95). A second fit, `b1_downstream_fixed` (`notebook/scm_downstream_fixed_fit.R`), pins it at 1 (prior Normal(1, 0.001), the standard way to fix an `mi()` coefficient in brms) and is otherwise identical.

```{r robustness-inheritance}
#| echo: false
fx_path <- here("results", "models_fit", "b1_downstream_fixed.rds")
if (file.exists(fx_path)) {
  library(brms)
  d_free <- as_draws_df(readRDS(here("results", "models_fit", "b1_downstream.rds")))
  d_fix  <- as_draws_df(readRDS(fx_path))
  qf2 <- function(x) sprintf("%.3f [%.3f, %.3f]", mean(x), quantile(x, .025), quantile(x, .975))
  rob <- tribble(
    ~Parameter, ~`Free slope (b1_downstream)`, ~`Fixed at 1 (b1_downstream_fixed)`,
    "Inheritance slope",  qf2(d_free$bsp_DOCout_miDOC_input),      qf2(d_fix$bsp_DOCout_miDOC_input),
    "Macrophytes",        qf2(d_free$bsp_DOCout_mimacrophy_abun_z), qf2(d_fix$bsp_DOCout_mimacrophy_abun_z),
    "Phytoplankton",      qf2(d_free$bsp_DOCout_miplankton_abun_z), qf2(d_fix$bsp_DOCout_miplankton_abun_z),
    "Soil litter cover",  qf2(d_free$bsp_DOCout_micover_litter_z),  qf2(d_fix$bsp_DOCout_micover_litter_z),
    "Water residence time", qf2(d_free$bsp_DOCout_miwater_res_time_z), qf2(d_fix$bsp_DOCout_miwater_res_time_z),
    "Intercept",          qf2(d_free$b_DOCout_Intercept),           qf2(d_fix$b_DOCout_Intercept))
  kable(rob, caption = "Posterior mean [95% CrI]. Fixing the inheritance slope at 1 is a robustness check, not a preferred model.")
} else {
  cat("Fit in progress (notebook/scm_downstream_fixed_fit.R) -- table will appear once results/models_fit/b1_downstream_fixed.rds exists.")
}
```

**Result: fixing the slope changes nothing else.** Both fits converge cleanly (max R-hat 1.00). Every mediator effect (macrophytes, phytoplankton, litter, water residence time) agrees within its credible interval whether the inheritance slope is estimated freely (0.945) or pinned at 1 — the largest shift is Phytoplankton, 0.141 to 0.130, well inside both intervals. Only the intercept moves (0.15 to 0.02), which is expected: fixing the slope at exactly 1 forces the intercept to absorb the small remaining offset. The mediator story of Figure 1b does not depend on which value the inheritance slope takes.

# 5. Q3 — Magnitude and seasonal modification: $R_s$ and $G$

The analysis uses the **158 streams with both a winter and a summer sampling**.

$R_s$, $B$ and $G$ are computed from posterior draws of each stream's concentrations (Bayesian measurement-error model, `notebook/concentrations_fit.qmd`; the model belongs in the Methods), so no derived quantity is ever formed from raw values and the uncertainty is carried through (`r nrow(readRDS(here("results","concentrations_thin.rds"))$S)` posterior draws per stream).

```{r q3-compute}
# LR holds one matrix per quantity: rows = 4000 posterior draws, columns = 158 streams
LR <- readRDS(here("results", "concentrations_thin.rds"))
R_s <- LR$R_s   # relative response in summer, per draw and stream
R_w <- LR$R_w   # relative response in winter
S_m <- LR$S     # inherited seasonal contrast
# medR: for each posterior draw, the median across streams -> a posterior
# distribution of the across-stream median (optionally on a subset of streams)
medR <- function(R, keep = rep(TRUE, ncol(R))) apply(R[, keep, drop = FALSE], 1, median)
# per-stream median |S|, used only for the small-denominator stability check
absS <- apply(abs(S_m), 2, median)
# qf: format a posterior as "median [95% credible interval]"
qf   <- function(x) sprintf("%.2f [%.2f, %.2f]", median(x), quantile(x, .025), quantile(x, .975))
q3 <- tibble(
  Quantity = c("Median $R_{summer}$ across streams", "Median $R_{winter}$ across streams", "$R_{summer} - R_{winter}$",
               "Median $R_{summer}$, excluding the 20% smallest $\\lvert S \\rvert$ (ratio-stability check)"),
  `Posterior median [95% CrI]` = c(qf(medR(R_s)), qf(medR(R_w)), qf(medR(R_s) - medR(R_w)), qf(medR(R_s, absS > quantile(absS, .2)))),
  `P(> 0)` = sprintf("%.3f", c(mean(medR(R_s) > 0), mean(medR(R_w) > 0), mean(medR(R_s) - medR(R_w) > 0),
                               mean(medR(R_s, absS > quantile(absS, .2)) > 0))))
kable(q3)
# streams whose summer relative response is positive (or negative) with P > 0.90
n_add <- sum(colMeans(R_s > 0) > 0.90); n_rem <- sum(colMeans(R_s < 0) > 0.90); n_str <- ncol(R_s)
# Seasonal gain G = B / S, per stream, from the same posterior draws
# beaver seasonal modification B = summer response - winter response, per draw
B_m <- LR$ds - LR$dw
# seasonal gain G = B / S: does beaver-engineering reinforce (G > 0),
# dampen (-1 < G < 0) or reverse (G < -1) the inherited seasonal contrast?
G_m <- B_m / S_m
g_med <- median(apply(G_m, 1, function(g) median(g)))  # posterior median of the across-stream median gain
```

**Manuscript placement: main text, Figure 2 (new; replaces the amplification-factor figure).**

```{r fig-R}
#| fig-height: 12.5
#| fig-cap: "**Relative response $R_s$ and seasonal gain $G$ of beaver-engineering.** (a) Posterior distribution of the median relative response across the 158 streams, in winter and summer; point: posterior median, bars: 66% and 95% credible intervals; dashed line: no response. (b) Relative response of each stream in summer: posterior median with 66% (thick) and 95% (thin) credible intervals, streams ranked by their median; the grey band marks $\\lvert R_{summer} \\rvert < 0.1$. The vertical axis is linear between $-0.1$ and $0.1$ and logarithmic beyond, so small and large responses are both readable. In panels (a) and (b) skyblue: source (beaver-engineering adds DOC, P > 0.90), coral: sink (removes DOC, P > 0.90), grey: no-evidence — the same colours as in the direction figure; panel (b) prints the percentage of the 158 streams in each colour directly on the plot. (c) Out of the 158 streams, what percentage reinforce ($G > 0$, make the inherited seasonal contrast bigger), dampen ($-1 < G < 0$, make it smaller) or reverse ($G < -1$, flip its sign)? Each stream is classified on every posterior draw, so each bar is a full posterior distribution of that percentage, with its own uncertainty — not a single count of streams sorted into fixed bins; point: posterior median percentage, bars: 66% and 95% credible intervals. $R_s$, $G$, $B$ and $S$ are defined in @tbl-variables."
# classify by posterior probability: add if P(R > 0) > 0.90, remove if P(R < 0) > 0.90, else no-evidence
cls <- function(p_add) case_when(p_add > 0.90 ~ "add", p_add < 0.10 ~ "remove", TRUE ~ "none")
lab <- c(add = "source (adds DOC)", remove = "sink (removes DOC)", none = "no-evidence")   # P > 0.90 stated once, in the caption

# ---- panel (a): across-stream median, winter and summer --------------------
pop <- bind_rows(tibble(season = "summer", v = medR(R_s)), tibble(season = "winter", v = medR(R_w))) |>
  mutate(season = factor(season, levels = c("winter", "summer"))) |>
  group_by(season) |> mutate(class = cls(mean(v > 0))) |> ungroup()
pa <- ggplot(pop, aes(x = v, y = season, fill = class)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
  stat_halfeye(.width = c(.66, .95), point_interval = "median_qi", alpha = .9) +  # 66% and 95% intervals
  scale_fill_manual(values = PAL, labels = lab, breaks = names(lab), name = NULL, drop = FALSE) +
  labs(x = expression("Median relative response across streams,  "*R[s]*" = "*Delta*DOC[s]*" / |S|"), y = NULL) +
  theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank(), legend.position = "top")

# ---- panel (b): each stream's summer relative response, ranked -------------
# per stream: posterior median, 66% and 95% credible intervals, and its class
per <- tibble(med  = apply(R_s, 2, median),
              lo95 = apply(R_s, 2, quantile, .025), hi95 = apply(R_s, 2, quantile, .975),
              lo66 = apply(R_s, 2, quantile, .17),  hi66 = apply(R_s, 2, quantile, .83),
              class = cls(colMeans(R_s > 0))) |> arrange(med) |> mutate(rank = row_number())
# percentage of the 158 streams in each colour: the legend names the colour,
# the in-panel text gives its share (top-left corner is empty of points)
pct_b <- prop.table(table(factor(per$class, levels = c("add", "remove", "none")))) * 100
pct_lab <- tibble(class = c("add", "remove", "none"),
                   txt = sprintf("%.0f%% of streams", pct_b[c("add", "remove", "none")]),
                   y = c(90, 30, 10))   # stacked top to bottom on the log axis
# axis transform: linear inside [-0.1, 0.1], logarithmic beyond
psl <- pseudo_log_trans(sigma = 0.1, base = 10)
pb <- ggplot(per, aes(x = rank, y = med, colour = class)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = -0.1, ymax = 0.1, fill = "grey88", alpha = .8) +
  geom_hline(yintercept = 0, colour = "grey45") +
  geom_linerange(aes(ymin = lo95, ymax = hi95), alpha = .45, linewidth = 0.4) +  # 95% interval (thin)
  geom_linerange(aes(ymin = lo66, ymax = hi66), alpha = .8, linewidth = 0.9) +   # 66% interval (thick)
  geom_point(size = 1.6) +
  geom_text(data = pct_lab, aes(x = 3, y = y, label = txt, colour = class), hjust = 0, size = 4, fontface = "bold", show.legend = FALSE) +
  scale_colour_manual(values = c(add = "#3B9AD1", remove = "#E8603C", none = "grey60"), labels = lab, breaks = names(lab), name = NULL) +
  scale_y_continuous(trans = psl, breaks = c(-10, -1, -0.1, 0, 0.1, 1, 10)) +
  labs(x = "Stream", y = expression(R[summer]*" = "*Delta*DOC[summer]*" / |S|")) +
  theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank(), legend.position = "top")
# ---- panel (c): posterior share of streams that reinforce/dampen/reverse ---
# classify every stream on EVERY posterior draw (no P > 0.90 threshold), then
# compute the proportion in each zone per draw -- this gives a posterior
# distribution of the proportion itself, propagating its own uncertainty,
# rather than a single count built from per-stream point classifications.
zone <- function(g) case_when(g > 0 ~ "reinforce", g > -1 ~ "dampen", TRUE ~ "reverse")
zmat <- apply(G_m, c(1, 2), zone)                       # same shape as G_m, one label per draw x stream
gprop <- as_tibble(t(apply(zmat, 1, function(row)        # proportion per zone, one row per draw
  c(reinforce = mean(row == "reinforce"), dampen = mean(row == "dampen"), reverse = mean(row == "reverse"))))) |>
  pivot_longer(everything(), names_to = "zone", values_to = "p") |>
  mutate(zone = factor(zone, levels = c("reverse", "dampen", "reinforce")))
gpal <- c(reinforce = "#8856A7", dampen = "#41AB5D", reverse = "#B5651D")   # light brown
pc <- ggplot(gprop, aes(x = p, y = zone, fill = zone)) +
  stat_halfeye(.width = c(.66, .95), point_interval = "median_qi") +   # 66% and 95% credible intervals
  scale_fill_manual(values = gpal, guide = "none") +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  labs(x = "Percentage of the 158 streams", y = expression("Seasonal gain,  "*G*" = "*B*" / "*S)) +
  theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank())
fig2 <- pa / pb / pc + plot_layout(heights = c(1, 1.3, 1)) + plot_annotation(tag_levels = "a")
ggsave(here("results", "figures", "plan", "fig_R.png"), fig2, width = 9, height = 11, dpi = 300)
fig2
# summary of panel (c) for the key message, one row per zone
gsum <- gprop |> group_by(zone) |> summarise(med = median(p), lo = quantile(p, .025), hi = quantile(p, .975))
gz <- function(z) gsum |> filter(zone == z)

# same +/-0.1 ROPE as panel (b)'s grey band, applied to R_summer, to compare
# against @fig-q1's ROPE logic (see the note below)
zone_r <- function(r) case_when(r > 0.1 ~ "source", r < -0.1 ~ "sink", TRUE ~ "negligible")
rzmat <- apply(R_s, c(1, 2), zone_r)
rprop <- as_tibble(t(apply(rzmat, 1, function(row)
  c(source = mean(row == "source"), negligible = mean(row == "negligible"), sink = mean(row == "sink"))))) |>
  pivot_longer(everything(), names_to = "zone", values_to = "p")
rz <- function(z) rprop |> filter(zone == z) |> summarise(med = median(p))
```

::: {.callout-important}
## Key message of @fig-R

**Panel (a) of @fig-R.** In winter, beaver-engineering neither adds nor removes DOC: the relative response is `r sprintf("%.2f", median(medR(R_w)))`. In summer it adds DOC: averaged across the 158 streams, the DOC added by beaver-engineering equals `r sprintf("%.0f%%", 100*median(medR(R_s)))` of the inherited seasonal contrast $S$ (95% CrI `r sprintf("%.0f–%.0f%%", 100*quantile(medR(R_s), .025), 100*quantile(medR(R_s), .975))`) — about one fifth of the DOC change that the seasons themselves produce in the upstream river.

**Panel (b) of @fig-R.** The streams do not behave alike. In summer, `r sprintf("%.0f%%", 100 * n_add / n_str)` of the beaver-engineered streams (`r n_add` of `r n_str`) are DOC sources; `r sprintf("%.0f%%", 100 * n_rem / n_str)` (`r n_rem`) are DOC sinks; the rest are no-evidence.

**Panel (c) of @fig-R.** The seasonal gain is heterogeneous: `r sprintf("%.0f%%", 100*gz("reinforce")$med)` of the streams (95% CrI `r sprintf("%.0f–%.0f%%", 100*gz("reinforce")$lo, 100*gz("reinforce")$hi)`) reinforce the inherited seasonal contrast, `r sprintf("%.0f%%", 100*gz("dampen")$med)` (`r sprintf("%.0f–%.0f%%", 100*gz("dampen")$lo, 100*gz("dampen")$hi)`) dampen it, and `r sprintf("%.0f%%", 100*gz("reverse")$med)` (`r sprintf("%.0f–%.0f%%", 100*gz("reverse")$lo, 100*gz("reverse")$hi)`) reverse it; the median gain across streams is `r sprintf("%.2f", g_med)`. Beaver-engineering does not shift the seasonal contrast in one direction; it redistributes it stream by stream.
:::

::: {.callout-note}
## How @fig-q1 and @fig-R fit together

@fig-q1 and @fig-R are not two conflicting results — they answer two different questions, so one naturally reads as "mostly nothing detectable" and the other as "most streams have a clear sign."

- **@fig-q1 asks: if someone visits a stream and takes one water sample, what would they most likely find?** This is a single field visit — the way almost every published beaver–DOC study, including the ones this manuscript is compared against, actually samples. The answer: mostly nothing an ecologist would call a real change.
- **@fig-R asks: what is each stream's true relative response, according to the best estimate the model can produce from everything it knows?** This is the paper's actual result — it does not simulate a single field visit, it reports the model's estimate of the real value. The answer: most streams have a clear sign.

Both questions are legitimate and both answers are correct; they are simply not the same question. (The technical reason the two differ — one reports a simulated single observation, the other reports the model's own estimate — is a Methods-level detail and does not need to be repeated at every figure.) For consistency, applying @fig-q1's "no real change" logic to $R_{summer}$ (a $\pm$0.1 band, matching the grey band already drawn in panel (b)) still leaves only `r sprintf("%.0f%%", 100*rz("negligible")$med)` of streams with no detectable response, versus `r sprintf("%.0f%%", 100*rz("source")$med)` source and `r sprintf("%.0f%%", 100*rz("sink")$med)` sink.

Methods sentence: state once that Figure 1a reports the posterior *predictive* distribution of a single new sampling (so it includes the instrument error), while $R_s$, $B$ and $G$ are computed from each stream's posterior *parameter* estimate (the model's estimate of the true concentration).
:::

# 6. Q4 — Intensity: is the response ordered by the number of dams or the water residence time?

Figure 2b ranks the streams by their own response. Here the same per-stream posteriors of $R_{summer}$ and $G$ are placed on the two gradients of beaver-engineering intensity we measured, and the ordering is estimated with a Bayesian model rather than a correlation coefficient: for each response and each gradient, a Student-*t* regression of the response (on the same pseudo-log scale as the figure axis, $t(x) = \operatorname{asinh}(x/0.2)/\ln 10$) on the standardized gradient, in which each stream enters with its posterior median and its posterior standard deviation as a known measurement error (`se()` in brms). The slope $\beta$ is the ordering: $P(\beta > 0)$ is the posterior probability that the response increases along the gradient.

**Manuscript placement: Appendix S1 (Figure S12), with one sentence in the Results.**

```{r fig-intensity}
#| fig-height: 8
#| fig-cap: "**Relative response and seasonal gain against the intensity of beaver-engineering.** Each point is a stream: posterior median of $R_{summer}$ (top row) or $G$ (bottom row) with 66% (thick) and 95% (thin) credible intervals, against the number of dams (left; jittered) and the summer water residence time (right; log axis). Colours as in Figure 2: source / sink / no-evidence for $R_{summer}$; reinforce / dampen / reverse / no-evidence for $G$ (classified at posterior probability 0.90). The vertical axes are linear between −0.1 and 0.1 and logarithmic beyond. Each panel title gives the slope $\\beta$ of a Bayesian Student-t regression of the response (pseudo-log scale) on the standardized gradient, with each stream's posterior standard deviation as its measurement error: posterior median [95% credible interval] and $P(\\beta > 0)$. $R_s$, $G$, $B$ and $S$ are defined in @tbl-variables."
library(brms)
# covariates per analysed stream: number of dams and summer water residence time (from m6)
cov <- read.csv(here("data", "processed", "m6.csv")) |> filter(season == "summer") |>
  select(site, beaver_territory, n_dams, wrt = water_res_time)
cv <- LR$d |> select(site, beaver_territory) |> left_join(cov, by = c("site", "beaver_territory"))
# per-stream classes for colouring (same rules as Figure 2)
r_class <- cls(colMeans(R_s > 0))
g_class <- apply(G_m, 2, function(g) {
  p <- c(reinforce = mean(g > 0), dampen = mean(g > -1 & g < 0), reverse = mean(g < -1))
  if (max(p) > 0.90) names(p)[which.max(p)] else "no-evidence" })
# the pseudo-log transform used on the figure axis (scales::pseudo_log_trans, sigma = 0.1)
plog <- function(x) asinh(x / 0.2) / log(10)
# Bayesian ordering: Student-t regression of the transformed response on the standardized
# gradient, each stream weighted by its own posterior uncertainty (known se)
slope_fit <- function(M, x, tag) {
  ok <- !is.na(x); TM <- plog(M[, ok])
  dd <- tibble(y = apply(TM, 2, median), se_y = apply(TM, 2, sd), x = as.numeric(scale(x[ok])))
  brm(bf(y | se(se_y, sigma = TRUE) ~ 1 + x), data = dd, family = student(),
      prior = c(prior(normal(0, 1), class = "b"), prior(normal(0, 2), class = "Intercept"),
                prior(student_t(3, 0, 1), class = "sigma"), prior(gamma(2, 0.1), class = "nu")),
      chains = 4, iter = 3000, warmup = 1000, seed = 7, backend = "cmdstanr", refresh = 0,
      file = here("results", "models_fit", paste0("intensity_", tag)))
}
fits <- list(R_dams = slope_fit(R_s, cv$n_dams, "R_dams"), R_wrt = slope_fit(R_s, log10(cv$wrt), "R_wrt"),
             G_dams = slope_fit(G_m, cv$n_dams, "G_dams"), G_wrt = slope_fit(G_m, log10(cv$wrt), "G_wrt"))
beta <- lapply(fits, function(f) as_draws_df(f)$b_x)
b_lab <- function(b) sprintf("beta = %.2f [%.2f, %.2f], P(beta > 0) = %.2f", median(b), quantile(b, .025), quantile(b, .975), mean(b > 0))
qs <- function(M) tibble(med = apply(M, 2, median), lo95 = apply(M, 2, quantile, .025), hi95 = apply(M, 2, quantile, .975),
                         lo66 = apply(M, 2, quantile, .17), hi66 = apply(M, 2, quantile, .83))
rpal <- c(add = "#3B9AD1", remove = "#E8603C", none = "grey60")
rlab <- c(add = "source", remove = "sink", none = "no-evidence")
gpal2 <- c(reinforce = "#8856A7", dampen = "#41AB5D", reverse = "#B5651D", `no-evidence` = "grey60")
panel <- function(M, x, xlab, class, pal, labs_, ylab, sub, logx = FALSE) {
  d <- qs(M) |> mutate(x = x, class = factor(class, levels = names(pal)))
  pos <- if (logx) "identity" else position_jitter(width = .15, height = 0, seed = 1)
  p <- ggplot(d, aes(x = x, y = med, colour = class)) +
    geom_hline(yintercept = 0, colour = "grey45") +
    geom_linerange(aes(ymin = lo95, ymax = hi95), alpha = .45, linewidth = 0.4, position = pos) +
    geom_linerange(aes(ymin = lo66, ymax = hi66), alpha = .8, linewidth = 0.9, position = pos) +
    geom_point(size = 1.6, position = pos) +
    scale_colour_manual(values = pal, labels = labs_, breaks = names(pal), name = NULL) +
    scale_y_continuous(trans = psl, breaks = c(-10, -1, -0.1, 0, 0.1, 1, 10)) +
    labs(x = xlab, y = ylab, subtitle = sub) + theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank(), legend.position = "top", plot.subtitle = element_text(size = 10))
  if (logx) p + scale_x_log10() else p
}
p1 <- panel(R_s, cv$n_dams, "Number of dams", r_class, rpal, rlab, expression(R[summer]), b_lab(beta$R_dams))
p2 <- panel(R_s, cv$wrt, "Water residence time in summer (s)", r_class, rpal, rlab, expression(R[summer]), b_lab(beta$R_wrt), logx = TRUE)
p3 <- panel(G_m, cv$n_dams, "Number of dams", g_class, gpal2, names(gpal2), expression("Seasonal gain, "*G), b_lab(beta$G_dams))
p4 <- panel(G_m, cv$wrt, "Water residence time in summer (s)", g_class, gpal2, names(gpal2), expression("Seasonal gain, "*G), b_lab(beta$G_wrt), logx = TRUE)
figI <- (p1 + p2) / (p3 + p4) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "a") &
  theme(legend.position = "top")
ggsave(here("results", "figures", "plan", "fig_intensity.png"), figI, width = 10, height = 8, dpi = 300)
figI
bt <- function(k) sprintf("%.2f [%.2f, %.2f]", median(beta[[k]]), quantile(beta[[k]], .025), quantile(beta[[k]], .975))
bp <- function(k) sprintf("%.2f", mean(beta[[k]] > 0))
```

::: {.callout-note}
## Key message of @fig-intensity

**The intensity of beaver-engineering orders the streams only weakly, and only for the summer response.** Along the water residence time, the summer relative response increases: slope $\beta$ = `r bt("R_wrt")` per SD of log residence time, $P(\beta > 0)$ = `r bp("R_wrt")` — streams that hold water longer tend to add more DOC in summer, the stream-level counterpart of the Q2 path water residence time → phytoplankton → downstream DOC. Along the number of dams the slope is `r bt("R_dams")` ($P(\beta > 0)$ = `r bp("R_dams")`). The seasonal gain $G$ is not ordered by either gradient (`r bt("G_dams")`, $P(\beta > 0)$ = `r bp("G_dams")` with the number of dams; `r bt("G_wrt")`, $P(\beta > 0)$ = `r bp("G_wrt")` with the water residence time). Sources, sinks, reinforcing and dampening streams occur at every number of dams and across the whole range of residence times: the intensity of beaver-engineering sets the stage, but it does not decide which streams add or remove DOC, nor which reinforce or dampen the inherited seasonal contrast.

Sentence for the Results (Appendix S1: Figure S12): the relative response in summer increased with the water residence time ($\beta$ = `r bt("R_wrt")` per SD of log residence time, $P(\beta > 0)$ = `r bp("R_wrt")`) and weakly with the number of dams (`r bt("R_dams")`, $P(\beta > 0)$ = `r bp("R_dams")`); the seasonal gain was ordered by neither (`r bt("G_dams")` and `r bt("G_wrt")`).
:::

# 7. Spatial coherence: is the beaver seasonal modification regional or local?

**Question (Discussion support):** *the inherited seasonal contrast is generated over whole floodplains — is the modification that beaver-engineering imposes on it organised at the same regional scale, or is it stream-specific?* The answer decides how the Discussion frames beaver-engineering: if $B$ were regionally coherent, beaver-engineered streams would act together as a landscape-scale control on the seasonal DOC signal; if $B$ is spatially incoherent, each beaver-engineered stream is an independent local control point on a regional signal — the framing the manuscript title claims.

For each stream we correlate its value with the mean of its 8 nearest neighbours (Spearman), computed on every posterior draw of the concentrations, so the coherence estimate carries the full uncertainty. $T = \text{DOC}_{down,summer} - \text{DOC}_{down,winter}$ is the **downstream seasonal contrast** ($T = S + B$).

**Manuscript placement: Appendix S1 (Figure S11), plus one sentence in the Discussion.**

```{r fig-spatial}
#| fig-height: 3.5
#| fig-cap: "**Spatial coherence of the inherited seasonal contrast $S$, the beaver seasonal modification $B$ and the downstream seasonal contrast $T$.** Posterior distribution of the Spearman correlation between each stream's value and the mean of its 8 nearest neighbours, computed on every posterior draw of the concentrations. Point: median; bars: 66% and 95% credible intervals; dashed line: no spatial coherence. $S$ and $B$ are defined in @tbl-variables ($T = S + B$)."
# stream coordinates (longitude, latitude of the upstream sampling point)
co <- read.csv(here("data", "processed", "m6.csv")) |> distinct(site, .keep_all = TRUE) |>
  select(site, lon = x_upstream, lat = y_upstream)
d2 <- left_join(LR$d, co, by = "site")
# planar distances in km (adequate at the scale of Switzerland)
kmx <- (d2$lon - mean(d2$lon)) * 111.32 * cos(mean(d2$lat) * pi / 180)
kmy <- (d2$lat - mean(d2$lat)) * 110.57
D <- as.matrix(dist(cbind(kmx, kmy))); diag(D) <- Inf
# for each stream, the indices of its 8 nearest neighbours (fixed across draws)
nb <- t(apply(D, 1, function(r) order(r)[1:8]))
# mean of the neighbours' values, for one vector of stream values
nbmean <- function(v) rowMeans(matrix(v[as.vector(nb)], nrow = length(v)))
# one Spearman correlation per posterior draw -> a posterior distribution of coherence
rho_draw <- function(M) apply(M, 1, function(v) cor(v, nbmean(v), method = "spearman"))
T_m <- S_m + B_m   # downstream seasonal contrast, per draw
sp <- bind_rows(tibble(q = "inherited seasonal contrast S", rho = rho_draw(S_m)),
                tibble(q = "beaver seasonal modification B", rho = rho_draw(B_m)),
                tibble(q = "downstream seasonal contrast T", rho = rho_draw(T_m))) |>
  mutate(q = factor(q, levels = c("beaver seasonal modification B", "downstream seasonal contrast T",
                                  "inherited seasonal contrast S")))
sp_med <- sp |> group_by(q) |> summarise(m = median(rho))
ps <- ggplot(sp, aes(x = rho, y = q)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
  stat_halfeye(.width = c(.66, .95), point_interval = "median_qi", fill = "grey80") +
  labs(x = "Spearman correlation with the mean of the 8 nearest neighbours", y = NULL) +
  theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank())
ggsave(here("results", "figures", "plan", "fig_spatial.png"), ps, width = 8, height = 3.5, dpi = 300)
ps
```

::: {.callout-note}
## Key message of @fig-spatial

**The inherited seasonal contrast is regional; the beaver seasonal modification is local.** $S$ is spatially coherent (Spearman `r sprintf("%.2f", sp_med$m[sp_med$q == "inherited seasonal contrast S"])`), the downstream seasonal contrast still inherits that regional structure (`r sprintf("%.2f", sp_med$m[sp_med$q == "downstream seasonal contrast T"])`), but $B$ shows no spatial coherence (`r sprintf("%.2f", sp_med$m[sp_med$q == "beaver seasonal modification B"])`, credible interval spanning zero): neighbouring streams do not modify the seasonal contrast in the same way.

Sentence for the Discussion: the seasonal DOC contrast that streams inherit is organised at the regional scale, whereas the modification imposed by beaver-engineering is stream-specific and spatially incoherent — each beaver-engineered stream reinforces, dampens or reverses a regionally inherited signal on its own terms.
:::

# 8. Model validation: can the SCM predict downstream DOC at a stream it has not seen?

The manuscript claims that the model can be used to anticipate the effect of beaver-engineering on stream DOC in Switzerland. That claim needs an out-of-sample test. Two are reported: (i) Pareto-smoothed importance-sampling leave-one-out cross-validation (PSIS-LOO) of the downstream-DOC SCM, in which every observed downstream concentration is predicted as if it had been left out of the fit; and (ii) a hold-out test, in which the SCM is refitted on 80% of the streams and predicts downstream DOC at the remaining 20% (`r if (file.exists(here("results","holdout_pred.rds"))) length(readRDS(here("results","holdout_pred.rds"))$test_sites) else "…"` streams), which it has never seen. At an unseen stream the site effect is unknown and the mediators are mostly unmeasured (they enter at their mean), so the hold-out prediction is what a user in Switzerland would actually have: upstream DOC, the dams, the channel gradient and solar radiation. For both tests, the derived $\Delta\text{DOC}$ (predicted downstream DOC minus upstream DOC) is scored as well, because that is the quantity the claim is about. Details of the derivation in `scm_outcome_choice.qmd`.

**Manuscript placement: Appendix S1 (Figure S13), with one sentence in the Methods and one in the Discussion.**

```{r validation}
#| echo: false
pred_path <- here("results", "scm_downstream_pred.rds"); hold_path <- here("results", "holdout_pred.rds")
have_pred <- file.exists(pred_path); have_hold <- file.exists(hold_path)
if (have_pred) P <- readRDS(pred_path)
if (have_hold) {
  H <- readRDS(hold_path)
  y_h <- H$test$DOC_out; mu_h <- colMeans(H$epred); q_h <- apply(H$pred, 2, quantile, c(.05, .95))
  d_obs_h <- H$test$DOC_out - H$test$DOC_input; d_pred_h <- mu_h - H$test$DOC_input
  hold_metrics <- c(R2 = 1 - sum((y_h - mu_h)^2) / sum((y_h - mean(y_h))^2), RMSE = sqrt(mean((y_h - mu_h)^2)),
                    cover90 = mean(y_h >= q_h[1, ] & y_h <= q_h[2, ]),
                    R2_d = 1 - sum((d_obs_h - d_pred_h)^2) / sum((d_obs_h - mean(d_obs_h))^2),
                    RMSE_d = sqrt(mean((d_obs_h - d_pred_h)^2)), RMSE_d_null = sqrt(mean((d_obs_h - mean(d_obs_h))^2)))
}
if (have_pred && have_hold) {
  lm <- P$loo_metrics; dm <- P$ddoc_metrics
  met <- function(y, mu, q = NULL) c(R2 = 1 - sum((y - mu)^2) / sum((y - mean(y))^2), RMSE = sqrt(mean((y - mu)^2)),
                                     cover = if (is.null(q)) NA else mean(y >= q[1, ] & y <= q[2, ]))
  # in-sample, full model (all 158 streams, site effects known)
  mu_in <- colMeans(P$ep_sub[, P$obs]); q_in <- apply(P$pp_sub[, P$obs], 2, quantile, c(.05, .95))
  in_full <- met(P$y, mu_in, q_in)
  d_in_full <- met(P$d_obs, colMeans(P$ep_sub[, P$both]) - P$dat$DOC_input[P$both])
  # in-sample, hold-out fit on its training streams
  y_tr <- H$train$DOC_out; mu_tr <- colMeans(H$epred_train); q_tr <- apply(H$pred_train, 2, quantile, c(.05, .95))
  in_train <- met(y_tr, mu_tr, q_tr)
  d_in_train <- met(y_tr - H$train$DOC_input, mu_tr - H$train$DOC_input)
  f2 <- function(x) sprintf("%.2f", x); fp <- function(x) sprintf("%.0f%%", 100 * x)
  kable(tribble(
    ~Quantity, ~Test, ~`In / out of sample`, ~`R²`, ~`RMSE (mg L⁻¹)`, ~`RMSE of predicting the mean`, ~`90% interval coverage`,
    "Downstream DOC", "Full model, all 158 streams", "in-sample", f2(in_full["R2"]), f2(in_full["RMSE"]), f2(sd(P$y)), fp(in_full["cover"]),
    "Downstream DOC", "Full model, PSIS-LOO", "out-of-sample", f2(lm["R2_loo"]), f2(lm["RMSE_loo"]), f2(sd(P$y)), fp(lm["cover90"]),
    "Downstream DOC", sprintf("Hold-out fit, %d training streams", length(unique(H$train$site))), "in-sample", f2(in_train["R2"]), f2(in_train["RMSE"]), f2(sd(y_tr)), fp(in_train["cover"]),
    "Downstream DOC", sprintf("Hold-out fit, %d unseen streams (%d samplings)", length(H$test_sites), length(y_h)), "out-of-sample", f2(hold_metrics["R2"]), f2(hold_metrics["RMSE"]), f2(sd(y_h)), fp(hold_metrics["cover90"]),
    "Derived ΔDOC", "Full model, all 158 streams", "in-sample", f2(d_in_full["R2"]), f2(d_in_full["RMSE"]), f2(sd(P$d_obs)), "—",
    "Derived ΔDOC", "Full model, PSIS-LOO", "out-of-sample", f2(dm["R2_loo"]), f2(dm["RMSE_loo"]), f2(dm["RMSE_null"]), "—",
    "Derived ΔDOC", "Hold-out fit, training streams", "in-sample", f2(d_in_train["R2"]), f2(d_in_train["RMSE"]), f2(sd(y_tr - H$train$DOC_input)), "—",
    "Derived ΔDOC", "Hold-out fit, unseen streams", "out-of-sample", f2(hold_metrics["R2_d"]), f2(hold_metrics["RMSE_d"]), f2(hold_metrics["RMSE_d_null"]), "—"),
    caption = sprintf("Predictive performance of the downstream-DOC SCM, in and out of sample. In-sample predictions include the stream random effects (the model has seen the stream); out-of-sample predictions do not. Hold-out fit: max R-hat %.3f.", H$rhat))
} else cat("Predictions not yet computed (notebook/R/scm_predictive_checks.R, notebook/R/scm_holdout.R).")
```

```{r fig-holdout}
#| fig-height: 4.5
#| fig-cap: "**Hold-out test of the downstream-DOC SCM.** The model was refitted on 80% of the streams and predicts the other 20%, which it has never seen. (a) Predicted (posterior mean, with 90% predictive interval) against observed downstream DOC at the unseen streams; dashed line: 1:1. (b) The derived local response, predicted downstream DOC minus upstream DOC, against the observed ΔDOC at the same samplings. $\\Delta\\text{DOC}_s$ is defined in @tbl-variables."
if (have_hold) {
  pa <- ggplot(tibble(y = y_h, mu = mu_h, lo = q_h[1, ], hi = q_h[2, ]), aes(y, mu)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
    geom_linerange(aes(ymin = lo, ymax = hi), colour = "grey70", alpha = .6) + geom_point(size = 1.6, colour = "#3B9AD1") +
    labs(x = expression("Observed downstream DOC (mg "*L^-1*")"), y = expression("Predicted downstream DOC (mg "*L^-1*")")) +
    theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank())
  pb <- ggplot(tibble(o = d_obs_h, p = d_pred_h), aes(o, p)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
    geom_hline(yintercept = 0, colour = "grey80") + geom_vline(xintercept = 0, colour = "grey80") +
    geom_point(size = 1.6, colour = "#3B9AD1") +
    labs(x = expression("Observed "*Delta*"DOC (mg "*L^-1*")"), y = expression("Predicted "*Delta*"DOC (mg "*L^-1*")")) +
    theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank())
  figH <- pa + pb + plot_annotation(tag_levels = "a")
  ggsave(here("results", "figures", "plan", "fig_holdout.png"), figH, width = 9, height = 4.5, dpi = 300)
  figH
}
```

::: {.callout-important}
## Key message of @fig-holdout, and what the manuscript may claim

`r if (have_pred && have_hold) sprintf("**Downstream DOC at an unseen Swiss stream is predictable; the local change is not.** The SCM predicts downstream DOC with R² = %.2f (hold-out) and %.2f (leave-one-out), an RMSE of about %.1f mg L⁻¹ against a spread of %.1f mg L⁻¹ in the data, and 90%% predictive intervals that contain %.0f%% of the unseen observations — calibrated. In-sample and out-of-sample accuracy are the same (R² = %.2f in sample against %.2f leave-one-out for the full model; %.2f in sample for the hold-out fit): the model does not overfit, and its R² on all streams is capped not by unseen streams but by a few very high concentrations in the heavy tail, which the Student-t likelihood deliberately does not chase (RMSE 2.7 mg L⁻¹ against a spread of 3.7). The hold-out streams happen to lie in the bulk of the distribution, hence their higher R². Almost all of the predictive skill is inheritance: downstream DOC is 95%% upstream DOC. The derived local response ΔDOC is predicted no better than its own mean (hold-out R² = %.2f; RMSE %.2f against %.2f for the mean), because ΔDOC is small relative to the inherited state and its sign is stream-specific (Q1, Figure 2b).", hold_metrics["R2"], P$loo_metrics["R2_loo"], hold_metrics["RMSE"], sd(y_h), 100 * hold_metrics["cover90"], in_full["R2"], P$loo_metrics["R2_loo"], in_train["R2"], hold_metrics["R2_d"], hold_metrics["RMSE_d"], hold_metrics["RMSE_d_null"]) else "Pending."`

So the sentence the manuscript can support is not "the model forecasts the beaver effect at any stream", but: **"the model predicts downstream DOC at unsampled Swiss streams from the upstream concentration and the measured drivers (hold-out R² given above), and it predicts that the beaver-mediated change will be small relative to that inherited concentration and of stream-specific sign — where it becomes a source or a sink cannot be forecast from dams, residence time or solar radiation alone (Q2, Q4)."** That is a weaker but defensible claim, and it is the claim the data make.
:::

# 9. Methods figure

**Manuscript placement: main text, Figure 3 (both panels redrawn: panel a shows the 158 analysed streams, panel b shows $S$, $\Delta\text{DOC}_s$, $R_s$ and $F$ in place of the amplification-factor quantities).**

The chunk below draws panel (a) directly from the analysed data — the red dots are exactly the 158 streams, so map and analysis can never disagree — plus the 16 producer-abundance streams (black triangles; coordinates in `data/processed/producer_sites_coordinates.csv`, taken from the field-campaign site list). Panel (b) is the schematic of the derived variables.

```{r fig-map}
#| fig-height: 11
#| fig-cap: "**Sampling design and derived variables.** (a) Sampling locations across Switzerland (elevation background): red dots are the 158 analysed streams; black triangles are the 16 streams where primary producer abundance was sampled. Water was sampled upstream of the first and downstream of the last beaver dam of each stream, once in winter and once in summer. (b) Schematic of the derived variables: the inherited seasonal contrast $S$ is generated over the floodplain area $A_\\text{floodplain}$; the local response $\\Delta\\text{DOC}_s$ is measured across the beaver-engineered stream of ponded area $A_\\text{beaver}$; the relative response is $R_s = \\Delta\\text{DOC}_s / \\lvert S \\rvert$ and the footprint contrast is $F = A_\\text{floodplain} / A_\\text{beaver}$. All variables are defined in @tbl-variables."
# panel (a): the 158 analysed streams (d, from the Q1 chunk) with coordinates,
# and the 16 producer-abundance streams from the field-campaign site list
pts  <- d |> distinct(site) |>
  left_join(read.csv(here("data", "processed", "m6.csv")) |> distinct(site, .keep_all = TRUE) |>
              select(site, lon = x_upstream, lat = y_upstream), by = "site")
prod <- read.csv(here("data", "processed", "producer_sites_coordinates.csv"))
# elevation background: Swiss DEM (30 arc-second, data/elevation/) with a
# hillshade underlay and a hypsometric tint, as in the original map
library(terra); library(tidyterra)
dem    <- rast(here("data", "elevation", "CHE_elv_msk.tif"))
hill   <- shade(terrain(dem, "slope", unit = "radians"),
                terrain(dem, "aspect", unit = "radians"), 40, 270)
greys  <- hcl.colors(1000, "Grays")
hillcol <- greys[round(scales::rescale(values(hill), to = c(1, 1000)))]
pmap <- ggplot() +
  geom_spatraster(data = hill, fill = hillcol, maxcell = Inf) +
  geom_spatraster(data = dem, maxcell = Inf) +
  scale_fill_hypso_tint_c(palette = "dem_poster", alpha = 0.5, na.value = NA, name = "Elevation (m)") +
  geom_point(data = pts, aes(lon, lat), colour = "red", size = 1.7, alpha = 0.9) +
  geom_point(data = prod, aes(longitude, latitude), colour = "black", size = 2.2, shape = 17) +
  annotate("text", x = 5.99, y = 47.75, label = "(a)", size = 6.5, hjust = 0) +
  theme_void() + theme(legend.position = "right")
# panel (b): the schematic, included as an image below the map
schem <- png::readPNG(here("results", "figures", "plan", "schematic_S_R_F.png"))
psch  <- ggplot() + annotation_custom(grid::rasterGrob(schem)) +
  annotate("text", x = 0.02, y = 0.97, label = "(b)", size = 6.5, hjust = 0) +
  xlim(0, 1) + ylim(0, 1) + theme_void()
fig3 <- pmap / psch + plot_layout(heights = c(1, 1.15))
ggsave(here("results", "figures", "plan", "fig_map_schematic.png"), fig3, width = 9, height = 11, dpi = 300, bg = "white")
fig3
```

# 10. Figures for the manuscript

```{r inventory}
#| echo: false
kable(tribble(
  ~`Manuscript figure`, ~Content, ~Answers, ~`This document`, ~Status,
  "Figure 1a", "Posterior predictive distribution of $\\Delta\\text{DOC}_s$ by season", "Q1", "@fig-q1", "existing",
  "Figure 1b", "Structural causal model, outcome downstream DOC, with estimated effects", "Q2", "@fig-q2", "**redrawn — outcome is downstream DOC**",
  "Figure 2",  "Relative response $R_s$: across streams (a), by stream (b); seasonal gain $G$ by stream (c)", "Q3", "@fig-R", "**new — replaces the amplification-factor figure**",
  "Figure 3",  "Map (158 analysed streams), sampling design and schematic of $S$, $\\Delta\\text{DOC}_s$, $R_s$, $F$", "Methods", "@fig-map", "**both panels redrawn**",
  "Figure S1–S9", "Hypothesised DAG, priors, checks, stream-level classification, total effects, connectivity model", "Q2 support", "—", "existing",
  "Figure S10", "Derivation of the variables", "all", "@fig-dag", "**new**",
  "Figure S11", "Spatial coherence of $S$, $B$ and $T$", "Discussion support", "@fig-spatial", "**new**",
  "Figure S12", "$R_{summer}$ and $G$ against the number of dams and the water residence time", "Q4", "@fig-intensity", "**new**",
  "Figure S13", "Hold-out prediction of downstream DOC and $\\Delta\\text{DOC}$", "Methods support", "@fig-holdout", "**new**"))
```

::: {.callout-warning}
## What changes in the manuscript

1.  **Abstract.** The sentence *"Compared to floodplain-scale DOC variation, beaver-engineering amplified DOC concentration changes 141 times in summer and 33 times in winter."* is replaced by: *"Relative to the inherited seasonal contrast of upstream DOC, beaver-engineering added DOC equal to `r sprintf("%.0f%%", 100*median(medR(R_s)))` of that contrast in summer (95% credible interval `r sprintf("%.0f–%.0f%%", 100*quantile(medR(R_s), .025), 100*quantile(medR(R_s), .975))`) and had no net effect in winter, on a footprint of about `r sprintf("%.1f%%", 100 * median(1 / d$F))` of the floodplain area."*
2.  **Results.** The amplification-factor section and its figure are removed; Q3 is answered by $R_s$ (Figure 2a, b) and by the seasonal gain $G$ (Figure 2c: share of streams that reinforce / dampen / reverse).
3.  **Methods.** The definitions of $S$, $B$, $G$, $R_s$ and $F$ and the measurement-error model on the concentrations replace the amplification-factor equations; one sentence states that the structural causal model is parameterized with downstream DOC as the outcome (inheritance slope 0.95) and $\Delta\text{DOC}_s$ is derived from the posterior; Figure 3b is redrawn.
4.  **Discussion.** One sentence on spatial coherence: the inherited seasonal contrast is regional, the beaver seasonal modification is local (Figure S11).
5.  **Whole manuscript.** Every mention of 176 streams is replaced by **158**, the streams actually analysed (abstract, introduction, methods, Figure 3a and its caption); the count appears with the same value everywhere.
6.  **Appendix S1.** Figure S10 (derivation of the variables), Figure S11 (spatial coherence), Figure S12 (response against intensity of beaver-engineering) and Figure S13 (hold-out validation) are added.
7.  **Methods and Discussion.** One sentence each on validation: the SCM predicts downstream DOC at unseen streams (hold-out R², coverage), but not the sign of the local change; the "predictive basis" claim of the Discussion is reworded accordingly (section 8).
:::

# 11. Reproducibility

```{r session}
#| echo: false
cat("Data: data/processed/m12.csv | concentration posteriors: notebook/concentrations_fit.qmd -> results/concentrations.rds (thinned copy concentrations_thin.rds is tracked)\n")
sessionInfo()$R.version$version.string
```

# References

::: {#refs}
:::
