---
title: "Posterior DOC concentrations per stream"
subtitle: "The measurement-error model behind S, B, G and R_s"
author: "Leonardo Capitani"
date: last-modified
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
execute:
  echo: true
  message: false
  warning: false
---

Every derived quantity of the manuscript — the inherited seasonal contrast $S$, the beaver seasonal modification $B$, the seasonal gain $G$ and the relative response $R_s$ — is a difference or a ratio of measured DOC concentrations. Forming them directly from the raw values would ignore the 0.2 mg L$^{-1}$ instrument error and, for the ratios, let small denominators explode. This document fits one Bayesian hierarchical model per measured concentration and computes every derived quantity from the posterior draws, so the measurement uncertainty is carried through to the end.

Output: `results/concentrations.rds` (all draws; a thinned copy with 4000 draws, `results/concentrations_thin.rds`, is tracked in git and used by `analysis_plan.qmd`).

```{r setup}
suppressMessages({library(brms); library(tidyverse); library(here)})
```

## 1. One row per stream

Each stream was sampled once in winter and once in summer, upstream and downstream of the beaver-engineered stream. We pair the two seasons and keep the streams where all four concentrations exist: **158 streams**.

```{r data}
m <- read.csv(here("data", "processed", "m12.csv"))
w <- m |> filter(season == "winter") |>
  select(site, beaver_territory, up_w = DOC_input, dn_w = DOC_out,
         Afp = A_floodplain_km2, Abv = A_beaver_km2)          # winter sampling
s <- m |> filter(season == "summer") |>
  select(site, beaver_territory, up_s = DOC_input, dn_s = DOC_out)  # summer sampling
d <- inner_join(w, s, by = c("site", "beaver_territory")) |>
  filter(complete.cases(up_w, dn_w, up_s, dn_s)) |>
  mutate(reach = factor(row_number()))   # stream id used as the grouping factor
nrow(d)
```

## 2. One measurement-error model per concentration

Four models, one for each measured concentration (upstream winter, downstream winter, upstream summer, downstream summer). Each model is fitted on the log scale, because concentrations are positive and right-skewed, and every observation carries its known instrument error, moved to the log scale by the delta method ($\text{se}_{\log} = 0.2 / \text{value}$). The stream enters as a random intercept: its deviation is the stream's true log concentration.

Priors are weakly informative: the intercept is centred at a log concentration of 1 (about 2.7 mg L$^{-1}$) with a wide scale, and the between-stream standard deviation gets a half-Student-t.

```{r fit}
pri <- c(prior(normal(1, 1.5), class = "Intercept"),
         prior(student_t(3, 0, 1), class = "sd"))
fitc <- function(v) {
  dd <- d |> mutate(
    y      = log(pmax(.data[[v]], 0.05)),        # log concentration (floored at 0.05)
    se_log = 0.2 / pmax(.data[[v]], 0.25))       # instrument error on the log scale
  # y | se(...): each observation carries its known measurement error;
  # (1 | reach): one true-concentration deviation per stream
  brm(bf(y | se(se_log, sigma = FALSE) ~ 1 + (1 | reach)),
      data = dd, prior = pri, chains = 4, cores = 4, iter = 6000, warmup = 2000,
      seed = 7, backend = "cmdstanr",
      control = list(adapt_delta = 0.99, max_treedepth = 12),
      file = here("results", "models_fit", paste0("conc_", v)), refresh = 0)
}
```

## 3. Derived quantities, per posterior draw

For each posterior draw and each stream: back-transform intercept plus stream deviation to mg L$^{-1}$, then compute the differences and ratios. Each derived quantity is therefore a full posterior distribution per stream — never a single number formed from raw values.

```{r derive}
# for each draw: intercept + stream deviation, back-transformed to mg/L
conc <- function(f) { x <- as_draws_df(f)
  exp(as.matrix(x[, grep("^r_reach\\[", names(x))]) + x$b_Intercept) }
L <- setNames(lapply(c("up_w", "dn_w", "up_s", "dn_s"), function(v) conc(fitc(v))),
              c("up_w", "dn_w", "up_s", "dn_s"))
S  <- L$up_s - L$up_w   # inherited seasonal contrast (upstream, summer - winter)
dw <- L$dn_w - L$up_w   # local response in winter
ds <- L$dn_s - L$up_s   # local response in summer
saveRDS(list(S = S, dw = dw, ds = ds,
             R_s = ds / abs(S), R_w = dw / abs(S),   # relative response by season
             up_s = L$up_s, up_w = L$up_w, d = d),
        here("results", "concentrations.rds"))
```

## 4. Thinned copy for the analysis plan

The full file is large; a thinned copy with 4000 draws is tracked in git and used by every downstream document.

```{r thin}
full <- readRDS(here("results", "concentrations.rds"))
keep <- round(seq(1, nrow(full$S), length.out = 4000))
thin <- lapply(full, function(x) if (is.matrix(x)) x[keep, ] else x)
saveRDS(thin, here("results", "concentrations_thin.rds"))
```

## Reproducibility

```{r session}
#| echo: false
sessionInfo()$R.version$version.string
```
