---
title: "Lateral connectivity analysis"
author: "Leonardo Capitani"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
knitr:
     opts_chunk:
        R.options:
         width: 120
format: 
      html:
          toc: true
          code-fold: true
          code-tools: true
          code-link: true
          embed-resources: true
          code-line-numbers: true
html-math-method: 
  method: mathjax
  url: https://cdnjs.cloudflare.com/ajax/libs/mathjax/2.7.2/MathJax.js?config=TeX-AMS-MML_HTMLorMML
execute:
  echo: true
  message: false
  warning: false 
  cache: false
  freeze: true  # never re-render during project render
editor: visual
code-annotations: select
editor_options: 
  chunk_output_type: console
bibliography: references.bib
---

```{r libraries, message=FALSE}
invisible(Sys.setlocale("LC_CTYPE", "en_US.UTF-8"))  # print non-ASCII (Delta, arrows) correctly

# Load libraries
library(tidyverse)
library(here)
library(DT)
library(ggdist)
library(scales)
library(readxl)
library(chron)# for time formatting
library(brms)
library(tidybayes)
library(dplyr)
library(stringr) 
library(bayesplot)
library(dagitty)
```

## Introduction

This notebook will present additional analysis in regard to the following research question:

::: callout-tip
## Research question:

Controlling for beaver engineering effects, to what extent does lateral stream connectivity influence dissolved organic carbon (DOC) concentration ?
:::

::: callout-important
Please refer to Kaspar Berger's Master thesis, chapter 4.12.4 and chapter 5.5 to get details about the definition of stream later connectivity.

In this study, we are going to use Berger's methodology to define stream lateral connectivity: `lateral_connectivity` is a categorical vcariable with three levels.

- `No_wetland`: no presence of wetland around the stream course,

- `yes_not connected`: presence of wetland not directly connected to the stream (figure 1a),

- `yes_connected` : presence of wetland and it is connected to the stream (figure 1b).
:::

![**The three levels of `lateral_connectivity`.** Two streams with contrasting lateral connectivity to an adjacent wetland: on the left a wetland present but not directly connected to the channel (`yes_not_connected`), on the right a wetland hydrologically connected to the channel (`yes_connected`). Sites with no wetland within the 50 m buffer are coded `No_wetland`. Image by Kaspar Berger.](../images/later_connectivty_berger-01.png){#fig-latconn}

Please note our main results:

![**The main structural causal model, for reference (manuscript Figure 1b).** Drivers of beaver-mediated DOC change: solid edges (red = negative, blue = positive) carry at least 90% posterior probability of a directional effect, dashed edges are inconclusive. Values are standardized average effects. The path this notebook extends is water residence time (`WRT`), to which `lateral_connectivity` is added as an additional parent.](../images/Capitani_Figure_2b.png){#fig-scm}

We do already know that the mediating path (WRT → ΔDOC) is weak because an increase in WRT influence 1-2% the `deltaDOC`.

## DAG

Let's define our causal diagram where predictors (`lateral_connectivity`) and response variable (`deltaDOC`) are connected by causal influences. Recall that WRT is the water residence time.

We do know that stream lateral connectivity to a wetland creates an additional storage and exchange compartment for stream water — when a wetland is hydrologically connected, water can move laterally into and out of that storage zone, increasing the time it spends in the system before continuing downstream. This mechanism is well established in the hydrology literature on stream-floodplain/wetland exchange, where connected off-channel storage areas are consistently shown to extend residence time relative to channels lacking such lateral exchange (transient storage zone theory; Bencala & Walters 1983; Wohl 2021, *Rev. Geophys.*).

::: {.callout-note appearance="simple"}
**Hypothesis:**

We hypothesize that if a stream is connected to the adjacent wetland then the beaver pond water residence time varies with the degree of lateral connectivity between the stream and adjacent wetlands.
:::

```{r dag}

dag <- dagitty('dag {
Max_dam_height [pos="-0.991,0.879"]
WRT [pos="-0.510,0.684"]
channel_gradient [pos="-0.518,0.295"]
delta_DOC [outcome,pos="0.990,0.636"]
lateral_connectivity [exposure,pos="-0.503,1.098"]
macrophytes [pos="0.373,0.249"]
number_dams [pos="-0.977,0.506"]
phytoplankton [pos="0.323,0.736"]
soil_litter [pos="0.288,1.289"]
solar_radiation [pos="-0.159,-0.285"]
upstream_DOC [pos="0.878,0.164"]
Max_dam_height -> WRT
WRT -> macrophytes
WRT -> phytoplankton
channel_gradient -> WRT
lateral_connectivity -> WRT
macrophytes -> delta_DOC
number_dams -> WRT
phytoplankton -> delta_DOC
soil_litter -> delta_DOC
solar_radiation -> WRT
solar_radiation -> macrophytes
solar_radiation -> phytoplankton
solar_radiation -> soil_litter
upstream_DOC -> delta_DOC
}

'
)

#plot(dag)
```

![**Extended DAG including `lateral_connectivity` (Appendix S1, Figure S8).** Identical to the main SCM except for one added causal path, from `lateral_connectivity` to water residence time (`WRT`). Because `lateral_connectivity` has no parents, no back-door path to `delta_DOC` exists and its total effect is identified without adjustment; the open front-door paths run through `WRT` and the primary producers.](../images/DAG_lateral_conn.png){#fig-daglateral}

So in this DAG, there are exactly **two open front-door paths**:

1.  lateral_connectivity → WRT → macrophytes → delta_DOC

2.  lateral_connectivity → WRT → phytoplankton → delta_DOC

If our question question is: how much does connectivity influence DOC overall, then:

The total effect of `lateral_connectivity` on `delta_DOC` is the *sum of all causal influence* flowing through every open path — direct arrow (if any) plus every mediated pathway (through `WRT`, `macrophytes`, `phytoplankton`, `soil_litter`, etc.).

```{r adjset_total}
# Step 3-4: backdoor paths and adjustment sets, for TOTAL effect
adjustmentSets(dag, exposure = "lateral_connectivity", outcome = "delta_DOC", effect = "total")
# list the open paths (the orange front-door paths in Appendix S1: Figure S8)
paths(dag, "lateral_connectivity", "delta_DOC")
```

Since there's no backdoor path (nothing causes `lateral_connectivity`), there's nothing to adjust for — you want all those mediator paths to stay open so their contribution is captured in the estimate. That's why the answer is `{}`.

::: {.callout-important appearance="simple"}
We conclude that:

**No open biasing paths.**

No adjustment is necessary to estimate the total effect of lateral_connectivity on delta_DOC.
:::

That means the *total* effect of `WRT` on `delta_DOC` is also identified with an empty adjustment set — we can regress `delta_DOC` directly on `WRT` alone, skipping `macrophytes` and `phytoplankton` entirely, and still get an unbiased estimate of WRT's total downstream effect.

## Load data

We are going to load the data sent by Kaspar Berger ([bergersumis\@gmail.com](mailto:bergersumis@gmail.com)).

```{r load_processed_data_frame}
l <- read_csv(here("data","processed","later_connectivity_berger.csv"))

# Standardize category labels to match your established naming scheme
l$lateral_connectivity <- recode(l$lateral_connectivity,
                                 "No" = "No_wetland",
                                 "Yes_not_connected" = "yes_not_connected",
                                 "Yes_connected" = "yes_connected")

# Resolve the two real conflicts: both AEG01_1 and AEG01_2 are confirmed
# "yes_not_connected" — overwrite all instances, then dedup
l$lateral_connectivity[l$site %in% c("AEG01_1", "AEG01_2")] <- "yes_not_connected"

# Now every duplicate (the 2 conflicts + the 3 exact dupes: TG02, MUE03, TG02_17)
# collapses to identical rows, so unique() removes all of them cleanly
l_clean <- unique(l)

# dataframe that we do not remove NAs observations 
m6 <- read_csv(here("data","processed","m6.csv")) |> mutate(season = as.factor(season)) 

```

Add the `later connectivity` variable to the original data frame:

```{r merge_later_conn}
# lateral_connectivity lives in a separate table `l`, keyed by site —
# merge it into your main data before fitting
# Re-merge and confirm the row count is now exactly right
m6_merged <- merge(m6, l_clean, by = "site", all.x = TRUE)

na_sites <- sort(unique(m6_merged$site[is.na(m6_merged$lateral_connectivity)]))

# Filter to summer observations only (one row per site), then build the same table
m6_summer <- m6_merged[m6_merged$season == "summer", ]   # adjust value if your coding differs (e.g. "Summer", "S")

tab_summer <- table(m6_summer$lateral_connectivity, useNA = "always")
pct_summer <- round(prop.table(tab_summer) * 100, 1)

result_summer <- data.frame(
  category = names(tab_summer),
  n        = as.vector(tab_summer),
  pct      = paste0(as.vector(pct_summer), "%")
)

result_summer
```

```{r}
#| label: na-sites-list
#| echo: false
#| output: asis

na_sites <- sort(unique(m6_merged$site[is.na(m6_merged$lateral_connectivity)]))

cat("::: {.callout-warning title=\"Sites missing `lateral_connectivity` classification\"}\n\n")
cat("**", length(na_sites), " sites** (", length(na_sites) * 2,
    " rows across both seasons) lack a connectivity category:\n\n", sep = "")
cat(paste0("`", na_sites, "`", collapse = ", "))
cat("\n\n:::\n")
```

### Drop NA rows:

```{r drop_nas}

m6_fit <- m6_merged[!is.na(m6_merged$lateral_connectivity), ]

nrow(m6_fit)   # should be 362 - 44 = 318

# set factor levels with No_wetland as reference
m6_fit$lateral_connectivity <- factor(
  m6_fit$lateral_connectivity,
  levels = c("No_wetland", "yes_not_connected", "yes_connected")
)

table(m6_fit$lateral_connectivity)  # confirm no NAs remain, counts look right
```

### Model formula

This is the 2-equation parsimonious version we settled on — `lateral_connectivity` enters only the `water_res_time_z` equation (its sole causal pathway per the DAG), and `delta_DOC` regresses on the mediator `WRT` alone (no backdoor adjustment needed, since the total effect is unconfounded):

```{r fit_model , message=FALSE, }

bform <- 
  bf(delta_DOC | mi(0.283) ~ 0 + Intercept + 
       mi(water_res_time_z) +
       (0 + Intercept | site),
     family = student()) + 
  
  bf(water_res_time_z | mi()  ~ 0 + n_dams_z + mi(dam_height_z) + slope_z + 
       lateral_connectivity + solar_z,
     family = gaussian()) +
  
  bf(dam_height_z | mi()  ~ 0 + Intercept, family = gaussian()) +
  
  set_rescor(FALSE)

#detach("package:rethinking", unload = TRUE)

# Informative priors: shared parameters carry the same priors as in the main SCM (notebook.qmd);
# the three connectivity-level coefficients get weakly-informative normal(0, 1) priors (WRT is z-scored).
prio_conn <- c(
  prior(normal(0, 2.5),       class = "b", coef = "Intercept",          resp = "deltaDOC"),
  prior(normal(0, 0.1),       class = "b", coef = "miwater_res_time_z", resp = "deltaDOC"),
  prior(student_t(3, 0, 2.5), class = "sigma",                          resp = "deltaDOC"),
  prior(gamma(2, 0.1),        class = "nu",                             resp = "deltaDOC"),
  prior(normal(0, 2),         class = "sd", coef = "Intercept", group = "site", resp = "deltaDOC"),
  prior(normal(0.15, 0.075),  class = "b", coef = "n_dams_z",           resp = "waterrestimez"),
  prior(normal(0.15, 0.075),  class = "b", coef = "midam_height_z",     resp = "waterrestimez"),
  prior(normal(0, 0.075),     class = "b", coef = "slope_z",            resp = "waterrestimez"),
  prior(normal(0, 0.075),     class = "b", coef = "solar_z",            resp = "waterrestimez"),
  prior(normal(0, 1),         class = "b", coef = "lateral_connectivityNo_wetland",        resp = "waterrestimez"),
  prior(normal(0, 1),         class = "b", coef = "lateral_connectivityyes_not_connected", resp = "waterrestimez"),
  prior(normal(0, 1),         class = "b", coef = "lateral_connectivityyes_connected",     resp = "waterrestimez"),
  prior(normal(0, 0.5),       class = "sigma",                          resp = "waterrestimez"),
  prior(normal(0, 1),         class = "b", coef = "Intercept",          resp = "damheightz"),
  prior(normal(0, 0.5),       class = "sigma",                          resp = "damheightz"))

fit_connectivity <- brm(bform, data = m6_fit, 
                        prior = prio_conn,
                        chains = 4, 
                        cores = 10, seed = 42,
                        backend = "cmdstanr",
                        file = here("results","models_fit","fitconnectivity_informative"),
                        sample_prior = FALSE
)

#summary(fit_connectivity)

```

Now we are going to extract the posterior draws from the model we fitted:

```{r draws}

draws <- as_draws_df(fit_connectivity)

mean_no_wetland    <- draws$b_waterrestimez_lateral_connectivityNo_wetland
mean_not_connected <- draws$b_waterrestimez_lateral_connectivityyes_not_connected
mean_connected     <- draws$b_waterrestimez_lateral_connectivityyes_connected

effect_wrt_not_connected <- mean_not_connected - mean_no_wetland
effect_wrt_connected     <- mean_connected - mean_no_wetland

s1 <- summarise_draws(data.frame(
  effect_wrt_not_connected = effect_wrt_not_connected,
  effect_wrt_connected     = effect_wrt_connected
))

wrt_to_doc <- draws$bsp_deltaDOC_miwater_res_time_z

total_effect_not_connected <- effect_wrt_not_connected * wrt_to_doc
total_effect_connected     <- effect_wrt_connected * wrt_to_doc

s2 <- summarise_draws(data.frame(
  total_effect_not_connected = total_effect_not_connected,
  total_effect_connected     = total_effect_connected
))

effects_df <- data.frame(
  not_connected = total_effect_not_connected,
  connected     = total_effect_connected
) %>%
  pivot_longer(everything(), names_to = "contrast", values_to = "effect") %>%
  mutate(contrast = recode(contrast,
                           "not_connected" = "Yes, not connected\nvs No wetland",
                           "connected"      = "Yes, connected\nvs No wetland"))
```

Then plot the results as contrasts (Appendix S1: Figure S9):

```{r}
#| label: fig-contrasts
#| fig-cap: "**Appendix S1, Figure S9. Effect of lateral stream-wetland connectivity.** Left: the contrasts `effect_wrt_not_connected` and `effect_wrt_connected`, i.e. the effect of each connectivity level on water residence time relative to sites without wetland. Right: `total_effect_not_connected` and `total_effect_connected`, the total mediated effect on $\\Delta$DOC, obtained by multiplying each contrast by the WRT to $\\Delta$DOC coefficient `wrt_to_doc`. Points are posterior medians; thick and thin bars are the 66% and 95% credible intervals; the dashed line marks zero."
#| fig-width: 11
#| fig-height: 4.5
df_s9 <- bind_rows(
  tibble(stage = "A", contrast = "Wetland, not connected vs no wetland", draw = effect_wrt_not_connected),
  tibble(stage = "A", contrast = "Wetland, connected vs no wetland",     draw = effect_wrt_connected),
  tibble(stage = "B", contrast = "Wetland, not connected vs no wetland", draw = total_effect_not_connected),
  tibble(stage = "B", contrast = "Wetland, connected vs no wetland",     draw = total_effect_connected)) |>
  mutate(contrast = factor(contrast, levels = rev(c("Wetland, not connected vs no wetland", "Wetland, connected vs no wetland"))))
lab_s9 <- as_labeller(c(A = "Connectivity%->%'water residence time'", B = "'Total effect on'~Delta*'DOC'"), label_parsed)
(p_s9 <- ggplot(df_s9, aes(x = draw, y = contrast)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey40") +
  stat_halfeye(.width = c(.66, .95), fill = "steelblue", alpha = .8) +
  facet_wrap(~stage, scales = "free_x", labeller = lab_s9) +
  labs(x = "Standardized effect", y = NULL) +
  theme_bw(base_size = 15) +
  theme(axis.text = element_text(color = "black"), panel.grid.minor = element_blank(),
        strip.background = element_rect(fill = "grey92")))
ggsave(here("results", "figures", "figS9_connectivity.png"), p_s9, width = 11, height = 4.5, dpi = 300)
```

::: {.callout-note}
## What @fig-contrasts shows

Naming each posterior quantity explicitly (standardized effect, 90% credible interval, posterior probability of being positive):

-   **`effect_wrt_not_connected`** — an unconnected wetland versus no wetland, effect on water residence time: `r sprintf("%+.2f", mean(effect_wrt_not_connected))` [`r sprintf("%.2f, %.2f", quantile(effect_wrt_not_connected,.05), quantile(effect_wrt_not_connected,.95))`], P(> 0) = `r sprintf("%.2f", mean(effect_wrt_not_connected > 0))`
-   **`effect_wrt_connected`** — a connected wetland versus no wetland, effect on water residence time: `r sprintf("%+.2f", mean(effect_wrt_connected))` [`r sprintf("%.2f, %.2f", quantile(effect_wrt_connected,.05), quantile(effect_wrt_connected,.95))`], P(> 0) = `r sprintf("%.2f", mean(effect_wrt_connected > 0))`
-   **`wrt_to_doc`** — the mediating coefficient from water residence time to $\Delta$DOC: `r sprintf("%+.2f", mean(wrt_to_doc))` [`r sprintf("%.2f, %.2f", quantile(wrt_to_doc,.05), quantile(wrt_to_doc,.95))`], P(> 0) = `r sprintf("%.2f", mean(wrt_to_doc > 0))`
-   **`total_effect_not_connected`**: `r sprintf("%+.2f", mean(total_effect_not_connected))` [`r sprintf("%.2f, %.2f", quantile(total_effect_not_connected,.05), quantile(total_effect_not_connected,.95))`], and **`total_effect_connected`**: `r sprintf("%+.2f", mean(total_effect_connected))` [`r sprintf("%.2f, %.2f", quantile(total_effect_connected,.05), quantile(total_effect_connected,.95))`]

**Lateral stream-wetland connectivity does not detectably change $\Delta$DOC through the water-residence-time pathway.** The estimates of the remaining paths are unchanged from the main SCM (@fig-scm), so the conclusions of the manuscript are robust to adding this causal path.
:::

### Conclusions:

::: callout-important
- We find null total effect of lateral connectivity on ΔDOC via the water-residence-time pathway: both 90% CIs overlap zero, with small magnitudes on the standardized scale.

- We also see the WRT → ΔDOC mediating path itself is weak, with its CI touching \~0.00 at the lower bound — limiting our ability to detect a total effect even if connectivity strongly affected WRT.

- We treat this as power-limited rather than strong evidence of no effect — only 13 sites (\~7%) are `yes_connected` and 11 (\~6%) are `yes_not_connected`, so wide intervals likely reflect limited data rather than a true null.
:::
