---
title: "ΔDOC or downstream DOC as the outcome of the structural causal model?"
subtitle: "A balanced comparison of the two parameterizations"
author: "Leonardo Capitani"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    embed-resources: true
    fig-width: 9
    fig-align: center
bibliography: references.bib
csl: ecology-letters.csl
execute:
  echo: true
  message: false
  warning: false
---

```{r setup}
#| include: false
library(tidyverse); library(ggdist); library(patchwork); library(knitr); library(here); library(bayesplot); library(scales)
invisible(Sys.setlocale("LC_CTYPE", "en_US.UTF-8"))
ME <- readRDS(here("results", "me_refits.rds")); tD <- ME$delta$tab; tO <- ME$down$tab
gme <- function(tb, par) tb[tb$par == par, ]
have_pred <- file.exists(here("results", "me_predictions.rds"))
if (have_pred) { P <- readRDS(here("results", "me_predictions.rds")); cD <- P$cmpD; cO <- P$cmpO; s_up <- P$s_up; mu_up <- P$mu_up } else { s_up <- ME$s_up; mu_up <- NA }
q <- function(x) sprintf("%.3f [%.3f, %.3f]", mean(x), quantile(x, .025), quantile(x, .975))
qm <- function(x) sprintf("%.3f [%.3f, %.3f]", median(x), quantile(x, .025), quantile(x, .975))
```

::: {.callout-tip appearance="simple"}
## The questions addressed

1.  Why is the response variable of the structural causal model (SCM) **downstream DOC** and not **ΔDOC**, the process we want to model? What does the link upstream DOC → ΔDOC represent, as against upstream DOC → downstream DOC? (@sec-upstream)
2.  Does the choice **change the inference**? (@sec-one-model, @sec-practical)
3.  Where is the evidence that the model predicts downstream DOC well (predicted vs observed), and **how is ΔDOC estimated a posteriori** from a model that never sees it? (@sec-ppc, @sec-derived)
4.  If ΔDOC is derived from the prediction, does the **specification of the SCM** bias it? Would it be safer to model ΔDOC directly from the data? (@sec-spec)
5.  Should both models account for the **measurement error of upstream DOC**? (@sec-error)
6.  **What do we gain** with downstream DOC over ΔDOC? (@sec-gain)
7.  If beavers can raise or lower downstream DOC *depending on how much comes in*, the structure needs the links upstream DOC → primary producers as well as the direct link: what do the two DAGs look like and what do they give? (@sec-dags, @sec-inflow)
8.  Our interest is beaver-engineering's effect on **ΔDOC**, not on downstream DOC; and if the model is on downstream DOC, should the figures not show it? (@sec-target)
9.  Can the extended SCMs **answer Q1 as well**, instead of the separate paired-contrast model? (@sec-q1)
:::

::: {.callout-tip appearance="simple"}
## The two models

| | **delta_SCM** (file `delta_SCM_me`) | **downstream_DOC** (file `downstream_DOC_me`) |
|---|---|---|
| Outcome | $\Delta\text{DOC} = \text{DOC}_{down} - \text{DOC}_{up}$ | $\text{DOC}_{down}$ |
| Instrument error on the outcome | 0.283 mg L$^{-1}$ — the errors of both readings, $\sqrt{0.2^2 + 0.2^2}$ | 0.2 mg L$^{-1}$ — one reading |
| Upstream DOC | latent true value (its reading carries the 0.2 mg L$^{-1}$ error), standardized, as a predictor of ΔDOC *and* subtracted inside ΔDOC | latent true value (0.2 mg L$^{-1}$ error), in mg L$^{-1}$, as a predictor with a freely estimated **inheritance slope** |
| Links upstream DOC → phytoplankton, → macrophytes | yes | yes |
| Mediators, priors, stream random effect, Student-*t* likelihood, sampler | identical | identical |
| ΔDOC | is the outcome | is derived afterwards: predicted downstream DOC minus upstream DOC, per posterior draw |

Both are fitted to the same data (360 samplings; producers at 16 streams) with the same DAG (@fig-dags). Both account for the measurement error of upstream DOC (@sec-error).
:::

::: {.callout-tip appearance="simple"}
## The research question the SCM answers

**Q1 — Direction.** *Do beaver-engineered streams act as sources or sinks of DOC?* Answered here with delta_SCM, whose outcome is ΔDOC itself (@sec-q1).

**Q2 — Drivers.** *Which abiotic and biotic drivers set the direction of the local DOC change?* Answered with downstream_DOC, which estimates the paths dams → water residence time → primary producers → DOC, solar radiation → producers → DOC, litter → DOC, and upstream DOC → producers, with upstream DOC as the incoming state.

Neither answers Q3 (magnitude), which uses the four-concentration model behind $S$, $R_s$ and $G$.
:::

::: {.callout-important}
## Short answer

-   The two models are one likelihood in two coordinate systems: the mediator effects are identical, and the link upstream DOC → ΔDOC (`r sprintf("%.3f", gme(tD, "bsp_deltaDOC_miDOC_input_z")$mean / s_up)` mg per mg) is the inheritance slope (`r sprintf("%.3f", gme(tO, "bsp_DOCout_miDOC_input")$mean)`) minus one — the net loss per unit inflow and the fraction passed on are the same shortfall read from two sides.
-   They handle the 0.2 mg L$^{-1}$ instrument error differently because ΔDOC is built from two readings: delta_SCM carries the upstream error twice (inside the 0.283 error of its outcome and on the predictor), correlated, which brms cannot model, while downstream_DOC carries each error once.
-   In practice they give the same mediator effects, the same ΔDOC for every sampling and the same predictive accuracy (@sec-practical, @sec-ppc); only the number reported directly differs, net loss per unit inflow or fraction inherited.
:::

# 1. The two DAGs {#sec-dags}

```{r dag-draw-fn}
#| echo: false
draw_dag <- function(tb, outcome, pref, up, up_lab, title) {
  nodes <- data.frame(name = c("Max dam\nheight", "Number of\ndams", "Channel\ngradient", "Water\nresidence time", "Solar\nradiation", "Macrophytes", "Phytoplankton", "Soil litter\ncover", up_lab, outcome),
    x = c(0.9, 0.9, 2.9, 3.2, 4.6, 6.6, 6.2, 6.2, 8.8, 8.8), y = c(6.6, 4.6, 2.0, 5.0, 9.3, 7.2, 5.0, 2.9, 8.8, 5.4))
  rownames(nodes) <- c("dh", "nd", "gr", "wrt", "sol", "mac", "pla", "lit", "up", "out")
  ed <- function(from, to, par, curv = 0, lx = 0, ly = 0.3, new = FALSE) { g <- gme(tb, par)
    tibble(from = from, to = to, label = sprintf("%.2f", g$mean), class = if (new) "new" else if (g$p_pos > 0.9) "pos" else if (g$p_pos < 0.1) "neg" else "none",
           solid = g$p_pos > 0.9 | g$p_pos < 0.1, curv = curv, lx = lx, ly = ly) }
  edges <- bind_rows(
    ed("dh", "wrt", "bsp_waterrestimez_midam_height_z"), ed("nd", "wrt", "b_waterrestimez_n_dams_z", lx = -0.1, ly = 0.35),
    ed("gr", "wrt", "b_waterrestimez_slope_z", lx = 0.45, ly = 0), ed("wrt", "pla", "bsp_planktonabunz_miwater_res_time_z"),
    ed("wrt", "mac", "bsp_macrophyabunz_miwater_res_time_z"), ed("wrt", "out", paste0("bsp_", pref, "_miwater_res_time_z"), curv = 0.18),
    ed("sol", "mac", "bsp_macrophyabunz_misolar_z", curv = -0.15, lx = 0.55, ly = 0.55), ed("sol", "pla", "bsp_planktonabunz_misolar_z", curv = 0.30, lx = -0.85, ly = 0.1),
    ed("sol", "lit", "bsp_coverlitterz_misolar_z", curv = 0.30), ed("mac", "out", paste0("bsp_", pref, "_mimacrophy_abun_z"), ly = 0.32),
    ed("pla", "out", paste0("bsp_", pref, "_miplankton_abun_z"), ly = 0.30), ed("lit", "out", paste0("bsp_", pref, "_micover_litter_z"), lx = -0.2, ly = 0.32),
    ed("up", "out", paste0("bsp_", pref, "_mi", up), lx = 0.45, ly = 0),
    ed("up", "mac", paste0("bsp_macrophyabunz_mi", up), curv = 0.25, lx = 0.35, ly = 0.35, new = TRUE),
    ed("up", "pla", paste0("bsp_planktonabunz_mi", up), curv = -0.35, lx = -0.9, ly = -0.9, new = TRUE))
  edges$label[!edges$solid & edges$class != "new"] <- ""
  a <- merge(edges, nodes, by.x = "from", by.y = "row.names"); a <- merge(a, nodes, by.x = "to", by.y = "row.names", suffixes = c("", "2"))
  dx <- a$x2 - a$x; dy <- a$y2 - a$y; L <- sqrt(dx^2 + dy^2)
  a$x0 <- a$x + dx / L * 0.85; a$y0 <- a$y + dy / L * 0.55; a$x1 <- a$x2 - dx / L * 0.90; a$y1 <- a$y2 - dy / L * 0.55
  a$xm <- (a$x0 + a$x1) / 2 + a$lx; a$ym <- (a$y0 + a$y1) / 2 + a$ly
  pal <- c(pos = "#3B9AD1", neg = "#E8603C", none = "grey55", new = "#8B0000"); arr <- arrow(length = unit(0.16, "cm"), type = "closed")
  pd <- ggplot()
  for (i in seq_len(nrow(a))) { lt <- if (a$class[i] == "new") (if (a$solid[i]) "solid" else "dotted") else if (a$solid[i]) "solid" else "dashed"
    pd <- pd + if (a$curv[i] == 0) geom_segment(data = a[i, ], aes(x = x0, y = y0, xend = x1, yend = y1), colour = pal[a$class[i]], linetype = lt, linewidth = 0.7, arrow = arr) else
      geom_curve(data = a[i, ], aes(x = x0, y = y0, xend = x1, yend = y1), curvature = a$curv[i], colour = pal[a$class[i]], linetype = lt, linewidth = 0.7, arrow = arr) }
  pd + geom_text(data = subset(a, label != ""), aes(x = xm, y = ym, label = label, colour = class), size = 3.8, show.legend = FALSE) +
    scale_colour_manual(values = c(pos = "black", neg = "black", none = "black", new = "#8B0000")) +
    geom_label(data = nodes[rownames(nodes) != "out", ], aes(x = x, y = y, label = name), size = 3.8, label.padding = unit(0.3, "lines"), label.r = unit(0.05, "lines")) +
    geom_label(data = nodes["out", ], aes(x = x, y = y, label = name), size = 4.1, fontface = "bold", label.padding = unit(0.45, "lines"), label.r = unit(0.9, "lines"), linewidth = 0.9, colour = "#C0208E") +
    labs(subtitle = title) + xlim(0, 10) + ylim(1.2, 10) + theme_void() + theme(plot.subtitle = element_text(size = 12, face = "bold"))
}
```

```{r fig-dags}
#| fig-height: 11
#| fig-cap: "**The two models as DAGs, with their fitted coefficients.** Top: delta_SCM — ΔDOC as the outcome; the upstream → ΔDOC coefficient is per SD of upstream DOC. Bottom: downstream_DOC — downstream DOC as the outcome; the upstream → downstream coefficient is the inheritance slope in mg per mg. Dark red: the links upstream DOC → macrophytes and → phytoplankton (per SD, or per mg L⁻¹, of upstream DOC). All other numbers are standardized effects. Solid: posterior probability of the sign above 0.90; dashed/dotted: no-evidence. Every arrow except the one from upstream DOC carries the same number in both graphs."
fig_me <- draw_dag(tD, "ΔDOC", "deltaDOC", "DOC_input_z", "Upstream\nDOC (SD)", "delta_SCM") /
          draw_dag(tO, "Downstream\nDOC", "DOCout", "DOC_input", "Upstream\nDOC (mg/L)", "downstream_DOC")
ggsave(here("results", "figures", "plan", "fig_refit_dags.png"), fig_me, width = 9, height = 11, dpi = 300, bg = "white"); fig_me
```

Max R-hat: `r sprintf("%.3f", ME$delta$rhat)` (delta_SCM), `r sprintf("%.3f", ME$down$rhat)` (downstream_DOC).

# 2. What the link from upstream DOC represents in each model {#sec-upstream}

::: {.callout-note appearance="simple"}
**Question:** What does the link upstream DOC → ΔDOC represent, as against upstream DOC → downstream DOC — and why keep upstream DOC as a separate term at all?
:::

The two arrows from upstream DOC in @fig-dags are one quantity seen from two sides. Write $\beta_{inh}$ for the coefficient of upstream DOC on downstream DOC (downstream_DOC) and $\beta_1$ for its coefficient on ΔDOC (delta_SCM, in mg per mg). Because $\Delta\text{DOC} = \text{DOC}_{down} - \text{DOC}_{up}$,

$$
\beta_1 = \beta_{inh} - 1 .
$$ {#eq-link}

::: {.callout-important}
## What each arrow says

-   $\beta_{inh}$ = `r sprintf("%.2f", gme(tO, "bsp_DOCout_miDOC_input")$mean)` mg per mg is the **fraction of the inflow passed on**, and its distance from one is the removal that scales with the inflow.
-   $\beta_1$ = `r sprintf("%.2f", gme(tD, "bsp_deltaDOC_miDOC_input_z")$mean / s_up)` mg per mg is the **net change per mg entering**, which mixes that removal with the $-1$ the subtraction contributes by construction.
-   Upstream DOC is kept as its own term because it is the one variable beaver engineering cannot cause, so it fixes the inherited baseline that $S$, $R_s$ and $G$ are defined on.
:::

# 3. One model, two parameterizations {#sec-one-model}

::: {.callout-note appearance="simple"}
**Question:** Does the choice of outcome change the inference?
:::

**delta_SCM**, with mediators $M_k$ (macrophytes, phytoplankton, litter cover, water residence time) and a stream random intercept $u$:

$$
\begin{aligned}
\Delta\text{DOC}_i &= \text{DOC}_{down,i} - \text{DOC}_{up,i} \\
\Delta\text{DOC}_i &\sim \text{Student-}t\!\left(\nu,\; \mu^{\Delta}_i,\; \sigma\right) \\
\mu^{\Delta}_i &= \alpha + \beta_1\,\text{DOC}_{up,i} + \sum_k \beta_k M_{k,i} + u_{site(i)}
\end{aligned}
$$ {#eq-delta}

**downstream_DOC**:

$$
\begin{aligned}
\text{DOC}_{down,i} &\sim \text{Student-}t\!\left(\nu,\; \mu_i,\; \sigma\right) \\
\mu_i &= \alpha' + \beta_{inh}\,\text{DOC}_{up,i} + \sum_k \beta'_k M_{k,i} + u'_{site(i)}
\end{aligned}
$$ {#eq-down}

Subtracting $\text{DOC}_{up,i}$ from both sides of @eq-down gives

$$
\text{DOC}_{down,i} - \text{DOC}_{up,i} = \alpha' + (\beta_{inh} - 1)\,\text{DOC}_{up,i} + \sum_k \beta'_k M_{k,i} + u'_{site(i)} + \varepsilon_i ,
$$ {#eq-subtract}

which is @eq-delta with $\beta_1 = \beta_{inh} - 1$ (@eq-link), $\beta_k = \beta'_k$ and the same $\nu$, $\sigma$, $u$. One likelihood, two coordinate systems. The fitted models confirm it:

```{r}
#| label: tbl-equivalence
#| tbl-cap: "The same SCM fitted as delta_SCM (ΔDOC outcome) and downstream_DOC (downstream-DOC outcome), both with the upstream measurement error and the producer links. Posterior mean [95% credible interval]. The inheritance slope implied by delta_SCM is 1 + β₁/SD(upstream DOC)."
#| echo: false
if (have_pred) kable(tribble(
  ~Parameter, ~`delta_SCM: outcome ΔDOC`, ~`downstream_DOC: outcome downstream DOC`,
  "Inheritance slope (mg per mg)", q(1 + cD[, "bsp_deltaDOC_miDOC_input_z"] / s_up), q(cO[, "bsp_DOCout_miDOC_input"]),
  "Upstream DOC → ΔDOC (mg per mg)", q(cD[, "bsp_deltaDOC_miDOC_input_z"] / s_up), q(cO[, "bsp_DOCout_miDOC_input"] - 1),
  "Macrophytes → outcome",         q(cD[, "bsp_deltaDOC_mimacrophy_abun_z"]),  q(cO[, "bsp_DOCout_mimacrophy_abun_z"]),
  "Phytoplankton → outcome",       q(cD[, "bsp_deltaDOC_miplankton_abun_z"]),  q(cO[, "bsp_DOCout_miplankton_abun_z"]),
  "Soil litter cover → outcome",   q(cD[, "bsp_deltaDOC_micover_litter_z"]),   q(cO[, "bsp_DOCout_micover_litter_z"]),
  "Water residence time → outcome",q(cD[, "bsp_deltaDOC_miwater_res_time_z"]), q(cO[, "bsp_DOCout_miwater_res_time_z"]),
  "Upstream DOC → macrophytes (per SD / per mg L⁻¹)", q(cD[, "bsp_macrophyabunz_miDOC_input_z"]), q(cO[, "bsp_macrophyabunz_miDOC_input"]),
  "Upstream DOC → phytoplankton (per SD / per mg L⁻¹)", q(cD[, "bsp_planktonabunz_miDOC_input_z"]), q(cO[, "bsp_planktonabunz_miDOC_input"]),
  "Between-stream SD",   q(cD[, "sd_site__deltaDOC_Intercept"]), q(cO[, "sd_site__DOCout_Intercept"]),
  "Residual scale σ",    q(cD[, "sigma_deltaDOC"]),              q(cO[, "sigma_DOCout"]),
  "Degrees of freedom ν",q(cD[, "nu_deltaDOC"]),                 q(cO[, "nu_DOCout"]),
  "Average ΔDOC (delta_SCM intercept; downstream_DOC evaluated at mean upstream DOC)",
     q(cD[, "b_deltaDOC_Intercept"]), q(cO[, "b_DOCout_Intercept"] + (cO[, "bsp_DOCout_miDOC_input"] - 1) * mu_up))) else cat("pending")
```

::: {.callout-important}
## Key message of @tbl-equivalence

-   Mediator effects, producer links, stream SD, residual scale and tail parameter are the same in the two models.
-   The two readings of the inheritance differ only at the second decimal, from the error correlation of @sec-error.
-   **Implication.** The choice of outcome sets which number is reported first, not what is inferred.
:::

# 4. Propagation of the measurement error in the two models {#sec-error}

::: {.callout-note appearance="simple"}
**Question:** Should both models account for the 0.2 mg L$^{-1}$ instrument error of upstream DOC, and do they propagate it in the same way?
:::

Both do now: in each model, the upstream reading is a noisy measurement of a latent true concentration (error 0.2 mg L$^{-1}$, or `r sprintf("%.3f", 0.2 / s_up)` on the standardized scale), and the latent value — not the reading — enters the outcome equation and the two producer equations. But the two models cannot propagate it in the same way, and the reason is the definition of ΔDOC:

-   **delta_SCM.** The outcome is built from two readings, so it carries their combined error, $\sqrt{0.2^2 + 0.2^2} = 0.283$ mg L$^{-1}$. The upstream reading is *also* the predictor, with its own 0.2 error. The same physical error therefore appears twice — once inside the outcome's error, once on the predictor — and the two are negatively correlated (a reading that is too high makes ΔDOC too low and the predictor too high). brms models the two errors as independent; the correlation is ignored.
-   **downstream_DOC.** The outcome is one reading with its own error (0.2); the upstream reading is the predictor with its own error (0.2). Each error appears exactly once, and independence is correct.

Does it matter? Before the upstream error was added, the inheritance slope was 0.945; with it, `r sprintf("%.3f", gme(tO, "bsp_DOCout_miDOC_input")$mean)` in downstream_DOC and `r sprintf("%.3f", 1 + gme(tD, "bsp_deltaDOC_miDOC_input_z")$mean / s_up)` implied by delta_SCM — the predicted ~1% attenuation (reliability of upstream DOC `r sprintf("%.3f", s_up^2 / (s_up^2 + 0.04))`), and the second-decimal gap between the two is the ignored correlation. The mediator effects do not move. So: a real difference in how the error is handled, correctly handled in one model and approximately in the other, and no practical consequence at this error size.

# 5. A practical comparison {#sec-practical}

::: {.callout-note appearance="simple"}
**Question:** In practice, what does each model give directly, and how well does each predict?
:::

```{r practical}
#| echo: false
if (have_pred) {
  m <- function(x) sprintf("%.2f", x)
  kable(tribble(
    ~` `, ~`delta_SCM`, ~`downstream_DOC`,
    "Number reported directly for upstream DOC", "net change per unit inflow: −0.05 mg per mg", "fraction inherited: 0.94 mg per mg",
    "Average ΔDOC", "the intercept, directly", "one subtraction: intercept + (β_inh − 1) × mean upstream DOC",
    "Mediator effects on the local change", "directly", "directly (identical numbers)",
    "Effect of the inflow on the producers (new links)", "directly", "directly (identical signs)",
    "ΔDOC of each sampling", "the fitted value", "one subtraction: predicted downstream DOC − upstream DOC",
    "Out-of-sample accuracy for its own outcome (LOO R², RMSE, 90% coverage)", sprintf("ΔDOC: R² %s, RMSE %s (mean-only %s), %s%%", m(P$D$metrics["R2"]), m(P$D$metrics["RMSE"]), m(P$D$metrics["RMSE_null"]), sprintf("%.0f", 100 * P$D$metrics["cover90"])),
      sprintf("downstream DOC: R² %s, RMSE %s (mean-only %s), %s%%", m(P$O$metrics["R2"]), m(P$O$metrics["RMSE"]), m(P$O$metrics["RMSE_null"]), sprintf("%.0f", 100 * P$O$metrics["cover90"])),
    "Out-of-sample accuracy for ΔDOC", sprintf("R² %s, RMSE %s", m(P$D$metrics["R2"]), m(P$D$metrics["RMSE"])), sprintf("R² %s, RMSE %s (derived)", m(P$O$ddoc_metrics["R2"]), m(P$O$ddoc_metrics["RMSE"])),
    "Measurement error", "outcome 0.283 + upstream 0.2, correlated (ignored)", "outcome 0.2 + upstream 0.2, independent (correct)",
    "Question it answers here", "Q1: direction of the local change", "Q2: drivers, and the inherited state used by S, R_s and G"),
    caption = sprintf("What each model gives, and how well it predicts (PSIS-LOO on the observed samplings; max R-hat %.3f and %.3f).", P$rhat["delta"], P$rhat["down"]))
} else cat("pending")
```

**Neither model predicts ΔDOC out of sample.** delta_SCM, which has ΔDOC as its outcome, does no better than its mean, and so does the derived ΔDOC of downstream_DOC: a property of the system, not of the parameterization.

# 6. Pros and cons

::: {.callout-note appearance="simple"}
**Question:** What does each parameterization give, and what does it cost?
:::

**delta_SCM — ΔDOC as the outcome**

-   *Pro:* models the process of interest directly; the outcome *is* the local change, and its average is the intercept.
-   *Pro:* the posterior predictive distribution of its outcome gives the sink / negligible / source shares of Q1 directly (@sec-q1).
-   *Con:* the upstream coefficient mixes the passed-on fraction with the inflow-scaled removal (@sec-upstream), which invites reading it as "high upstream DOC causes greater retention".
-   *Con:* the upstream error appears twice, correlated, and the correlation is ignored (@sec-error).
-   *Con:* the inherited state is not a quantity of the model; $S$, $R_s$, $G$ and the spatial result then rest on a concept the SCM does not state.

**downstream_DOC — downstream DOC as the outcome**

-   *Pro:* the upstream coefficient is the inheritance, a directly reportable number with units, and its shortfall is the inflow-scaled removal.
-   *Pro:* the mediators read as what beaver engineering *adds* to an inherited state.
-   *Pro:* each measurement error appears once.
-   *Con:* ΔDOC is one subtraction away; a reader who wants "the model of the local change" has to follow it (@sec-derived).
-   *Con:* it requires re-describing the Methods equation and re-drawing the SCM figure.

Q1 is answered with delta_SCM and Q2 with downstream_DOC: each question is read off the parameterization that states its quantity directly, and @tbl-equivalence shows that this splits the reporting, not the inference.

# 7. Predictive accuracy of the two models {#sec-ppc}

::: {.callout-note appearance="simple"}
**Question:** How well does each model predict its own outcome, and ΔDOC, out of sample — predicted against observed?
:::

Pareto-smoothed importance-sampling leave-one-out cross-validation (PSIS-LOO): each observed value is predicted as if it had been left out of the fit. delta_SCM predicts ΔDOC; downstream_DOC predicts downstream DOC, and ΔDOC by subtraction.

```{r fig-ppc}
#| fig-height: 4.3
#| fig-cap: "**Predicted vs observed, both models (leave-one-out).** (a) delta_SCM: LOO-predicted ΔDOC against the observed ΔDOC of each sampling. (b) downstream_DOC: LOO-predicted downstream DOC against the observed concentration. (c) downstream_DOC: ΔDOC derived from the LOO prediction against the observed ΔDOC. Segments: 90% LOO predictive interval where available; dashed: 1:1."
if (have_pred) {
  pp <- function(x, y, lo = NULL, hi = NULL, xl, yl) { d <- tibble(x, y); p <- ggplot(d, aes(x, y)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45")
    if (!is.null(lo)) p <- p + geom_linerange(aes(ymin = lo, ymax = hi), colour = "grey70", alpha = .6)
    p + geom_point(size = 1.2, colour = "#3B9AD1", alpha = .8) + labs(x = xl, y = yl) + theme_bw(base_size = 11) + theme(panel.grid.minor = element_blank()) }
  ib <- match(P$both, P$obsO)
  pa <- pp(P$d_obs, P$D$mu_loo, P$D$q_loo[1, ], P$D$q_loo[2, ], expression("Observed "*Delta*"DOC (mg "*L^-1*")"), expression("delta_SCM: LOO-predicted "*Delta*"DOC"))
  pb <- pp(P$y_out, P$O$mu_loo, P$O$q_loo[1, ], P$O$q_loo[2, ], expression("Observed downstream DOC (mg "*L^-1*")"), "downstream_DOC: LOO-predicted downstream DOC")
  pc <- pp(P$d_obs, P$O$mu_loo[ib] - P$up_both, NULL, NULL, expression("Observed "*Delta*"DOC (mg "*L^-1*")"), expression("downstream_DOC: derived "*Delta*"DOC"))
  pa + pb + pc + plot_annotation(tag_levels = "a")
}
```

```{r ppc-table}
#| echo: false
if (have_pred) { m <- function(x) sprintf("%.2f", x)
  kable(tribble(
    ~Model, ~`Quantity predicted`, ~`LOO R²`, ~`LOO RMSE (mg L⁻¹)`, ~`RMSE of predicting the mean`, ~`90% interval coverage`,
    "delta_SCM", "ΔDOC (its outcome)", m(P$D$metrics["R2"]), m(P$D$metrics["RMSE"]), m(P$D$metrics["RMSE_null"]), sprintf("%.0f%%", 100 * P$D$metrics["cover90"]),
    "downstream_DOC", "downstream DOC (its outcome)", m(P$O$metrics["R2"]), m(P$O$metrics["RMSE"]), m(P$O$metrics["RMSE_null"]), sprintf("%.0f%%", 100 * P$O$metrics["cover90"]),
    "downstream_DOC", "ΔDOC (derived)", m(P$O$ddoc_metrics["R2"]), m(P$O$ddoc_metrics["RMSE"]), m(P$O$ddoc_metrics["RMSE_null"]), "—"),
    caption = "Out-of-sample accuracy of the two models (PSIS-LOO on the observed samplings).") }
```

::: {.callout-important}
## Key message of @fig-ppc

`r if (have_pred) sprintf("- Both models predict ΔDOC out of sample no better than its mean (delta_SCM R² %.2f; downstream_DOC derived R² %.2f), because the local change is small and stream-specific (Q1).", P$D$metrics["R2"], P$O$ddoc_metrics["R2"]) else "- pending"`
`r if (have_pred) sprintf("- downstream_DOC predicts downstream DOC with R² %.2f, RMSE %.2f against %.2f for the mean, and calibrated 90%% intervals (%.0f%% coverage), almost all of it inheritance.", P$O$metrics["R2"], P$O$metrics["RMSE"], P$O$metrics["RMSE_null"], 100 * P$O$metrics["cover90"]) else ""`
-   **Implication.** Predictive accuracy does not separate the two parameterizations: the same information is in both, and what can and cannot be predicted is a property of the system.
:::

# 8. The a-posteriori derivation of ΔDOC {#sec-derived}

::: {.callout-note appearance="simple"}
**Question:** How is ΔDOC estimated a posteriori from a model that never sees it, and is it the same ΔDOC that delta_SCM fits directly?
:::

ΔDOC is computed after fitting, once per posterior draw $d$ and sampling $i$, from two numbers the draw already contains:

$$
\begin{aligned}
\Delta\text{DOC}^{(d)}_i &= \mu^{(d)}_i - \text{DOC}_{up,i} \\
&= \alpha'^{(d)} + \left(\beta_{inh}^{(d)} - 1\right)\text{DOC}_{up,i} + \sum_k \beta'^{(d)}_k M_{k,i} + u'^{(d)}_{site(i)}
\end{aligned}
$$ {#eq-derived}

(with the latent upstream value of the draw). The seasonal average is the same operation averaged over the samplings of a season:

$$
\overline{\Delta\text{DOC}}^{(d)}_{s} = \frac{1}{n_s}\sum_{i \in s} \Delta\text{DOC}^{(d)}_i
$$ {#eq-avg}

```{r fig-ddoc}
#| fig-height: 4.5
#| fig-cap: "**ΔDOC from the two models.** (a) The ΔDOC of every sampling: derived from downstream_DOC (@eq-derived) against fitted directly by delta_SCM; posterior medians; dashed: 1:1. (b) The seasonal average ΔDOC from both models (@eq-avg); point: median, bars: 66% and 95% credible intervals; open diamonds: the paired-contrast model on the concentrations alone (+0.14 mg L⁻¹ summer, 0.00 winter)."
if (have_pred) {
  pa <- ggplot(tibble(x = P$D$dd_med, y = P$O$dd_med), aes(x, y)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
    geom_hline(yintercept = 0, colour = "grey80") + geom_vline(xintercept = 0, colour = "grey80") +
    geom_point(size = 1.4, colour = "#3B9AD1", alpha = .8) +
    labs(x = expression("Fitted "*Delta*"DOC, delta_SCM (mg "*L^-1*")"), y = expression("Derived "*Delta*"DOC, downstream_DOC (mg "*L^-1*")")) +
    theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank())
  av <- bind_rows(as_tibble(P$D$avg_season) |> mutate(model = "delta_SCM"), as_tibble(P$O$avg_season) |> mutate(model = "downstream_DOC")) |>
    pivot_longer(c(summer, winter), names_to = "season", values_to = "v") |> mutate(season = factor(season, levels = c("winter", "summer")))
  q1 <- tibble(season = factor(c("winter", "summer"), levels = c("winter", "summer")), v = c(0.00, 0.14))
  pb <- ggplot(av, aes(x = v, y = season, colour = model)) + geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
    stat_pointinterval(.width = c(.66, .95), point_interval = "median_qi", position = position_dodge(width = .5)) +
    geom_point(data = q1, aes(x = v, y = season), inherit.aes = FALSE, shape = 23, size = 3, fill = "white", colour = "#E8603C", stroke = 1.2) +
    scale_colour_manual(values = c(delta_SCM = "#E8603C", downstream_DOC = "#3B9AD1"), name = NULL) +
    labs(x = expression("Seasonal average "*Delta*"DOC (mg "*L^-1*")"), y = NULL) +
    theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank(), legend.position = "top")
  pa + pb + plot_annotation(tag_levels = "a")
}
```

```{r ddoc-numbers}
#| echo: false
if (have_pred) {
  ddoc_msg1 <- sprintf("- The ΔDOC of every sampling is the same in the two models (correlation %.3f, median absolute difference %.3f mg L⁻¹): the subtraction recovers what delta_SCM fits.", cor(P$D$dd_med, P$O$dd_med), median(abs(P$D$dd_med - P$O$dd_med)))
  ddoc_msg2 <- sprintf("- Seasonal averages: delta_SCM %s (summer) and %s (winter); downstream_DOC %s and %s; the paired-contrast model of Q1, from the concentrations alone, +0.14 [0.05, 0.22] and 0.00 [−0.06, 0.07].", qm(P$D$avg_season[, "summer"]), qm(P$D$avg_season[, "winter"]), qm(P$O$avg_season[, "summer"]), qm(P$O$avg_season[, "winter"]))
} else ddoc_msg1 <- ddoc_msg2 <- "- pending"
```

::: {.callout-important}
## Key message of @fig-ddoc

`r ddoc_msg1`
`r ddoc_msg2`
-   **Implication.** Nothing about ΔDOC is lost by not making it the outcome; the derived quantity is the fitted quantity.
:::

# 9. Answering Q1 with the extended models {#sec-q1}

::: {.callout-note appearance="simple"}
**Question:** Q1 (direction) has so far been answered with a separate paired-contrast model on the concentrations; can the extended SCMs answer it too, so that one model serves both questions?
:::

Q1 asks two things: the average ΔDOC per season, and the posterior predictive shares of sink, negligible and source responses for a single sampling (ROPE ±0.5 mg L$^{-1}$). Both come from either extended model: the seasonal average from @eq-avg, and the shares from the posterior predictive distribution of ΔDOC for a new sampling — directly in delta_SCM, and as (predicted downstream DOC − upstream DOC) in downstream_DOC. delta_SCM is the model that states ΔDOC as its outcome, so it is the one used for Q1; @fig-dag-q1 is its graph.

```{r fig-dag-q1}
#| fig-height: 5.5
#| fig-cap: "**The model behind Q1: delta_SCM, with ΔDOC as the outcome.** Same graph and same data as the Q2 model, with the local change as the response instead of the downstream concentration; the coefficient on upstream DOC is per SD. Dark red: the links upstream DOC → macrophytes and → phytoplankton. Solid: posterior probability of the sign above 0.90; dashed/dotted: no-evidence."
draw_dag(tD, "ΔDOC", "deltaDOC", "DOC_input_z", "Upstream\nDOC (SD)", "delta_SCM — the model for Q1")
```

```{r q1-from-scm}
#| echo: false
if (have_pred) {
  rope <- 0.5
  shares <- function(PP, season, lab) { bind_rows(lapply(c("winter", "summer"), function(s) { X <- PP[, season == s]
    tibble(model = lab, season = s, sink = rowMeans(X < -rope), negligible = rowMeans(abs(X) <= rope), source = rowMeans(X > rope)) })) }
  ppD <- P$D$pp_sub[, P$both]; ppO <- P$O$pp_sub[, P$both] - matrix(P$up_both, nrow(P$O$pp_sub), length(P$both), byrow = TRUE)
  sh <- bind_rows(shares(ppD, P$season_both, "delta_SCM"), shares(ppO, P$season_both, "downstream_DOC")) |>
    pivot_longer(c(sink, negligible, source), names_to = "response", values_to = "p") |>
    mutate(response = factor(response, levels = c("sink", "negligible", "source")), season = factor(season, levels = c("winter", "summer")))
  ref <- tibble(season = rep(c("winter", "summer"), each = 3), response = rep(c("sink", "negligible", "source"), 2), p = c(.12, .77, .11, .09, .75, .15)) |>
    mutate(response = factor(response, levels = c("sink", "negligible", "source")), season = factor(season, levels = c("winter", "summer")))
  shs <- sh |> group_by(model, season, response) |> summarise(med = median(p), .groups = "drop")
  gs <- function(mo, se, re) shs$med[shs$model == mo & shs$season == se & shs$response == re]
}
```

```{r fig-q1}
#| fig-height: 4.5
#| fig-cap: "**Q1 answered from the two extended SCMs.** Posterior predictive shares of a single sampling showing a sink (ΔDOC < −0.5 mg L⁻¹), a negligible (within ±0.5) or a source (> 0.5) response, by season and model; point: median, bars: 66% and 95% credible intervals. Open diamonds: the shares given by the paired-contrast model on the concentrations alone (winter 12 / 77 / 11%, summer 9 / 75 / 15%)."
if (have_pred) ggplot(sh, aes(x = p, y = response, colour = model)) +
  stat_pointinterval(.width = c(.66, .95), point_interval = "median_qi", position = position_dodge(width = .5)) +
  geom_point(data = ref, aes(x = p, y = response), inherit.aes = FALSE, shape = 23, size = 3, fill = "white", colour = "black", stroke = 1) +
  facet_wrap(~ season) + scale_colour_manual(values = c(delta_SCM = "#E8603C", downstream_DOC = "#3B9AD1"), name = NULL) +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  labs(x = "Posterior predictive share of a single sampling", y = NULL) + theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank(), legend.position = "top")
```

::: {.callout-important}
## Key message of @fig-q1

`r if (have_pred) sprintf("- The extended SCMs reproduce the Q1 answer: negligible responses in %.0f%% (winter) and %.0f%% (summer) of single samplings under downstream_DOC (delta_SCM %.0f%% and %.0f%%), against 77%% and 75%% from the paired-contrast model, with sources slightly more likely than sinks in summer in every model.", 100*gs("downstream_DOC","winter","negligible"), 100*gs("downstream_DOC","summer","negligible"), 100*gs("delta_SCM","winter","negligible"), 100*gs("delta_SCM","summer","negligible")) else "- pending"`
-   The seasonal averages are those of @fig-ddoc b: the same ordering as the paired-contrast model, with overlapping intervals, from a model that also carries the drivers.
-   **Implication (the option).** Q1 can be answered from delta_SCM alone, at the price that its numbers then depend on the SCM specification, with the mediators imputed at 342 of 360 samplings.
:::

# 10. Dependence of the derived ΔDOC on the model specification {#sec-spec}

::: {.callout-note appearance="simple"}
**Question:** If ΔDOC is derived from the model's prediction, does adding or removing a variable from the SCM change it — and bias the inference built on it? Would it be safer to model ΔDOC directly from the data?
:::

The concern is right in general, and it is answered in two parts.

**No conclusion is drawn from the SCM's ΔDOC.** Two different quantities are called ΔDOC:

| | Where it comes from | Depends on the SCM's variables? | Used for |
|---|---|---|---|
| **ΔDOC from the data** | the paired-contrast measurement-error model on the raw concentrations (Q1) and the four-concentration model behind $S$, $R_s$, $B$, $G$ (Q3) — no mediators, no drivers | **no** | every conclusion on direction, magnitude and seasonal gain |
| **ΔDOC from the SCM** | $\mu_i - \text{DOC}_{up,i}$, the SCM's *expected* local modification given the drivers (@eq-derived) | **yes**, as any fitted value does | interpreting Q2: which pathways contribute (the decomposition), and the checks in this document |

So ΔDOC is already modelled directly from the data wherever a conclusion about ΔDOC is drawn. The SCM's ΔDOC is a fitted value, and fitted values depend on the specification in both parameterizations equally, because they are the same likelihood (@eq-subtract; @fig-ddoc a). How far they move when the structure changes is measured by the structural-sensitivity analysis in `analysis_plan.qmd`, section 9 (alternative DAGs, unmeasured confounding, priors).

# 11. The gain: the meaning of the coefficient on upstream DOC {#sec-gain}

::: {.callout-note appearance="simple"}
**Question:** What do we actually gain with downstream DOC as the outcome instead of ΔDOC?
:::

The gain is the meaning of one number (@sec-upstream): `r sprintf("%.2f", gme(tD, "bsp_deltaDOC_miDOC_input_z")$mean)` per SD on ΔDOC becomes `r sprintf("%.2f", gme(tO, "bsp_DOCout_miDOC_input")$mean)` mg per mg inherited, whose distance from one *is* the removal that scales with the inflow. With it come the inherited state as a stated quantity, one appearance of each measurement error (@sec-error), and an inflow-through-producers effect of a variable that is not also inside the outcome (@sec-inflow). Not gained: fit, mediator estimates, predictive accuracy (@sec-practical).

# 12. The inflow through the producers: direct, indirect and total effects {#sec-inflow}

::: {.callout-note appearance="simple"}
**Question:** With the links upstream DOC → producers in the DAG, does the baseline modulate the beaver modification through the producers, and by how much?
:::

```{r inflow-effects}
#| echo: false
if (have_pred) {
  dir_O <- cO[, "bsp_DOCout_miDOC_input"]
  indO_mac <- cO[, "bsp_macrophyabunz_miDOC_input"] * cO[, "bsp_DOCout_mimacrophy_abun_z"]
  indO_pla <- cO[, "bsp_planktonabunz_miDOC_input"] * cO[, "bsp_DOCout_miplankton_abun_z"]
  indD_mac <- cD[, "bsp_macrophyabunz_miDOC_input_z"] * cD[, "bsp_deltaDOC_mimacrophy_abun_z"] / s_up
  indD_pla <- cD[, "bsp_planktonabunz_miDOC_input_z"] * cD[, "bsp_deltaDOC_miplankton_abun_z"] / s_up
  dir_D <- cD[, "bsp_deltaDOC_miDOC_input_z"] / s_up
  ind_O <- indO_mac + indO_pla; ind_D <- indD_mac + indD_pla
  tot_O <- dir_O - 1 + ind_O; tot_D <- dir_D + ind_D
  eff <- bind_rows(
    tibble(model = "downstream_DOC", effect = "direct on downstream DOC (inheritance)", v = dir_O),
    tibble(model = "downstream_DOC", effect = "indirect via macrophytes", v = indO_mac),
    tibble(model = "downstream_DOC", effect = "indirect via phytoplankton", v = indO_pla),
    tibble(model = "downstream_DOC", effect = "total on ΔDOC", v = tot_O),
    tibble(model = "delta_SCM", effect = "direct on ΔDOC", v = dir_D),
    tibble(model = "delta_SCM", effect = "indirect via macrophytes", v = indD_mac),
    tibble(model = "delta_SCM", effect = "indirect via phytoplankton", v = indD_pla),
    tibble(model = "delta_SCM", effect = "total on ΔDOC", v = tot_D))
  inflow_msg1 <- sprintf("- **The links are supported for macrophytes, not for phytoplankton.** Upstream DOC → macrophytes %s per mg L⁻¹ (P > 0 = %.2f); upstream DOC → phytoplankton %s (P > 0 = %.2f); the same signs in delta_SCM.", q(cO[, "bsp_macrophyabunz_miDOC_input"]), mean(cO[, "bsp_macrophyabunz_miDOC_input"] > 0), q(cO[, "bsp_planktonabunz_miDOC_input"]), mean(cO[, "bsp_planktonabunz_miDOC_input"] > 0))
  inflow_msg2 <- sprintf("- **The indirect route is an order of magnitude below the direct one.** Indirect effect of the inflow on ΔDOC through the producers: %s mg per mg (downstream_DOC), %s (delta_SCM); total effect of the inflow on ΔDOC: %s and %s.", qm(ind_O), qm(ind_D), qm(tot_O), qm(tot_D))
} else inflow_msg1 <- inflow_msg2 <- "- pending"
```

```{r fig-inflow}
#| fig-height: 4
#| fig-cap: "**How the inflow reaches the local change, in mg per mg of upstream DOC, in both models.** Posterior distributions of the direct effect of upstream DOC (on downstream DOC in downstream_DOC — the inheritance; on ΔDOC in delta_SCM), the two indirect effects through the producers, and the total effect on ΔDOC; point: median, bars: 66% and 95% credible intervals; dashed: 0 and 1."
if (have_pred) {
  lv <- c("total on ΔDOC", "indirect via phytoplankton", "indirect via macrophytes", "direct on ΔDOC", "direct on downstream DOC (inheritance)")
  ggplot(eff |> mutate(effect = factor(effect, levels = lv)), aes(x = v, y = effect, colour = model)) +
    geom_vline(xintercept = c(0, 1), linetype = "dashed", colour = "grey45") +
    stat_pointinterval(.width = c(.66, .95), point_interval = "median_qi", position = position_dodge(width = .5)) +
    scale_colour_manual(values = c(delta_SCM = "#E8603C", downstream_DOC = "#3B9AD1"), name = NULL) +
    labs(x = "mg L⁻¹ of DOC per mg L⁻¹ of upstream DOC", y = NULL) + theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank(), legend.position = "top")
}
```

::: {.callout-important}
## Key message of @fig-inflow

`r inflow_msg1`
`r inflow_msg2`
-   **Implication.** The baseline does modulate the beaver modification, through macrophytes only, and the modulation is small against the direct shortfall: the more DOC enters, the lower the delta (about −0.05 mg per mg), and the producers do not reverse that. The two models agree on every number; only downstream_DOC names the inheritance.
:::

# 13. The target quantity and the figures {#sec-target}

::: {.callout-note appearance="simple"}
**Question:** Our interest is the effect of beaver-engineering on ΔDOC, not on downstream DOC, which depends almost entirely on upstream DOC; we want to see how the delta changes with the input. And if the model is now on downstream DOC, should the figures not show downstream DOC?
:::

The target is the local modification, and the parameterization does not change it:

-   **The estimand is the local modification.** Every effect in downstream_DOC — the mediator coefficients, the derived ΔDOC, the pathway decomposition — is an effect on $\mu_i - \text{DOC}_{up,i}$, identical draw for draw to the fitted ΔDOC of delta_SCM (@fig-ddoc a). Downstream DOC is the *response* of the likelihood, not the object of inference.
-   **"How the delta changes with the input" is one coefficient, estimated in both models.** It is $\beta_1 = \beta_{inh} - 1$ = `r if (have_pred) qm(cO[, "bsp_DOCout_miDOC_input"] - 1) else ""` mg per mg (downstream_DOC) and `r if (have_pred) qm(cD[, "bsp_deltaDOC_miDOC_input_z"] / s_up) else ""` (delta_SCM): the delta falls as the input rises — the beaver system dampens a rich inflow — plus the producer route of @sec-inflow. @fig-target shows it on the data.
-   **The figures.** The graph of the Q2 model (@fig-dags, bottom) has downstream DOC as its outcome node because that is what it predicts, and the predicted-versus-observed check (@fig-ppc b) is on downstream DOC for the same reason. Everything reported as a result — the direction shares (@fig-q1), the seasonal averages (@fig-ddoc b), $R_s$ and $G$, the pathway decomposition — is on ΔDOC or on quantities built from it.

```{r fig-target}
#| fig-height: 4.5
#| fig-cap: "**The local modification as a function of the input.** ΔDOC of every sampling (posterior median; derived from downstream_DOC) against the observed upstream DOC; black line and band: the expected ΔDOC at the mean of the mediators, α + (β_inh − 1)·DOC_up, posterior median and 95% credible interval; dashed: zero. The slope of the line is β₁ in mg per mg."
if (have_pred) {
  dm <- P$O$dd_med; up <- P$up_both
  grid <- seq(min(up), max(up), length.out = 80)
  line <- sapply(grid, function(u) cO[, "b_DOCout_Intercept"] + (cO[, "bsp_DOCout_miDOC_input"] - 1) * u)
  ln <- tibble(up = grid, med = apply(line, 2, median), lo = apply(line, 2, quantile, .025), hi = apply(line, 2, quantile, .975))
  ggplot(tibble(up, dm), aes(up, dm)) + geom_hline(yintercept = 0, linetype = "dashed", colour = "grey45") +
    geom_ribbon(data = ln, aes(x = up, ymin = lo, ymax = hi), inherit.aes = FALSE, fill = "black", alpha = .15) +
    geom_line(data = ln, aes(up, med), colour = "black", linewidth = 0.9) +
    geom_point(size = 1.3, colour = "#3B9AD1", alpha = .8) +
    labs(x = expression("Upstream DOC (mg "*L^-1*")"), y = expression(Delta*"DOC, posterior median (mg "*L^-1*")")) +
    theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank())
}
```

```{r target-numbers}
#| echo: false
if (have_pred) {
  b1 <- cO[, "bsp_DOCout_miDOC_input"] - 1
  tgt_msg1 <- sprintf("- The local change shifts by %s mg L⁻¹ per additional mg L⁻¹ of DOC entering the stream, so a rich inflow is dampened and a poor one enriched.", qm(b1))
  tgt_msg2 <- sprintf("- The zero crossing sits at %.1f mg L⁻¹ of upstream DOC: below it the expected ΔDOC is positive, above it negative.", median(-cO[, "b_DOCout_Intercept"] / b1))
} else tgt_msg1 <- tgt_msg2 <- "- pending"
```

::: {.callout-important}
## Key message of @fig-target

`r tgt_msg1`
`r tgt_msg2`
-   **Implication.** The delta is read against the input, and the input is a term of the model, which is why upstream DOC is kept explicit rather than folded into the outcome.
:::

# 14. Reproducibility

```{r session}
#| echo: false
cat("Models: delta_SCM = results/models_fit/delta_SCM_me.rds, downstream_DOC = results/models_fit/downstream_DOC_me.rds (notebook/R/me_refits.R)\n",
    "Predictions and LOO: notebook/R/me_predictions.R -> results/me_predictions.rds\n")
sessionInfo()$R.version$version.string
```

# References

::: {#refs}
:::
