---
title: "Option: rebuilding the SCM with downstream DOC as the target"
subtitle: "Josh's proposal, tested on the real data, with a cost and risk assessment"
author: "Leonardo Capitani"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    embed-resources: true
    fig-width: 9
    fig-align: center
execute:
  echo: true
  message: false
  warning: false
editor: visual
---

```{r setup}
#| include: false
library(tidyverse); library(brms); library(ggdist); library(patchwork)
library(knitr); library(here); library(glue)
invisible(Sys.setlocale("LC_CTYPE", "en_US.UTF-8"))  # so non-ASCII prints correctly
OK <- c(blue="#0072B2", vermillion="#D55E00", green="#009E73", amber="#E69F00")
```

::: {.callout-tip appearance="simple"}
## The proposal being evaluated

> "Could we try one more thing instead, which is to change the target variable from delta DOC to just the downstream DOC concentration? Upstream DOC would then still feed into this. We can then estimate delta DOC directly from the posteriors, winter and summer separately etc. That is: instead of using B as the target variable, retain winter and summer DOC as the target, and then use the posterior distributions to get distributions of B, gamma, lambda etc."
>
> — Josh Larsen

This document tests that proposal rather than describing it. **The short answer: the idea is correct, and the correct implementation of it is already what the manuscript does.** The free-slope version is not just expensive, it degrades the estimate. Evidence below.
:::

# 1. What the proposal changes

The current analysis has two separate models:

| Model | Target | Purpose |
|------------------|------------------|-------------------------------------|
| Q1 measurement-error model | raw upstream and downstream DOC, season × location | direction of ΔDOC |
| Q2 structural causal model | ΔDOC with propagated error 0.283 mg L⁻¹ | drivers |

Josh's proposal merges them: one model with **downstream DOC as the target**, upstream DOC as a parent, season-specific paths, and every contrast (ΔDOC, B, S, G, M) derived from the posterior. Conceptually this is attractive: the contrasts inherit full uncertainty instead of being computed from point values.

# 2. The critical question: what does "upstream DOC feeds into this" mean?

There are two ways to let upstream DOC enter the downstream equation, and they are **different estimands**:

$$\text{(a) free slope: } \;\text{DOC}_{down} = \alpha_s + \beta_s \,\text{DOC}_{up} + \dots \qquad
\text{(b) offset: }\; \text{DOC}_{down} = \alpha_s + 1\cdot\text{DOC}_{up} + \dots$$

Under (b), $\alpha_s$ **is** ΔDOC in season *s*. Under (a) there is no single ΔDOC: the implied difference $\alpha_s + (\beta_s - 1)\,\text{DOC}_{up}$ depends on how much DOC the stream was carrying. The paired design was chosen precisely to remove that dependence.

So the proposal only reproduces the intended quantity if $\beta \approx 1$. Let us check.

```{r data}
m <- read.csv(here("data","processed","m6.csv"))
wide <- m |> select(site, season, up = DOC_input, dn = DOC_out) |>
  filter(!is.na(up), !is.na(dn)) |>
  mutate(season = factor(season), site = factor(site))
glue("Observations: {nrow(wide)} | sites: {nlevels(wide$site)} | mean upstream DOC: {round(mean(wide$up),2)} mg/L")
```

```{r slope}
#| results: hide
f_slope <- brm(bf(dn | mi(0.2) ~ 0 + season + season:up, family = student()),
  data = wide,
  prior = c(prior(normal(0, 2), class = "b"), prior(student_t(3, 0, 2.5), class = "sigma"),
            prior(gamma(2, 0.1), class = "nu")),
  chains = 4, cores = 4, iter = 3000, warmup = 1000, seed = 7,
  backend = "cmdstanr", file = here("results","models_fit","inheritance_slope"), refresh = 0)
```

```{r slope-table}
#| echo: false
ds <- as_draws_df(f_slope)
kable(bind_rows(lapply(c("summer","winter"), function(s){
  b <- ds[[paste0("b_season", s, ":up")]]
  tibble(Season = s, `Posterior mean slope` = round(mean(b), 2),
         `2.5%` = round(quantile(b, .025), 2), `97.5%` = round(quantile(b, .975), 2),
         `P(0.9 < slope < 1.1)` = round(mean(b > 0.9 & b < 1.1), 2))})),
  caption = "Bayesian robust regression of downstream on upstream DOC (Student-t likelihood, 0.2 mg/L error on the response).")
```

::: {.callout-note}
## Inheritance really is close to one-to-one

The posterior slope of downstream on upstream DOC is **0.93 [0.88, 0.97] in summer and 0.95 [0.92, 0.99] in winter**. A heavy-tailed likelihood handles the few extreme upstream concentrations without trimming anything.

Downstream DOC tracks upstream DOC essentially one-for-one, so Josh's premise is empirically correct. The slope is nonetheless slightly *below* one, which is itself interpretable: at high incoming concentrations the reach retains proportionally a little more DOC than it passes on.
:::

# 3. Three pilot fits

All three use the same data, the same Student-*t* likelihood, the same site random intercept and the same 0.2 mg L⁻¹ instrument error on the response.

```{r fits}
#| results: hide
# (a) FREE SLOPE, upstream as a measurement-error covariate -- the literal proposal
f_free <- brm(bf(dn | mi(0.2) ~ 0 + season + season:me(up, 0.2) + (1|site), family = student()),
  data = wide,
  prior = c(prior(normal(0, 1), class = "b"),
            prior(normal(1, 0.3), class = "b", coef = "seasonsummer:meup0.2"),
            prior(normal(1, 0.3), class = "b", coef = "seasonwinter:meup0.2"),
            prior(student_t(3, 0, 2.5), class = "sigma"), prior(gamma(2, 0.1), class = "nu"),
            prior(normal(0, 1), class = "sd"),
            prior(normal(2.5, 2), class = "meanme"), prior(normal(0, 3), class = "sdme")),
  chains = 4, cores = 4, iter = 3000, warmup = 1000, seed = 7,
  control = list(adapt_delta = 0.95, max_treedepth = 12),
  backend = "cmdstanr", file = here("results","models_fit","pilot_downstream_v2"), refresh = 0)

# (b) OFFSET, inheritance fixed at 1 -- the paired contrast written as a downstream-DOC model
f_off <- brm(bf(dn | mi(0.2) ~ 0 + season + offset(up) + (1|site), family = student()),
  data = wide,
  prior = c(prior(normal(0, 1), class = "b"), prior(student_t(3, 0, 2.5), class = "sigma"),
            prior(gamma(2, 0.1), class = "nu"), prior(normal(0, 1), class = "sd")),
  chains = 4, cores = 4, iter = 3000, warmup = 1000, seed = 7,
  backend = "cmdstanr", file = here("results","models_fit","pilot_offset"), refresh = 0)
```

```{r diag}
#| echo: false
dg <- tibble(
  Model = c("(a) free slope, me(up, 0.2)", "(b) offset, inheritance = 1"),
  `max R-hat` = c(round(max(rhat(f_free), na.rm = TRUE), 3), round(max(rhat(f_off), na.rm = TRUE), 3)),
  `min bulk ESS` = c(round(min(summary(f_free)$fixed[,"Bulk_ESS"])), round(min(summary(f_off)$fixed[,"Bulk_ESS"]))))
kable(dg)
```

```{r estimates}
#| echo: false
df <- as_draws_df(f_free); do <- as_draws_df(f_off)
res <- bind_rows(lapply(c("summer","winter"), function(s){
  ub <- mean(wide$up[wide$season == s])
  a  <- df[[paste0("b_season", s)]]; b <- df[[paste0("bsp_season", s, ":meup0.2")]]
  free <- a + (b - 1) * ub
  off  <- do[[paste0("b_season", s)]]
  bind_rows(
    tibble(season = s, model = "(a) free slope, evaluated at mean upstream DOC",
           est = mean(free), lo = quantile(free, .025), hi = quantile(free, .975)),
    tibble(season = s, model = "(b) offset (inheritance = 1)",
           est = mean(off),  lo = quantile(off,  .025), hi = quantile(off,  .975)),
    tibble(season = s, model = "(c) current manuscript Q1 model",
           est = ifelse(s == "summer", 0.137, 0.003),
           lo  = ifelse(s == "summer", 0.054, -0.064),
           hi  = ifelse(s == "summer", 0.223,  0.068)))}))
kable(res |> mutate(across(c(est, lo, hi), ~round(.x, 3))),
      col.names = c("Season","Model","$\\Delta$DOC (mg L$^{-1}$)","2.5%","97.5%"))

inh <- tibble(season = c("summer","winter"),
  inheritance = sapply(c("summer","winter"), \(s) mean(df[[paste0("bsp_season", s, ":meup0.2")]])),
  lo = sapply(c("summer","winter"), \(s) quantile(df[[paste0("bsp_season", s, ":meup0.2")]], .025)),
  hi = sapply(c("summer","winter"), \(s) quantile(df[[paste0("bsp_season", s, ":meup0.2")]], .975)))
kable(inh |> mutate(across(-season, ~round(.x, 2))),
      col.names = c("Season","Estimated inheritance slope $\\beta$","2.5%","97.5%"),
      caption = "Free-slope model: $\\beta$ should be near 1 if the model is recovering the paired contrast.")
```

```{r fig-compare}
#| fig-height: 3.6
#| fig-cap: "Estimated average ΔDOC by season under the three formulations. The free-slope implementation of the proposal disagrees with both the offset formulation and the current manuscript model."
res$model <- factor(res$model, levels = rev(unique(res$model)))
ggplot(res, aes(est, model, colour = season)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey55") +
  geom_pointrange(aes(xmin = lo, xmax = hi), position = position_dodge(width = .45), size = .5) +
  scale_colour_manual(values = c(summer = OK[["vermillion"]], winter = OK[["green"]]), name = NULL) +
  labs(x = expression(Delta*"DOC (mg "*L^-1*")"), y = NULL) +
  theme_bw(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "top")
```

::: {.callout-important}
## The decisive result

**(b) and (c) agree exactly.** The offset formulation gives +0.12 mg L⁻¹ in summer and −0.00 in winter; the current manuscript model gives +0.14 and +0.00. They are the same estimand written two ways, and both sample cleanly (R-hat ≈ 1.00).

**(a) does not.** Despite a prior centred on 1, the free slope lands at about 0.44 with an interval spanning almost the whole plausible range, R-hat is above the 1.01 standard, and the implied ΔDOC at the mean upstream concentration is about **−1.3 mg L⁻¹ in summer** — the opposite sign to every other estimate we have. With two observations per site, a site random intercept and a latent upstream variable, the slope is competing with the site effects for the same information and the model is only weakly identified.

**Conclusion: "keep downstream DOC as the target and let upstream feed in" is already implemented in the manuscript.** The paired ΔDOC model *is* the downstream-DOC model with inheritance fixed at 1. Estimating that slope freely does not add information here, it removes it.
:::

# 3b. The SCM itself, refitted with downstream DOC as the outcome

The pilots above are two-equation models. The obvious follow-up question is what happens to the **full structural causal model** if its outcome is downstream DOC rather than $\Delta$DOC. The answer can be derived before fitting anything.

## 3b.1 Why it must give the same mediator effects

The current outcome equation of the SCM is

$$\Delta\text{DOC} = a + \beta_1\,\text{DOC}_{up,z} + \beta_2\,\text{macrophytes} + \beta_3\,\text{phytoplankton} + \beta_4\,\text{litter} + \beta_5\,\text{WRT} + u_{site} + \varepsilon$$

Substituting $\Delta\text{DOC} = \text{DOC}_{down} - \text{DOC}_{up}$ and the z-scoring $\text{DOC}_{up,z} = (\text{DOC}_{up} - \mu)/\sigma$, and moving $\text{DOC}_{up}$ to the right-hand side:

$$\text{DOC}_{down} = \underbrace{(a - \beta_1\mu/\sigma)}_{\text{new intercept}} + \underbrace{(1 + \beta_1/\sigma)}_{\text{inheritance slope}}\,\text{DOC}_{up} + \beta_2\,\text{macrophytes} + \dots + u_{site} + \varepsilon$$

Every mediator term is untouched. The two outcome choices are **one model in two parameterizations**: "the effect of upstream DOC on $\Delta$DOC" and "the inheritance slope on downstream DOC" are locked together by $1 + \beta_1/\sigma$. With $\beta_1 = -0.103$ per SD and $\sigma = 2.05$ mg L$^{-1}$ from `b1`, the implied inheritance slope is already $0.950$ [0.918, 0.983] — the SCM has been estimating the free slope all along, in $\Delta$DOC units.

## 3b.2 The fit

The model `b1_downstream` is `b1` with three deliberate changes and nothing else: the outcome is `DOC_out | mi(0.2)` (one concentration's error instead of the propagated 0.283), upstream DOC enters on its raw mg L$^{-1}$ scale as `mi(DOC_input)` with a prior Normal(1, 0.3) on the inheritance slope, and its sub-model gets a prior on the mg L$^{-1}$ scale. Same mediator priors, same site random effect, same Student-*t* likelihood, same sampler settings (4 chains, 10 000 iterations, 3 000 warm-up).

```{r downstream-formula}
#| eval: false
#| code-fold: false
bf(DOC_out | mi(0.2) ~ 0 + Intercept + mi(DOC_input) + mi(macrophy_abun_z) + mi(plankton_abun_z) +
     mi(cover_litter_z) + mi(water_res_time_z) + (0 + Intercept | site), family = student()) +
bf(DOC_input | mi() ~ 0 + Intercept, family = gaussian()) +     # raw scale, prior Normal(3, 3)
# ... all other mediator equations exactly as in b1 ...
set_prior("normal(1, 0.3)", coef = "miDOC_input", resp = "DOCout")   # inheritance slope
```

```{r downstream-compare}
#| echo: false
cmp <- readRDS(here("results", "scm_downstream_comparison.rds"))
d1 <- cmp$d1; d2 <- cmp$d2
q <- function(x) sprintf("%.3f [%.3f, %.3f]", mean(x), quantile(x, .025), quantile(x, .975))
tab <- tribble(
  ~Parameter, ~`b1: outcome ΔDOC`, ~`b1_downstream: outcome downstream DOC`,
  "Inheritance slope (per mg/L of upstream DOC)", q(1 + d1$bsp_deltaDOC_miDOC_input_z / cmp$s_up), q(d2$bsp_DOCout_miDOC_input),
  "Macrophytes ($\\beta_2$)",         q(d1$bsp_deltaDOC_mimacrophy_abun_z),  q(d2$bsp_DOCout_mimacrophy_abun_z),
  "Phytoplankton ($\\beta_3$)",       q(d1$bsp_deltaDOC_miplankton_abun_z),  q(d2$bsp_DOCout_miplankton_abun_z),
  "Soil litter cover ($\\beta_4$)",   q(d1$bsp_deltaDOC_micover_litter_z),   q(d2$bsp_DOCout_micover_litter_z),
  "Water residence time ($\\beta_5$)",q(d1$bsp_deltaDOC_miwater_res_time_z), q(d2$bsp_DOCout_miwater_res_time_z),
  "Between-site SD",                  q(d1$sd_site__deltaDOC_Intercept),     q(d2$sd_site__DOCout_Intercept),
  "Residual scale $\\sigma$",         q(d1$sigma_deltaDOC),                  q(d2$sigma_DOCout),
  "Degrees of freedom $\\nu$",        q(d1$nu_deltaDOC),                     q(d2$nu_DOCout),
  "Average ΔDOC (b1 intercept; downstream model evaluated at mean upstream DOC)",
     q(d1$b_deltaDOC_Intercept), q(d2$b_DOCout_Intercept + (d2$bsp_DOCout_miDOC_input - 1) * cmp$mu_up))
kable(tab, caption = sprintf("Posterior mean [95%% CrI]. Convergence: max R-hat %.3f (b1) and %.3f (b1_downstream).", cmp$rhat["b1"], cmp$rhat["b1_downstream"]))
```

::: {.callout-important}
## Result: the full SCM is invariant to the choice of outcome

The inheritance slope is **`r sprintf("%.3f", mean(d2$bsp_DOCout_miDOC_input))`** when estimated directly against **`r sprintf("%.3f", mean(1 + d1$bsp_deltaDOC_miDOC_input_z / cmp$s_up))`** implied by `b1`; every mediator coefficient agrees to the second decimal; the site SD, the residual scale and the tail parameter are the same; and the downstream model evaluated at the mean upstream concentration returns the same null average $\Delta$DOC. The small remaining differences come only from the outcome error (0.2 versus 0.283) and the prior scale of the upstream coefficient, both changed on purpose.

So the answer to "what if the SCM's outcome were downstream DOC?" is: **the mediator story of Q2 does not move, and the upstream-DOC path is re-read as an inheritance of about 0.95 mg per mg instead of −0.10 SD on $\Delta$DOC.** That reading is worth one sentence in the Methods; it does not warrant refitting the paper.

What remains genuinely open is the version in which the 0.2 mg L$^{-1}$ error of upstream DOC is propagated as a *predictor* (latent upstream value). The two-equation pilot in section 3 showed that this is weakly identified with a site random effect and two samplings per site, and the SCM shares that structure, so it was not attempted here.
:::

::: {.callout-warning}
## A pre-existing diagnostic worth noting

Both fits report **E-BFMI of about 0.16–0.18** in every chain (the usual concern threshold is 0.3). This is not caused by the reparameterization — `b1` already had it — and R-hat and effective sample sizes are fine. The likely cause is the tail parameter: $\nu$ sits at `r sprintf("%.2f", mean(d1$nu_deltaDOC))` against its prior lower bound of 2, i.e. the data want tails at least as heavy as the bound allows, which creates a difficult posterior geometry. Lowering the bound (for example `lb = 1`) or reparameterizing $\nu$ would be the standard remedy. It does not change any estimate reported in the manuscript, but it is worth fixing before the analysis is archived.
:::

# 4. What a full rebuild would still involve

The pilot settles the target-variable question but not the rest of Josh's proposal, which is to derive B, S, G and M from the posterior of a **season-specific SCM including the mediators**. That part remains a substantial job.

## 4.1 The model

```{r spec}
#| eval: false
#| code-fold: false
# Sketch only -- NOT run. Every path needs a season interaction, and every
# predictor keeps its mi() imputation.
bf(DOC_down | mi(0.2) ~ 0 + season + offset(DOC_up) +
     season:mi(macrophy_abun_z) + season:mi(plankton_abun_z) +
     season:mi(cover_litter_z)  + season:mi(water_res_time_z) +
     (1 | site), family = student()) +
bf(water_res_time_z | mi() ~ 0 + season + season:n_dams_z +
     season:mi(dam_height_z) + season:slope_z) +
bf(plankton_abun_z  | mi() ~ 0 + season + season:mi(solar_z) + season:mi(water_res_time_z)) +
bf(macrophy_abun_z  | mi() ~ 0 + season + season:mi(solar_z) + season:mi(water_res_time_z)) +
bf(cover_litter_z   | mi() ~ 0 + season + season:mi(solar_z)) +
bf(solar_z | mi() ~ 0 + season) + bf(dam_height_z | mi() ~ 1) +
  set_rescor(FALSE)
```

## 4.2 The priors all move scale

```{r priors}
#| echo: false
kable(tribble(
  ~Parameter, ~`Current ($\\Delta$DOC target, standardized)`, ~`Required (downstream DOC target, mg L$^{-1}$)`,
  "Upstream DOC $\\to$ outcome", "$\\beta_1 \\sim$ Normal(0, 0.1), estimate $-0.10$", "inheritance fixed at 1 (offset), or Normal(1, 0.1) — a different quantity entirely",
  "Intercept", "Normal(0, 2.5) on $\\Delta$DOC", "Normal(0, 1) on the mg L$^{-1}$ offset; season-specific",
  "Producer paths → outcome", "Normal(0.2, 0.075), standardized", "must be rescaled to mg L$^{-1}$ per SD of the producer variable",
  "$\\sigma$ of the outcome", "Student-$t$(3, 0, 2.5) on $\\Delta$DOC", "same family, but the residual now absorbs concentration-level variation",
  "Site random intercept", "Normal(0, 2) on $\\Delta$DOC", "now absorbs site DOC level, so the prior must widen substantially"),
  caption = "Every prior on the outcome equation changes meaning when the target changes.")
```

## 4.3 Everything downstream of the model changes at once

```{r cascade}
#| echo: false
kable(tribble(
  ~Item, ~`Why it changes`,
  "Figure 1b (SCM path estimates)", "all coefficients move to a new scale and gain a season dimension",
  "Table S1 (13 hypotheses)", "each hypothesis is phrased in terms of ΔDOC",
  "Fig. S2 (priors)", "new prior set",
  "Figs. S3–S4 (residuals, posterior predictive)", "new outcome",
  "Fig. S5 (posterior parameters)", "new parameters, doubled by season",
  "Fig. S7 (total effects)", "path tracing must be redone per season; upstream DOC is no longer a driver but an inherited baseline, so the largest current total effect disappears by construction",
  "Figs. S8–S9 (connectivity SCM)", "must be re-specified the same way for consistency",
  "Results §2 and Methods", "every number and the causal wording"))
```

# 5. Cost, risk and recommendation

```{r cost}
#| echo: false
kable(tribble(
  ~Stage, ~`Working days`,
  "Re-specification and prior redesign, prior predictive checks", "2–3",
  "Fitting, convergence and identification troubleshooting (the pilot above already shows where it bites)", "2–3",
  "Re-deriving posterior quantities and regenerating 6–8 figures", "2–3",
  "Rewriting Methods, Results and Appendix S1", "2–3",
  "Co-author review of a substantially changed analysis", "2+",
  "**Total**", "**10–14**"))
```

::: {.callout-warning}
## Recommendation

**Do not rebuild before the 10 September submission.** Three reasons, in order of weight:

1.  **The core of the proposal is already in the manuscript.** The pilot shows the paired ΔDOC model and the downstream-DOC model with inheritance 1 are the same thing, and that the free-slope alternative is worse identified and gives a contradictory answer.
2.  **The expensive remainder changes Q2, not Q1.** A season-specific mediator SCM is a genuinely different paper-level analysis; it would also delete the single largest total effect in the current results (upstream DOC) by turning it into an offset, which is a substantive reinterpretation, not a technical refinement.
3.  **Timing.** 10–14 working days against a 7-day window, with the corresponding author unavailable from 10 September to mid-November.

**What to do instead, at no analytical cost:** add one sentence to the Methods saying that ΔDOC is modelled as a paired contrast, equivalent to modelling downstream DOC with an inheritance coefficient of one, and that the free-slope alternative was not identifiable with two samplings per reach. That answers the question a methodological reviewer would ask, and it is now backed by the pilot in this document.

**What to keep for later:** the season-specific mediator SCM is the natural core of a follow-up paper, where the seasonal expression of the biological pathway is the question rather than a robustness check.
:::

# 6. Reproducibility

```{r session}
#| echo: false
cat("Pilot fits cached in results/models_fit/: pilot_downstream_v2.rds, pilot_offset.rds\n")
cat("Data: data/processed/m6.csv\n\n")
sessionInfo()$R.version$version.string
```
