---
title: "Beaver DOC: analysis plan for the revised submission"
subtitle: "The research questions, the variables that answer them, and the figures that go into the manuscript"
author: "Leonardo Capitani"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-tools: true
    embed-resources: true
    fig-width: 9
    fig-align: center
bibliography: ../references.bib
csl: ../ecology-letters.csl
execute:
  echo: true
  message: false
  warning: false
editor: visual
---

```{r setup}
#| include: false
library(tidyverse); library(scales); library(patchwork); library(ggdist); library(knitr); library(here)
invisible(Sys.setlocale("LC_CTYPE", "en_US.UTF-8"))
set.seed(7)
# One colour per meaning, used in every figure: DOC added (source) = skyblue, DOC removed (sink) = coral,
# negligible / no-evidence = lightgrey. Never used for anything else.
PAL <- c(add = "skyblue", remove = "coral", none = "lightgrey")
dir.create(here("results", "figures", "plan"), recursive = TRUE, showWarnings = FALSE)
```

::: {.callout-tip appearance="simple"}
## Purpose

This document fixes the story of the manuscript: three research questions, the variables that answer them, and the figures that go into the main text and into Appendix S1. Everything not listed here is left out. Figure numbers in this document (Figure 1, 2, …) are its own; the manuscript figure each one becomes is written in bold above it.
:::

# 1. The variables

The subject of every effect statement is **beaver-engineering**, the physical modification beavers create, because the upstream–downstream contrast measures the whole beaver-engineered stream. The unit of analysis is the stream.

```{r}
#| label: tbl-variables
#| tbl-cap: "The derived variables used throughout this document, referenced from every figure caption below."
#| echo: false
kable(tribble(
  ~Symbol, ~Name, ~Definition, ~`Meaning (unit)`, ~`Question it answers`,
  "$\\text{DOC}_{up,s}$, $\\text{DOC}_{down,s}$", "upstream DOC, downstream DOC", "measured",
  "DOC concentration entering / leaving the beaver-engineered stream in season $s$ ($w$ = winter, $s$ = summer) **(mg L$^{-1}$)**",
  "What is the DOC concentration before and after the beaver-engineered stream in each season?",
  "$\\Delta\\text{DOC}_s$", "local response", "$\\text{DOC}_{down,s} - \\text{DOC}_{up,s}$",
  "the change in DOC that beaver-engineering produces in season $s$; positive = enrichment, negative = depletion **(mg L$^{-1}$)**",
  "Does beaver-engineering increase or decrease DOC, and by how much, in each season?",
  "$S$", "inherited seasonal contrast", "$\\text{DOC}_{up,summer} - \\text{DOC}_{up,winter}$",
  "the winter-to-summer change of upstream DOC, i.e. what the upstream river corridor delivers to the stream **(mg L$^{-1}$)**",
  "How much does DOC change between winter and summer in the river before it reaches the beaver-engineered stream?",
  "$R_s$", "relative response", "$\\Delta\\text{DOC}_s / \\lvert S \\rvert$",
  "how much DOC beaver-engineering adds ($R_s > 0$) or removes ($R_s < 0$) in season $s$, as a fraction of the inherited seasonal contrast **(dimensionless)**",
  "How much does beaver-engineering add or remove DOC, relative to the inherited seasonal contrast?",
  "$B$", "beaver seasonal modification", "$\\Delta\\text{DOC}_{summer} - \\Delta\\text{DOC}_{winter}$",
  "how much beaver-engineering changes the winter-to-summer DOC contrast **(mg L$^{-1}$)**",
  "Does beaver-engineering change the seasonal DOC contrast, beyond its effect within each season?",
  "$G$", "seasonal gain", "$B / S$",
  "$G > 0$: beaver-engineering **reinforces** the inherited seasonal contrast; $-1 < G < 0$: **dampens** it; $G < -1$: **reverses** it **(dimensionless)**",
  "Does beaver-engineering reinforce, dampen or reverse the inherited seasonal contrast?",
  "$A_\\text{beaver}$, $A_\\text{floodplain}$", "areas", "mapped",
  "ponded footprint of the beaver-engineered stream; contributing catchment floodplain area **(km$^2$)**",
  "How large is the beaver-engineered footprint, and how large is the floodplain that generates $S$?",
  "$F$", "footprint contrast", "$A_\\text{floodplain} / A_\\text{beaver}$",
  "how much larger the floodplain area is than the beaver-engineered footprint **(dimensionless)**",
  "How small is the beaver-engineered footprint relative to the floodplain that generates $S$?"))
```

**Manuscript placement: Appendix S1 (Figure S10).**

![**Derivation of the variables.** Top: the same causal structure operates in winter and summer; downstream DOC inherits upstream DOC and is modified by local hydrological and biological processes. Bottom (dashed): $\Delta\text{DOC}_s$, $S$, $R_s$ and $F$ are calculations from the concentrations, not causal nodes. Solid arrows are causal pathways; the green arrow is the persistence of the hydrologic template between seasons. Variable definitions: @tbl-variables.](../../images/DAG_variable_derivation.png){#fig-dag}

# 2. Research questions and hypotheses

**Q1 — Direction.** *Does beaver-engineering consistently increase or decrease DOC downstream compared with upstream?*
Hypothesis: if a stream is impounded by beaver dams, then the local response $\Delta\text{DOC}_s$ will not have a consistent sign across streams, because impoundment simultaneously starts DOC-producing processes — exudation and senescence of phytoplankton and macrophytes, leaching of litter and inundated soils [@søndergaard1985; @sell1992; @aitkenhead1999] — and DOC-removing processes — microbial mineralization, photodegradation and sedimentation during the longer residence time [@johnston2017; @nummi2018] — whose balance depends on the stream; syntheses accordingly report enrichment, no change and depletion [@larsen2021; @grudzinski2022]. We further expect the balance to tilt toward enrichment in summer, when production exceeds removal [@mulholland1997].

**Q2 — Drivers.** *Which abiotic and biotic drivers set the direction of the local response?*
Hypothesis: if the number of dams increases, then water residence time increases because each dam impounds water and slows flow [@westbrook2006; @majerova2015]; if water residence time and solar radiation increase, then phytoplankton and macrophyte abundance increase because nutrients are retained and light is available for photosynthesis [@reynolds2006; @finger2007; @feuchtmayr2009]; if primary producer abundance and soil litter cover increase, then $\Delta\text{DOC}_s$ increases because exudation, senescence and litter leaching release DOC [@reitsema2018; @gessner1999]; if upstream DOC increases, then $\Delta\text{DOC}_s$ decreases because a larger incoming load is proportionally retained or mineralized within the pond [@larson2007; @wohl2013]. The thirteen path-specific hypotheses are in Appendix S1, Table S1.

**Q3 — Magnitude and seasonal modification.** *How much does beaver-engineering add or remove DOC relative to the inherited seasonal contrast, and does it reinforce, dampen or reverse that contrast?*
Hypothesis, magnitude: if the season is summer, then the relative response $R_{summer}$ will be positive and a substantial fraction of the inherited seasonal contrast, because higher light and temperature raise autochthonous DOC production inside the pond independently of the upstream corridor [@mulholland1997; @battin2023]; if the season is winter, then $R_{winter}$ will be close to zero because primary production is minimal.
Hypothesis, modification: if the summer response exceeds the winter response ($B > 0$ where $S > 0$), then the seasonal gain $G$ will be positive and beaver-engineering will reinforce the inherited seasonal contrast; because the balance of DOC-producing and DOC-removing processes is stream-specific (Q1), we expect $G$ to be heterogeneous across streams, with reinforcement, dampening and reversal all present.
Because $A_\text{beaver}$ is a very small part of $A_\text{floodplain}$ (footprint contrast $F \gg 1$), both responses arise on a footprint far smaller than the corridor that generates $S$.

**Q4 — Intensity.** *Does the local response scale with the intensity of beaver-engineering — the number of dams and the water residence time — or is it organised by something else?*
Hypothesis: if the number of dams and the water residence time increase, then $R_{summer}$ and $G$ will become more positive, because longer impoundment retains nutrients and creates habitat for the primary producers that add DOC in summer [@westbrook2006; @majerova2015; @feuchtmayr2009]. Because Q2 found only weak paths from the dams through water residence time, we expect at most a weak ordering of the streams along these gradients. Q4 is descriptive and belongs in Appendix S1; it answers the question a reader asks after Figure 4: "are the streams on the right the ones with more dams?"

**Q5 — Mechanism of the sign.** *Why do some beaver-engineered streams enrich and others deplete DOC, and why does the same stream change role between seasons?*
Hypothesis: if enrichment occurs where the biological and organic-matter pathways of the SCM (phytoplankton, macrophytes, litter) contribute most, and depletion where these contributions are weak and removal dominates, then decomposing the SCM's predicted local modification into pathway contributions will separate sources from sinks, and the summer-minus-winter change of the producer contributions will track the beaver seasonal modification $B$ [@mulholland1997; @battin2023]. Removal processes (mineralization, photodegradation) were not measured, so whatever the measured pathways cannot explain is reported as an unresolved component, not attributed.

The division of labour between the two quantities is fixed: **$R_s$ answers "how much, in season $s$" (Figure 2a, b); $G$ answers "what does that do to the seasonal contrast itself" (Figure 2c)**.

| Question | Answered by | This document | Manuscript |
|---|---|---|---|
| Q1 Direction | $\Delta\text{DOC}_s$ | @fig-q1 | Figure 1a (main text) |
| Q2 Drivers | structural causal model | @fig-q2 | Figure 4 (main text) |
| Q3 Magnitude and seasonal modification | $R_s$ (figure), $G$ (text), $F$ as context | @fig-R | Figure 2 (main text) |
| Q4 Intensity | $R_{summer}$ and $G$ against the number of dams and the water residence time (Bayesian slope) | @fig-intensity | Figure S12 (Appendix S1) |
| Spatial coherence (Discussion support) | $S$, $B$, downstream seasonal contrast $T$ | @fig-spatial | Figure S11 (Appendix S1) |
| Q5 Mechanism of the sign | pathway decomposition of the SCM's local modification | @fig-decomp, @fig-decomp-season | Figure S18 (Appendix S1) |
| Model validation (Methods support) | leave-one-out and hold-out prediction of downstream DOC and $\Delta\text{DOC}$ | @fig-holdout | Figure S13 (Appendix S1) |
| Methods | sampling design; $S$, $\Delta\text{DOC}_s$, $R_s$, $F$ | @fig-map | Figure 3 (main text) |
| Derivation of the variables | — | @fig-dag | Figure S10 (Appendix S1) |

# 3. Q1 — Direction

**Manuscript placement: main text, Figure 1a (existing, unchanged).**

![**Direction of the local response.** Posterior predictive distribution of $\Delta\text{DOC}_s$ for a single sampling of a beaver-engineered stream, in winter and summer, from the measurement-error model on the raw upstream and downstream concentrations. Coral: posterior predictive probability of a sink response ($\Delta\text{DOC}_s < -0.5$ mg L$^{-1}$); grey: negligible response (ROPE $\pm 0.5$ mg L$^{-1}$); skyblue: source response ($\Delta\text{DOC}_s > 0.5$ mg L$^{-1}$). Point: median; bars: 66% and 95% intervals. $\Delta\text{DOC}_s$ is defined in @tbl-variables.](../../overleaf/figures/media/Figure_2a.png){#fig-q1}

```{r q1-data}
# raw data: one row per sampling; pair each stream's winter and summer sampling
m <- read.csv(here("data", "processed", "m12.csv"))
w <- m |> filter(season == "winter") |> select(site, beaver_territory, up_w = DOC_input, dn_w = DOC_out,
                                                 Afp = A_floodplain_km2, Abv = A_beaver_km2)
s <- m |> filter(season == "summer") |> select(site, beaver_territory, up_s = DOC_input, dn_s = DOC_out)
# the 158 streams with both seasons complete; local responses and footprint contrast
d <- inner_join(w, s, by = c("site", "beaver_territory")) |> filter(complete.cases(up_w, dn_w, up_s, dn_s)) |>
  mutate(dw = dn_w - up_w,      # local response in winter
         ds = dn_s - up_s,      # local response in summer
         F  = Afp / Abv)        # footprint contrast
# shares of enrichment by season, and streams that switch direction between seasons
sw <- d |> summarise(n = n(), winter_enrich = mean(dw > 0), summer_enrich = mean(ds > 0),
                     dep_to_enr = sum(dw < 0 & ds > 0), enr_to_dep = sum(dw > 0 & ds < 0))
```

::: {.callout-note}
## Key message of @fig-q1

-   **Average local response: negligible.** +0.14 mg L$^{-1}$ in summer, 0.00 mg L$^{-1}$ in winter, both inside the ROPE of $\pm$0.5 mg L$^{-1}$.
-   **One sampling of one stream is highly variable.** Winter: 77% negligible, 11% source, 12% sink; summer: 75% negligible, 15% source, 9% sink (posterior predictive shares).
-   **Across the `r sw$n` streams:** `r sprintf("%.0f%%", 100*sw$winter_enrich)` show enrichment in winter and `r sprintf("%.0f%%", 100*sw$summer_enrich)` in summer; `r sw$dep_to_enr` switch from winter depletion to summer enrichment, `r sw$enr_to_dep` the other way.
-   **Implication.** Beaver-engineering has no consistent direction: it is not a DOC source or a DOC sink but a control point whose sign is set locally, with a summer tilt toward release. A single field visit will most often find no ecologically relevant change — which explains the contradictory site studies in the literature.
:::

# 4. Q2 — Drivers

**Manuscript placement: main text, Figure 4 (new, standalone — promoted from panel (b) of Figure 2, given the added causal links from upstream DOC to the primary producers and upstream DOC's own measurement error; Figure 2 keeps only its direction panel).**

The chunk below draws the structural causal model as it will appear in the manuscript, using the model comparison in `notebook/correspondence/scm_outcome_choice.qmd`. The layout is fixed by hand (node positions and edge routing); the numbers are the posterior mean effects of model `downstream_DOC_me` (`notebook/fitting/me_refits.R`) — the same downstream-DOC SCM as before, refit with a latent (measurement-error-corrected) upstream DOC and two additional links, upstream DOC → macrophytes and upstream DOC → phytoplankton, testing whether the incoming DOC also shapes the local response indirectly through the producers. Each edge's line type encodes the posterior probability of its direction; the two new links are drawn in dark red.

```{r dag-draw-fn}
#| echo: false
ME <- readRDS(here("results", "me_refits.rds")); tO <- ME$down$tab
gme <- function(tb, par) tb[tb$par == par, ]
draw_dag <- function(tb, outcome, pref, up, up_lab, title) {
  nodes <- data.frame(name = c("Max dam\nheight", "Number of\ndams", "Channel\ngradient", "Water\nresidence time", "Solar\nradiation", "Macrophytes", "Phytoplankton", "Soil litter\ncover", up_lab, outcome),
    x = c(0.9, 0.9, 2.9, 3.2, 4.6, 6.6, 6.2, 6.2, 8.8, 8.8), y = c(6.6, 4.6, 2.0, 5.0, 9.3, 7.2, 5.0, 2.9, 8.8, 5.4))
  rownames(nodes) <- c("dh", "nd", "gr", "wrt", "sol", "mac", "pla", "lit", "up", "out")
  ed <- function(from, to, par, curv = 0, lx = 0, ly = 0.3, new = FALSE) { g <- gme(tb, par)
    tibble(from = from, to = to, label = sprintf("%.2f", g$mean), class = if (new) "new" else if (g$p_pos > 0.9) "pos" else if (g$p_pos < 0.1) "neg" else "none",
           solid = g$p_pos > 0.9 | g$p_pos < 0.1, curv = curv, lx = lx, ly = ly) }
  edges <- bind_rows(
    ed("dh", "wrt", "bsp_waterrestimez_midam_height_z"), ed("nd", "wrt", "b_waterrestimez_n_dams_z", lx = -0.1, ly = 0.35),
    ed("gr", "wrt", "b_waterrestimez_slope_z", lx = 0.45, ly = 0), ed("wrt", "pla", "bsp_planktonabunz_miwater_res_time_z"),
    ed("wrt", "mac", "bsp_macrophyabunz_miwater_res_time_z"), ed("wrt", "out", paste0("bsp_", pref, "_miwater_res_time_z"), curv = 0.18),
    ed("sol", "mac", "bsp_macrophyabunz_misolar_z", curv = -0.15, lx = 0.55, ly = 0.55), ed("sol", "pla", "bsp_planktonabunz_misolar_z", curv = 0.30, lx = -0.85, ly = 0.1),
    ed("sol", "lit", "bsp_coverlitterz_misolar_z", curv = 0.30), ed("mac", "out", paste0("bsp_", pref, "_mimacrophy_abun_z"), ly = 0.32),
    ed("pla", "out", paste0("bsp_", pref, "_miplankton_abun_z"), ly = 0.30), ed("lit", "out", paste0("bsp_", pref, "_micover_litter_z"), lx = -0.2, ly = 0.32),
    ed("up", "out", paste0("bsp_", pref, "_mi", up), lx = 0.45, ly = 0),
    ed("up", "mac", paste0("bsp_macrophyabunz_mi", up), curv = 0.25, lx = 0.35, ly = 0.35, new = TRUE),
    ed("up", "pla", paste0("bsp_planktonabunz_mi", up), curv = -0.35, lx = -0.9, ly = -0.9, new = TRUE))
  edges$label[!edges$solid & edges$class != "new"] <- ""
  a <- merge(edges, nodes, by.x = "from", by.y = "row.names"); a <- merge(a, nodes, by.x = "to", by.y = "row.names", suffixes = c("", "2"))
  dx <- a$x2 - a$x; dy <- a$y2 - a$y; L <- sqrt(dx^2 + dy^2)
  a$x0 <- a$x + dx / L * 0.85; a$y0 <- a$y + dy / L * 0.55; a$x1 <- a$x2 - dx / L * 0.90; a$y1 <- a$y2 - dy / L * 0.55
  a$xm <- (a$x0 + a$x1) / 2 + a$lx; a$ym <- (a$y0 + a$y1) / 2 + a$ly
  pal <- c(pos = "#3B9AD1", neg = "#E8603C", none = "grey55", new = "#8B0000"); arr <- arrow(length = unit(0.16, "cm"), type = "closed")
  pd <- ggplot()
  for (i in seq_len(nrow(a))) { lt <- if (a$class[i] == "new") (if (a$solid[i]) "solid" else "dotted") else if (a$solid[i]) "solid" else "dashed"
    pd <- pd + if (a$curv[i] == 0) geom_segment(data = a[i, ], aes(x = x0, y = y0, xend = x1, yend = y1), colour = pal[a$class[i]], linetype = lt, linewidth = 0.7, arrow = arr) else
      geom_curve(data = a[i, ], aes(x = x0, y = y0, xend = x1, yend = y1), curvature = a$curv[i], colour = pal[a$class[i]], linetype = lt, linewidth = 0.7, arrow = arr) }
  pd + geom_text(data = subset(a, label != ""), aes(x = xm, y = ym, label = label, colour = class), size = 4.2, show.legend = FALSE) +
    scale_colour_manual(values = c(pos = "black", neg = "black", none = "black", new = "#8B0000")) +
    geom_label(data = nodes[rownames(nodes) != "out", ], aes(x = x, y = y, label = name), size = 4.1, label.padding = unit(0.3, "lines"), label.r = unit(0.05, "lines")) +
    geom_label(data = nodes["out", ], aes(x = x, y = y, label = name), size = 4.4, fontface = "bold", label.padding = unit(0.45, "lines"), label.r = unit(0.9, "lines"), linewidth = 0.9, colour = "#C0208E") +
    annotate("text", x = 0.15, y = 9.8, label = title, size = 6.5, hjust = 0) +
    xlim(0, 10) + ylim(1.2, 10) + theme_void()
}
```

```{r fig-q2}
#| fig-height: 5.6
#| fig-cap: "**Drivers of downstream DOC.** Structural causal model with downstream DOC as the outcome (model `downstream_DOC_me`) and the posterior mean standardized effects; the upstream-DOC path is the inheritance slope (0.94, mg per mg). Skyblue: positive and coral: negative effects with posterior probability above 0.90; dashed grey: no-evidence. Dark red: the two added links, upstream DOC → macrophytes and → phytoplankton (per mg L⁻¹ of upstream DOC); solid dark red meets the same 0.90 threshold, dotted does not."
pq2 <- draw_dag(tO, "Downstream\nDOC", "DOCout", "DOC_input", "Upstream\nDOC", "")
ggsave(here("images", "Figure_1b_downstream.png"), pq2, width = 9, height = 5.6, dpi = 300, bg = "white")
pq2
```

::: {.callout-note}
## Key message of @fig-q2

-   **Inheritance dominates.** Downstream DOC carries `r sprintf("%.2f", gme(tO,"bsp_DOCout_miDOC_input")$mean)` mg per mg of upstream DOC (P > 0 ≈ 1) — by far the largest path.
-   **Local modification through biology.** Phytoplankton +`r sprintf("%.2f", gme(tO,"bsp_DOCout_miplankton_abun_z")$mean)`, macrophytes +`r sprintf("%.2f", gme(tO,"bsp_DOCout_mimacrophy_abun_z")$mean)`, soil litter +`r sprintf("%.2f", gme(tO,"bsp_DOCout_micover_litter_z")$mean)` (standardized effects on downstream DOC); solar radiation feeds the producers (+`r sprintf("%.2f", gme(tO,"bsp_planktonabunz_misolar_z")$mean)` phytoplankton, +`r sprintf("%.2f", gme(tO,"bsp_macrophyabunz_misolar_z")$mean)` macrophytes).
-   **The dams act only indirectly and weakly.** Number of dams +`r sprintf("%.2f", gme(tO,"b_waterrestimez_n_dams_z")$mean)` and channel gradient `r sprintf("%.2f", gme(tO,"b_waterrestimez_slope_z")$mean)` on water residence time; residence time +`r sprintf("%.2f", gme(tO,"bsp_planktonabunz_miwater_res_time_z")$mean)` on phytoplankton; no direct residence-time → DOC path.
-   **The incoming DOC also reaches the producers, macrophytes only.** Upstream DOC → macrophytes `r sprintf("%.2f", gme(tO,"bsp_macrophyabunz_miDOC_input")$mean)` per mg L⁻¹ (P > 0 = `r sprintf("%.2f", gme(tO,"bsp_macrophyabunz_miDOC_input")$p_pos)`); upstream DOC → phytoplankton `r sprintf("%.2f", gme(tO,"bsp_planktonabunz_miDOC_input")$mean)` (P > 0 = `r sprintf("%.2f", gme(tO,"bsp_planktonabunz_miDOC_input")$p_pos)`, no-evidence) — a DOC-rich inflow goes with more macrophytes, not more phytoplankton.
-   **Implication.** A beaver-engineered stream inherits its DOC and then adds to it; the addition is decided by light and primary producers, not by the size of the engineering, and a richer inflow itself feeds the macrophytes a little further. The producer effects are prior-informed (16 summer streams; section 9) and should be presented as such.
:::

**Parameterization.** The structural causal model is reported with **downstream DOC as the outcome**: upstream DOC enters as a latent true concentration (its reading carries a 0.2 mg L$^{-1}$ instrument error) with a freely estimated inheritance slope, and $\Delta\text{DOC}_s$ is computed from the posterior afterwards as $\text{DOC}_{down,s} - \text{DOC}_{up,s}$, so upstream DOC is never both a predictor and part of the outcome. This reading replaces the older wording of Q2, which described the upstream-DOC path as upstream DOC *causing greater retention*; downstream DOC instead **inherits** the upstream state (slope `r sprintf("%.2f", gme(tO,"bsp_DOCout_miDOC_input")$mean)` mg per mg), and the hydrological/biological pathways are local modifications around that inherited background, not retention of it. Two causal links complete the picture: upstream DOC also feeds the primary producers directly, tested by adding upstream DOC → macrophytes and upstream DOC → phytoplankton to the DAG (supported for macrophytes, not for phytoplankton, above). This model (`downstream_DOC_me`, fit in `notebook/fitting/me_refits.R`, compared against the ΔDOC parameterization in `notebook/correspondence/scm_outcome_choice.qmd`) is the $\Delta$DOC model in a different parameterization: every mediator effect agrees to the second decimal, and the upstream-DOC path is re-read as an inheritance slope of **`r sprintf("%.2f", gme(tO,"bsp_DOCout_miDOC_input")$mean)` [`r sprintf("%.2f", gme(tO,"bsp_DOCout_miDOC_input")$lo)`, `r sprintf("%.2f", gme(tO,"bsp_DOCout_miDOC_input")$hi)`]** — downstream DOC inherits `r sprintf("%.0f", 100*gme(tO,"bsp_DOCout_miDOC_input")$mean)`% of the upstream concentration, mg per mg. Figure 4 (@fig-q2) replaces the former panel (b) of Figure 2; the Methods gain one sentence stating the parameterization, the inheritance slope, and the two added producer links.

**Robustness check: does fixing the inheritance slope at 1 change anything else?** `downstream_DOC_me` estimates the slope freely (prior Normal(1, 0.3), posterior `r sprintf("%.2f", gme(tO,"bsp_DOCout_miDOC_input")$mean)`). A second fit, `downstream_DOC_me_fixed` (`notebook/fitting/downstream_DOC_me_fixed_fit.R`), pins it at 1 (prior Normal(1, 0.001), the standard way to fix an `mi()` coefficient in brms) and is otherwise identical, including the two upstream → producer links and the upstream measurement error.

```{r robustness-inheritance}
#| echo: false
fx_path <- here("results", "models_fit", "downstream_DOC_me_fixed.rds")
if (file.exists(fx_path)) {
  library(brms)
  fit_free <- readRDS(here("results", "models_fit", "downstream_DOC_me.rds")); fit_fix <- readRDS(fx_path)
  d_free <- as_draws_df(fit_free); d_fix  <- as_draws_df(fit_fix)
  qf2 <- function(x) sprintf("%.3f [%.3f, %.3f]", mean(x), quantile(x, .025), quantile(x, .975))
  rob <- tribble(
    ~Parameter, ~`Free slope (downstream_DOC_me)`, ~`Fixed at 1 (downstream_DOC_me_fixed)`,
    "Inheritance slope",  qf2(d_free$bsp_DOCout_miDOC_input),      qf2(d_fix$bsp_DOCout_miDOC_input),
    "Macrophytes",        qf2(d_free$bsp_DOCout_mimacrophy_abun_z), qf2(d_fix$bsp_DOCout_mimacrophy_abun_z),
    "Phytoplankton",      qf2(d_free$bsp_DOCout_miplankton_abun_z), qf2(d_fix$bsp_DOCout_miplankton_abun_z),
    "Soil litter cover",  qf2(d_free$bsp_DOCout_micover_litter_z),  qf2(d_fix$bsp_DOCout_micover_litter_z),
    "Water residence time", qf2(d_free$bsp_DOCout_miwater_res_time_z), qf2(d_fix$bsp_DOCout_miwater_res_time_z),
    "Upstream DOC → macrophytes", qf2(d_free$bsp_macrophyabunz_miDOC_input), qf2(d_fix$bsp_macrophyabunz_miDOC_input),
    "Upstream DOC → phytoplankton", qf2(d_free$bsp_planktonabunz_miDOC_input), qf2(d_fix$bsp_planktonabunz_miDOC_input),
    "Intercept",          qf2(d_free$b_DOCout_Intercept),           qf2(d_fix$b_DOCout_Intercept))
  kable(rob, caption = "Posterior mean [95% CrI]. Fixing the inheritance slope at 1 is a robustness check, not a preferred model.")
  # largest shift among the mediator/producer-link paths, free vs fixed
  shift_pars <- c("bsp_DOCout_mimacrophy_abun_z", "bsp_DOCout_miplankton_abun_z", "bsp_DOCout_micover_litter_z",
                   "bsp_DOCout_miwater_res_time_z", "bsp_macrophyabunz_miDOC_input", "bsp_planktonabunz_miDOC_input")
  shift_lab <- c(bsp_DOCout_mimacrophy_abun_z = "macrophytes", bsp_DOCout_miplankton_abun_z = "phytoplankton",
                 bsp_DOCout_micover_litter_z = "soil litter cover", bsp_DOCout_miwater_res_time_z = "water residence time",
                 bsp_macrophyabunz_miDOC_input = "upstream DOC → macrophytes", bsp_planktonabunz_miDOC_input = "upstream DOC → phytoplankton")
  shifts <- sapply(shift_pars, function(p) abs(mean(d_free[[p]]) - mean(d_fix[[p]])))
  worst <- names(which.max(shifts))
  rob_msg <- sprintf("Both fits converge cleanly (max R-hat %.2f and %.2f). Every mediator effect and both upstream → producer links agree within their credible intervals whether the inheritance slope is estimated freely (%.2f) or pinned at 1 — the largest shift is %s, %.3f to %.3f. Only the intercept moves (%.2f to %.2f), which is expected: fixing the slope at exactly 1 forces the intercept to absorb the small remaining offset.",
    max(rhat(fit_free), na.rm = TRUE), max(rhat(fit_fix), na.rm = TRUE), mean(d_free$bsp_DOCout_miDOC_input),
    shift_lab[worst], mean(d_free[[worst]]), mean(d_fix[[worst]]), mean(d_free$b_DOCout_Intercept), mean(d_fix$b_DOCout_Intercept))
} else { rob_msg <- "pending"
  cat("Fit in progress (notebook/fitting/downstream_DOC_me_fixed_fit.R) -- table will appear once results/models_fit/downstream_DOC_me_fixed.rds exists.")
}
```

**Result: fixing the slope changes nothing else.** `r rob_msg` The mediator story of Figure 4 does not depend on which value the inheritance slope takes.

# 5. Q3 — Magnitude and seasonal modification: $R_s$ and $G$

The analysis uses the **158 streams with both a winter and a summer sampling**.

$R_s$, $B$ and $G$ are computed from posterior draws of each stream's concentrations (Bayesian measurement-error model, `notebook/concentrations_fit.qmd`; the model belongs in the Methods), so no derived quantity is ever formed from raw values and the uncertainty is carried through (`r nrow(readRDS(here("results","concentrations_thin.rds"))$S)` posterior draws per stream).

```{r q3-compute}
# LR holds one matrix per quantity: rows = 4000 posterior draws, columns = 158 streams
LR <- readRDS(here("results", "concentrations_thin.rds"))
R_s <- LR$R_s   # relative response in summer, per draw and stream
R_w <- LR$R_w   # relative response in winter
S_m <- LR$S     # inherited seasonal contrast
# medR: for each posterior draw, the median across streams -> a posterior
# distribution of the across-stream median (optionally on a subset of streams)
medR <- function(R, keep = rep(TRUE, ncol(R))) apply(R[, keep, drop = FALSE], 1, median)
# per-stream median |S|, used only for the small-denominator stability check
absS <- apply(abs(S_m), 2, median)
# qf: format a posterior as "median [95% credible interval]"
qf   <- function(x) sprintf("%.2f [%.2f, %.2f]", median(x), quantile(x, .025), quantile(x, .975))
q3 <- tibble(
  Quantity = c("Median $R_{summer}$ across streams", "Median $R_{winter}$ across streams", "$R_{summer} - R_{winter}$",
               "Median $R_{summer}$, excluding the 20% smallest $\\lvert S \\rvert$ (ratio-stability check)"),
  `Posterior median [95% CrI]` = c(qf(medR(R_s)), qf(medR(R_w)), qf(medR(R_s) - medR(R_w)), qf(medR(R_s, absS > quantile(absS, .2)))),
  `P(> 0)` = sprintf("%.3f", c(mean(medR(R_s) > 0), mean(medR(R_w) > 0), mean(medR(R_s) - medR(R_w) > 0),
                               mean(medR(R_s, absS > quantile(absS, .2)) > 0))))
kable(q3)
# streams whose summer relative response is positive (or negative) with P > 0.90
n_add <- sum(colMeans(R_s > 0) > 0.90); n_rem <- sum(colMeans(R_s < 0) > 0.90); n_str <- ncol(R_s)
# Seasonal gain G = B / S, per stream, from the same posterior draws
# beaver seasonal modification B = summer response - winter response, per draw
B_m <- LR$ds - LR$dw
# seasonal gain G = B / S: does beaver-engineering reinforce (G > 0),
# dampen (-1 < G < 0) or reverse (G < -1) the inherited seasonal contrast?
G_m <- B_m / S_m
g_med <- median(apply(G_m, 1, function(g) median(g)))  # posterior median of the across-stream median gain
```

**Manuscript placement: main text, Figure 2 (new; replaces the amplification-factor figure).**

```{r fig-R}
#| fig-height: 12.5
#| fig-cap: "**Relative response $R_s$ and seasonal gain $G$ of beaver-engineering.** (a) Posterior distribution of the median relative response across the 158 streams, in winter and summer; point: posterior median, bars: 66% and 95% credible intervals; dashed line: no response. (b) Relative response of each stream in summer: posterior median with 66% (thick) and 95% (thin) credible intervals, streams ranked by their median; the grey band marks $\\lvert R_{summer} \\rvert < 0.1$. The vertical axis is linear between $-0.1$ and $0.1$ and logarithmic beyond, so small and large responses are both readable. In panels (a) and (b) skyblue: source (beaver-engineering adds DOC, P > 0.90), coral: sink (removes DOC, P > 0.90), grey: no-evidence — the same colours as in the direction figure; panel (b) prints the percentage of the 158 streams in each colour directly on the plot. (c) Out of the 158 streams, what percentage reinforce ($G > 0$, make the inherited seasonal contrast bigger), dampen ($-1 < G < 0$, make it smaller) or reverse ($G < -1$, flip its sign)? Each stream is classified on every posterior draw, so each bar is a full posterior distribution of that percentage, with its own uncertainty — not a single count of streams sorted into fixed bins; point: posterior median percentage, bars: 66% and 95% credible intervals. $R_s$, $G$, $B$ and $S$ are defined in @tbl-variables."
# classify by posterior probability: add if P(R > 0) > 0.90, remove if P(R < 0) > 0.90, else no-evidence
cls <- function(p_add) case_when(p_add > 0.90 ~ "add", p_add < 0.10 ~ "remove", TRUE ~ "none")
lab <- c(add = "source (adds DOC)", remove = "sink (removes DOC)", none = "no-evidence")   # P > 0.90 stated once, in the caption

# ---- panel (a): across-stream median, winter and summer --------------------
pop <- bind_rows(tibble(season = "summer", v = medR(R_s)), tibble(season = "winter", v = medR(R_w))) |>
  mutate(season = factor(season, levels = c("winter", "summer"))) |>
  group_by(season) |> mutate(class = cls(mean(v > 0))) |> ungroup()
pa <- ggplot(pop, aes(x = v, y = season, fill = class)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
  stat_halfeye(.width = c(.66, .95), point_interval = "median_qi", alpha = .9) +  # 66% and 95% intervals
  scale_fill_manual(values = PAL, labels = lab, breaks = names(lab), name = NULL, drop = FALSE) +
  labs(x = expression("Median relative response across streams,  "*R[s]*" = "*Delta*DOC[s]*" / |S|"), y = NULL) +
  theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank(), legend.position = "top")

# ---- panel (b): each stream's summer relative response, ranked -------------
# per stream: posterior median, 66% and 95% credible intervals, and its class
per <- tibble(med  = apply(R_s, 2, median),
              lo95 = apply(R_s, 2, quantile, .025), hi95 = apply(R_s, 2, quantile, .975),
              lo66 = apply(R_s, 2, quantile, .17),  hi66 = apply(R_s, 2, quantile, .83),
              class = cls(colMeans(R_s > 0))) |> arrange(med) |> mutate(rank = row_number())
# percentage of the 158 streams in each colour: the legend names the colour,
# the in-panel text gives its share (top-left corner is empty of points)
pct_b <- prop.table(table(factor(per$class, levels = c("add", "remove", "none")))) * 100
pct_lab <- tibble(class = c("add", "remove", "none"),
                   txt = sprintf("%.0f%% of streams", pct_b[c("add", "remove", "none")]),
                   y = c(90, 30, 10))   # stacked top to bottom on the log axis
# axis transform: linear inside [-0.1, 0.1], logarithmic beyond
psl <- pseudo_log_trans(sigma = 0.1, base = 10)
pb <- ggplot(per, aes(x = rank, y = med, colour = class)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = -0.1, ymax = 0.1, fill = "grey88", alpha = .8) +
  geom_hline(yintercept = 0, colour = "grey45") +
  geom_linerange(aes(ymin = lo95, ymax = hi95), alpha = .45, linewidth = 0.4) +  # 95% interval (thin)
  geom_linerange(aes(ymin = lo66, ymax = hi66), alpha = .8, linewidth = 0.9) +   # 66% interval (thick)
  geom_point(size = 1.6) +
  geom_text(data = pct_lab, aes(x = 3, y = y, label = txt, colour = class), hjust = 0, size = 4, fontface = "bold", show.legend = FALSE) +
  scale_colour_manual(values = c(add = "#3B9AD1", remove = "#E8603C", none = "grey60"), labels = lab, breaks = names(lab), name = NULL) +
  scale_y_continuous(trans = psl, breaks = c(-10, -1, -0.1, 0, 0.1, 1, 10)) +
  labs(x = "Stream", y = expression(R[summer]*" = "*Delta*DOC[summer]*" / |S|")) +
  theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank(), legend.position = "top")
# ---- panel (c): posterior share of streams that reinforce/dampen/reverse ---
# classify every stream on EVERY posterior draw (no P > 0.90 threshold), then
# compute the proportion in each zone per draw -- this gives a posterior
# distribution of the proportion itself, propagating its own uncertainty,
# rather than a single count built from per-stream point classifications.
zone <- function(g) case_when(g > 0 ~ "reinforce", g > -1 ~ "dampen", TRUE ~ "reverse")
zmat <- apply(G_m, c(1, 2), zone)                       # same shape as G_m, one label per draw x stream
gprop <- as_tibble(t(apply(zmat, 1, function(row)        # proportion per zone, one row per draw
  c(reinforce = mean(row == "reinforce"), dampen = mean(row == "dampen"), reverse = mean(row == "reverse"))))) |>
  pivot_longer(everything(), names_to = "zone", values_to = "p") |>
  mutate(zone = factor(zone, levels = c("reverse", "dampen", "reinforce")))
gpal <- c(reinforce = "#8856A7", dampen = "#41AB5D", reverse = "#B5651D")   # light brown
pc <- ggplot(gprop, aes(x = p, y = zone, fill = zone)) +
  stat_halfeye(.width = c(.66, .95), point_interval = "median_qi") +   # 66% and 95% credible intervals
  scale_fill_manual(values = gpal, guide = "none") +
  scale_x_continuous(labels = label_percent(accuracy = 1)) +
  labs(x = "Percentage of the 158 streams", y = expression("Seasonal gain,  "*G*" = "*B*" / "*S)) +
  theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank())
fig2 <- pa / pb / pc + plot_layout(heights = c(1, 1.3, 1)) + plot_annotation(tag_levels = "a")
ggsave(here("results", "figures", "plan", "fig_R.png"), fig2, width = 9, height = 11, dpi = 300)
fig2
# summary of panel (c) for the key message, one row per zone
gsum <- gprop |> group_by(zone) |> summarise(med = median(p), lo = quantile(p, .025), hi = quantile(p, .975))
gz <- function(z) gsum |> filter(zone == z)

# same +/-0.1 ROPE as panel (b)'s grey band, applied to R_summer, to compare
# against @fig-q1's ROPE logic (see the note below)
zone_r <- function(r) case_when(r > 0.1 ~ "source", r < -0.1 ~ "sink", TRUE ~ "negligible")
rzmat <- apply(R_s, c(1, 2), zone_r)
rprop <- as_tibble(t(apply(rzmat, 1, function(row)
  c(source = mean(row == "source"), negligible = mean(row == "negligible"), sink = mean(row == "sink"))))) |>
  pivot_longer(everything(), names_to = "zone", values_to = "p")
rz <- function(z) rprop |> filter(zone == z) |> summarise(med = median(p))
```

::: {.callout-important}
## Key message of @fig-R

-   **(a) Magnitude.** Winter: relative response `r sprintf("%.2f", median(medR(R_w)))` — nothing added. Summer: beaver-engineering adds DOC equal to `r sprintf("%.0f%%", 100*median(medR(R_s)))` of the inherited seasonal contrast $S$ (95% CrI `r sprintf("%.0f–%.0f%%", 100*quantile(medR(R_s), .025), 100*quantile(medR(R_s), .975))`).
-   **(b) Heterogeneity.** In summer `r sprintf("%.0f%%", 100 * n_add / n_str)` of streams (`r n_add` of `r n_str`) are DOC sources, `r sprintf("%.0f%%", 100 * n_rem / n_str)` (`r n_rem`) are sinks, the rest no-evidence.
-   **(c) Seasonal gain.** `r sprintf("%.0f%%", 100*gz("reinforce")$med)` (`r sprintf("%.0f–%.0f%%", 100*gz("reinforce")$lo, 100*gz("reinforce")$hi)`) of streams reinforce the inherited seasonal contrast, `r sprintf("%.0f%%", 100*gz("dampen")$med)` dampen it, `r sprintf("%.0f%%", 100*gz("reverse")$med)` reverse it; median gain `r sprintf("%.2f", g_med)`.
-   **Implication.** Beaver-engineering adds about one fifth of the seasonal DOC signal in summer and nothing in winter, and it does not shift the seasonal contrast in one direction — it redistributes it stream by stream. Ecologically, beaver ponds are seasonal control points on a regionally inherited signal, not amplifiers of it.
:::

::: {.callout-note}
## How @fig-q1 and @fig-R fit together

@fig-q1 and @fig-R are not two conflicting results — they answer two different questions, so one naturally reads as "mostly nothing detectable" and the other as "most streams have a clear sign."

- **@fig-q1 asks: if someone visits a stream and takes one water sample, what would they most likely find?** This is a single field visit — the way almost every published beaver–DOC study, including the ones this manuscript is compared against, actually samples. The answer: mostly nothing an ecologist would call a real change.
- **@fig-R asks: what is each stream's true relative response, according to the best estimate the model can produce from everything it knows?** This is the paper's actual result — it does not simulate a single field visit, it reports the model's estimate of the real value. The answer: most streams have a clear sign.

Both questions are legitimate and both answers are correct; they are simply not the same question. (The technical reason the two differ — one reports a simulated single observation, the other reports the model's own estimate — is a Methods-level detail and does not need to be repeated at every figure.) For consistency, applying @fig-q1's "no real change" logic to $R_{summer}$ (a $\pm$0.1 band, matching the grey band already drawn in panel (b)) still leaves only `r sprintf("%.0f%%", 100*rz("negligible")$med)` of streams with no detectable response, versus `r sprintf("%.0f%%", 100*rz("source")$med)` source and `r sprintf("%.0f%%", 100*rz("sink")$med)` sink.

Methods sentence: state once that Figure 1a reports the posterior *predictive* distribution of a single new sampling (so it includes the instrument error), while $R_s$, $B$ and $G$ are computed from each stream's posterior *parameter* estimate (the model's estimate of the true concentration).
:::

# 6. Q4 — Intensity of beaver-engineering and the response

::: {.callout-note appearance="simple"}
**Question:** Is the local response ordered by the intensity of beaver-engineering — the number of dams and the water residence time — or by something else?
:::

Figure 2b ranks the streams by their own response. Here the same per-stream posteriors of $R_{summer}$ and $G$ are placed on the two gradients of beaver-engineering intensity we measured, and the ordering is estimated with a Bayesian model rather than a correlation coefficient: for each response and each gradient, a Student-*t* regression of the response (on the same pseudo-log scale as the figure axis, $t(x) = \operatorname{asinh}(x/0.2)/\ln 10$) on the standardized gradient, in which each stream enters with its posterior median and its posterior standard deviation as a known measurement error (`se()` in brms). The slope $\beta$ is the ordering: $P(\beta > 0)$ is the posterior probability that the response increases along the gradient.

**Manuscript placement: Appendix S1 (Figure S12), with one sentence in the Results.**

```{r fig-intensity}
#| fig-height: 8
#| fig-cap: "**Relative response and seasonal gain against the intensity of beaver-engineering.** Each point is a stream: posterior median of $R_{summer}$ (top row) or $G$ (bottom row) with 66% (thick) and 95% (thin) credible intervals, against the number of dams (left; jittered) and the summer water residence time (right; log axis). Colours as in Figure 2: source / sink / no-evidence for $R_{summer}$; reinforce / dampen / reverse / no-evidence for $G$ (classified at posterior probability 0.90). The vertical axes are linear between −0.1 and 0.1 and logarithmic beyond. The black line and band are the posterior median and 95% credible interval of a Bayesian Student-t regression of the response (pseudo-log scale) on the standardized gradient, with each stream's posterior standard deviation as its measurement error, drawn back on the response scale; each panel title gives its slope $\\beta$ (posterior median [95% credible interval]) and $P(\\beta > 0)$. $R_s$, $G$, $B$ and $S$ are defined in @tbl-variables."
library(brms)
# covariates per analysed stream: number of dams and summer water residence time (from m6)
cov <- read.csv(here("data", "processed", "m6.csv")) |> filter(season == "summer") |>
  select(site, beaver_territory, n_dams, wrt = water_res_time)
cv <- LR$d |> select(site, beaver_territory) |> left_join(cov, by = c("site", "beaver_territory"))
# per-stream classes for colouring (same rules as Figure 2)
r_class <- cls(colMeans(R_s > 0))
g_class <- apply(G_m, 2, function(g) {
  p <- c(reinforce = mean(g > 0), dampen = mean(g > -1 & g < 0), reverse = mean(g < -1))
  if (max(p) > 0.90) names(p)[which.max(p)] else "no-evidence" })
# the pseudo-log transform used on the figure axis (scales::pseudo_log_trans, sigma = 0.1)
plog <- function(x) asinh(x / 0.2) / log(10)
# Bayesian ordering: Student-t regression of the transformed response on the standardized
# gradient, each stream weighted by its own posterior uncertainty (known se)
slope_fit <- function(M, x, tag) {
  ok <- !is.na(x); TM <- plog(M[, ok])
  dd <- tibble(y = apply(TM, 2, median), se_y = apply(TM, 2, sd), x = as.numeric(scale(x[ok])))
  brm(bf(y | se(se_y, sigma = TRUE) ~ 1 + x), data = dd, family = student(),
      prior = c(prior(normal(0, 1), class = "b"), prior(normal(0, 2), class = "Intercept"),
                prior(student_t(3, 0, 1), class = "sigma"), prior(gamma(2, 0.1), class = "nu")),
      chains = 4, iter = 3000, warmup = 1000, seed = 7, backend = "cmdstanr", refresh = 0,
      file = here("results", "models_fit", paste0("intensity_", tag)))
}
fits <- list(R_dams = slope_fit(R_s, cv$n_dams, "R_dams"), R_wrt = slope_fit(R_s, log10(cv$wrt), "R_wrt"),
             G_dams = slope_fit(G_m, cv$n_dams, "G_dams"), G_wrt = slope_fit(G_m, log10(cv$wrt), "G_wrt"))
beta <- lapply(fits, function(f) as_draws_df(f)$b_x)
b_lab <- function(b) sprintf("beta = %.2f [%.2f, %.2f], P(beta > 0) = %.2f", median(b), quantile(b, .025), quantile(b, .975), mean(b > 0))
qs <- function(M) tibble(med = apply(M, 2, median), lo95 = apply(M, 2, quantile, .025), hi95 = apply(M, 2, quantile, .975),
                         lo66 = apply(M, 2, quantile, .17), hi66 = apply(M, 2, quantile, .83))
rpal <- c(add = "#3B9AD1", remove = "#E8603C", none = "grey60")
rlab <- c(add = "source", remove = "sink", none = "no-evidence")
gpal2 <- c(reinforce = "#8856A7", dampen = "#41AB5D", reverse = "#B5651D", `no-evidence` = "grey60")
inv_plog <- function(y) 0.2 * sinh(y * log(10))   # back from the pseudo-log scale
# posterior regression line of the slope model, drawn on the original scale of the response
fit_line <- function(fit, x_raw, logx) {
  ok <- !is.na(x_raw); xr <- if (logx) log10(x_raw[ok]) else x_raw[ok]
  grid <- seq(min(xr), max(xr), length.out = 60)
  nd <- tibble(x = (grid - mean(xr)) / sd(xr), se_y = 0)
  ep <- posterior_epred(fit, newdata = nd)                  # draws x grid, pseudo-log scale
  tibble(x = if (logx) 10^grid else grid, med = inv_plog(apply(ep, 2, median)),
         lo = inv_plog(apply(ep, 2, quantile, .025)), hi = inv_plog(apply(ep, 2, quantile, .975)))
}
panel <- function(M, x, xlab, class, pal, labs_, ylab, sub, fit, logx = FALSE) {
  d <- qs(M) |> mutate(x = x, class = factor(class, levels = names(pal)))
  pos <- if (logx) "identity" else position_jitter(width = .15, height = 0, seed = 1)
  ln <- fit_line(fit, x, logx)
  p <- ggplot(d, aes(x = x, y = med)) +
    geom_hline(yintercept = 0, colour = "grey45") +
    geom_linerange(aes(ymin = lo95, ymax = hi95, colour = class), alpha = .45, linewidth = 0.4, position = pos) +
    geom_linerange(aes(ymin = lo66, ymax = hi66, colour = class), alpha = .8, linewidth = 0.9, position = pos) +
    geom_point(aes(colour = class), size = 1.6, position = pos) +
    geom_ribbon(data = ln, aes(x = x, ymin = lo, ymax = hi), inherit.aes = FALSE, fill = "black", alpha = .15) +
    geom_line(data = ln, aes(x = x, y = med), inherit.aes = FALSE, colour = "black", linewidth = 0.9) +
    scale_colour_manual(values = pal, labels = labs_, breaks = names(pal), name = NULL) +
    scale_y_continuous(trans = psl, breaks = c(-10, -1, -0.1, 0, 0.1, 1, 10)) +
    labs(x = xlab, y = ylab, subtitle = sub) + theme_bw(base_size = 12) +
    theme(panel.grid.minor = element_blank(), legend.position = "top", plot.subtitle = element_text(size = 10))
  if (logx) p + scale_x_log10() else p
}
ylab_R <- expression(R[summer] == frac(DOC[list(down, summer)] - DOC[list(up, summer)], "|" * DOC[list(up, summer)] - DOC[list(up, winter)] * "|"))
ylab_G <- expression(G == frac(Delta * DOC[summer] - Delta * DOC[winter], DOC[list(up, summer)] - DOC[list(up, winter)]))
p1 <- panel(R_s, cv$n_dams, "Number of dams", r_class, rpal, rlab, ylab_R, b_lab(beta$R_dams), fits$R_dams)
p2 <- panel(R_s, cv$wrt, "Water residence time in summer (s)", r_class, rpal, rlab, ylab_R, b_lab(beta$R_wrt), fits$R_wrt, logx = TRUE)
p3 <- panel(G_m, cv$n_dams, "Number of dams", g_class, gpal2, names(gpal2), ylab_G, b_lab(beta$G_dams), fits$G_dams)
p4 <- panel(G_m, cv$wrt, "Water residence time in summer (s)", g_class, gpal2, names(gpal2), ylab_G, b_lab(beta$G_wrt), fits$G_wrt, logx = TRUE)
figI <- (p1 + p2) / (p3 + p4) + plot_layout(guides = "collect") + plot_annotation(tag_levels = "a") &
  theme(legend.position = "top")
ggsave(here("results", "figures", "plan", "fig_intensity.png"), figI, width = 10, height = 8, dpi = 300)
figI
bt <- function(k) sprintf("%.2f [%.2f, %.2f]", median(beta[[k]]), quantile(beta[[k]], .025), quantile(beta[[k]], .975))
bp <- function(k) sprintf("%.2f", mean(beta[[k]] > 0))
```

::: {.callout-note}
## Key message of @fig-intensity

-   **Water residence time orders the summer response, weakly.** $\beta$ = `r bt("R_wrt")` per SD of log residence time, $P(\beta > 0)$ = `r bp("R_wrt")`.
-   **The number of dams does not.** $\beta$ = `r bt("R_dams")`, $P(\beta > 0)$ = `r bp("R_dams")`.
-   **The seasonal gain is ordered by neither.** `r bt("G_dams")` (dams) and `r bt("G_wrt")` (residence time).
-   **Implication.** Longer impoundment favours DOC release in summer (the stream-level counterpart of the Q2 path residence time → phytoplankton → DOC), but sources, sinks, reinforcing and dampening streams occur at every level of engineering: the intensity of beaver-engineering sets the stage, it does not decide the sign. Sentence for Appendix S1 (Figure S12) as above.
:::

# 7. Spatial coherence of the seasonal contrasts

::: {.callout-note appearance="simple"}
**Question** (Discussion support)**:** *the inherited seasonal contrast is generated over whole floodplains — is the modification that beaver-engineering imposes on it organised at the same regional scale, or is it stream-specific?* The answer decides how the Discussion frames beaver-engineering: if $B$ were regionally coherent, beaver-engineered streams would act together as a landscape-scale control on the seasonal DOC signal; if $B$ is spatially incoherent, each beaver-engineered stream is an independent local control point on a regional signal — the framing the manuscript title claims.
:::

For each stream we correlate its value with the mean of its 8 nearest neighbours (Spearman), computed on every posterior draw of the concentrations, so the coherence estimate carries the full uncertainty. $T = \text{DOC}_{down,summer} - \text{DOC}_{down,winter}$ is the **downstream seasonal contrast** ($T = S + B$).

**Manuscript placement: Appendix S1 (Figure S11), plus one sentence in the Discussion.**

```{r fig-spatial}
#| fig-height: 3.5
#| fig-cap: "**Spatial coherence of the inherited seasonal contrast $S$, the beaver seasonal modification $B$ and the downstream seasonal contrast $T$.** Posterior distribution of the Spearman correlation between each stream's value and the mean of its 8 nearest neighbours, computed on every posterior draw of the concentrations. Point: median; bars: 66% and 95% credible intervals; dashed line: no spatial coherence. $S$ and $B$ are defined in @tbl-variables ($T = S + B$)."
# stream coordinates (longitude, latitude of the upstream sampling point)
co <- read.csv(here("data", "processed", "m6.csv")) |> distinct(site, .keep_all = TRUE) |>
  select(site, lon = x_upstream, lat = y_upstream)
d2 <- left_join(LR$d, co, by = "site")
# planar distances in km (adequate at the scale of Switzerland)
kmx <- (d2$lon - mean(d2$lon)) * 111.32 * cos(mean(d2$lat) * pi / 180)
kmy <- (d2$lat - mean(d2$lat)) * 110.57
D <- as.matrix(dist(cbind(kmx, kmy))); diag(D) <- Inf
# for each stream, the indices of its 8 nearest neighbours (fixed across draws)
nb <- t(apply(D, 1, function(r) order(r)[1:8]))
# mean of the neighbours' values, for one vector of stream values
nbmean <- function(v) rowMeans(matrix(v[as.vector(nb)], nrow = length(v)))
# one Spearman correlation per posterior draw -> a posterior distribution of coherence
rho_draw <- function(M) apply(M, 1, function(v) cor(v, nbmean(v), method = "spearman"))
T_m <- S_m + B_m   # downstream seasonal contrast, per draw
sp <- bind_rows(tibble(q = "inherited seasonal contrast S", rho = rho_draw(S_m)),
                tibble(q = "beaver seasonal modification B", rho = rho_draw(B_m)),
                tibble(q = "downstream seasonal contrast T", rho = rho_draw(T_m))) |>
  mutate(q = factor(q, levels = c("beaver seasonal modification B", "downstream seasonal contrast T",
                                  "inherited seasonal contrast S")))
sp_med <- sp |> group_by(q) |> summarise(m = median(rho))
ps <- ggplot(sp, aes(x = rho, y = q)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
  stat_halfeye(.width = c(.66, .95), point_interval = "median_qi", fill = "grey80") +
  labs(x = "Spearman correlation with the mean of the 8 nearest neighbours", y = NULL) +
  theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank())
ggsave(here("results", "figures", "plan", "fig_spatial.png"), ps, width = 8, height = 3.5, dpi = 300)
ps
```

::: {.callout-note}
## Key message of @fig-spatial

-   **The inherited seasonal contrast is regional.** Neighbour correlation of $S$: `r sprintf("%.2f", sp_med$m[sp_med$q == "inherited seasonal contrast S"])`.
-   **The downstream contrast inherits that structure.** $T$: `r sprintf("%.2f", sp_med$m[sp_med$q == "downstream seasonal contrast T"])`.
-   **The beaver seasonal modification is local.** $B$: `r sprintf("%.2f", sp_med$m[sp_med$q == "beaver seasonal modification B"])`, credible interval spanning zero.
-   **Implication.** Neighbouring streams inherit the same seasonal signal but do not modify it in the same way: each beaver-engineered stream reinforces, dampens or reverses a regionally organised signal on its own terms. This rules out a regional gradient masquerading as a beaver effect, and it is the empirical basis for calling beaver ponds *local* control points. Sentence for the Discussion as above.
:::

# 8. Model validation: out-of-sample prediction

::: {.callout-note appearance="simple"}
**Question:** Can the SCM predict downstream DOC, and the local change, at a stream it has not seen?
:::

The manuscript claims that the model can be used to anticipate the effect of beaver-engineering on stream DOC in Switzerland. That claim needs an out-of-sample test. Two are reported for `downstream_DOC_me`: (i) Pareto-smoothed importance-sampling leave-one-out cross-validation (PSIS-LOO), in which every observed downstream concentration is predicted as if it had been left out of the fit; and (ii) a hold-out test, in which the SCM is refitted on 80% of the streams and predicts downstream DOC at the remaining 20% (`r if (file.exists(here("results","me_holdout.rds"))) readRDS(here("results","me_holdout.rds"))$n_test_streams else "…"` streams), which it has never seen. At an unseen stream the site effect is unknown and the mediators are mostly unmeasured (they enter at their mean), so the hold-out prediction is what a user in Switzerland would actually have: upstream DOC, the dams, the channel gradient and solar radiation. For both tests, the derived $\Delta\text{DOC}$ is scored as well, because that is the quantity the claim is about; it is computed per posterior draw $d$ and sampling $i$ as

$$
\Delta\text{DOC}^{(d)}_i = \mu^{(d)}_i - \text{DOC}_{up,i}
$$ {#eq-derived-plan}

with $\mu_i$ the SCM's expected downstream DOC. Details of the derivation in `notebook/correspondence/scm_outcome_choice.qmd`.

**Manuscript placement: Appendix S1 (Figure S13), with one sentence in the Methods and one in the Discussion.**

```{r validation}
#| echo: false
pred_path <- here("results", "me_predictions.rds"); hold_path <- here("results", "me_holdout.rds")
have_pred <- file.exists(pred_path); have_hold <- file.exists(hold_path)
if (have_pred) P <- readRDS(pred_path)
if (have_hold) {
  H <- readRDS(hold_path)
  y_h <- H$obs_o_test; mu_h <- H$mu_o_test; q_h <- H$q_o_test
  d_obs_h <- H$obs_d_test; d_pred_h <- H$dd_test_pred_med; q_dh <- H$dd_q_test
}
if (have_pred && have_hold) {
  met <- function(y, mu, q = NULL) c(R2 = 1 - sum((y - mu)^2) / sum((y - mean(y))^2), RMSE = sqrt(mean((y - mu)^2)),
                                     cover = if (is.null(q)) NA else mean(y >= q[1, ] & y <= q[2, ]))
  # in-sample, full model (all 158 streams, site effects known)
  mu_in <- colMeans(P$O$ep_sub[, P$obsO]); q_in <- apply(P$O$pp_sub[, P$obsO], 2, quantile, c(.05, .95))
  in_full <- met(P$y_out, mu_in, q_in)
  ib_full <- match(P$both, P$obsO)
  d_in_full <- met(P$d_obs, colMeans(P$O$ep_sub[, P$both]) - P$up_both, apply(sweep(P$O$pp_sub[, P$both], 2, P$up_both), 2, quantile, c(.05, .95)))
  # LOO, full model
  loo_full <- met(P$y_out, P$O$mu_loo, P$O$q_loo)
  loo_d <- met(P$d_obs, P$O$mu_loo[ib_full] - P$up_both, P$O$dd_q_loo)
  n_all_streams <- length(unique(read.csv(here("data", "processed", "m6.csv"))$site))
  n_train_streams <- n_all_streams - H$n_test_streams
  f2 <- function(x) sprintf("%.2f", x); fp <- function(x) sprintf("%.0f%%", 100 * x)
  val_msg <- sprintf("- **Downstream DOC at an unseen stream is predictable.** Hold-out R² = %.2f, RMSE %.2f mg L⁻¹ (spread of the data %.2f), 90%% predictive intervals containing %.0f%% of the unseen observations — calibrated.\n- **No overfitting.** In-sample R² %.2f, leave-one-out %.2f (full model); %.2f in sample for the hold-out fit. The R² across all samplings is capped by a few very high concentrations in the heavy tail (RMSE %.1f against a spread of %.1f), not by unseen streams.\n- **The local change is not predictable.** Derived ΔDOC: hold-out R² = %.2f, RMSE %.2f against %.2f for predicting its mean.",
    H$O_out["R2"], H$O_out["RMSE"], H$O_out["RMSE_null"], 100 * H$O_out["cover90"], in_full["R2"], loo_full["R2"], H$O_in["R2"], loo_full["RMSE"], sd(P$y_out), H$Od_out["R2"], H$Od_out["RMSE"], H$Od_out["RMSE_null"])
  kable(tribble(
    ~Quantity, ~Test, ~`In / out of sample`, ~`R²`, ~`RMSE (mg L⁻¹)`, ~`RMSE of predicting the mean`, ~`90% interval coverage`,
    "Downstream DOC", sprintf("Full model, all %d samplings", length(P$y_out)), "in-sample", f2(in_full["R2"]), f2(in_full["RMSE"]), f2(sd(P$y_out)), fp(in_full["cover"]),
    "Downstream DOC", "Full model, PSIS-LOO", "out-of-sample", f2(loo_full["R2"]), f2(loo_full["RMSE"]), f2(sd(P$y_out)), fp(loo_full["cover"]),
    "Downstream DOC", sprintf("Hold-out fit, %d training streams", n_train_streams), "in-sample", f2(H$O_in["R2"]), f2(H$O_in["RMSE"]), f2(H$O_in["RMSE_null"]), fp(H$O_in["cover90"]),
    "Downstream DOC", sprintf("Hold-out fit, %d unseen streams (%d samplings)", H$n_test_streams, length(y_h)), "out-of-sample", f2(H$O_out["R2"]), f2(H$O_out["RMSE"]), f2(H$O_out["RMSE_null"]), fp(H$O_out["cover90"]),
    "Derived ΔDOC", sprintf("Full model, all %d samplings", length(P$d_obs)), "in-sample", f2(d_in_full["R2"]), f2(d_in_full["RMSE"]), f2(sd(P$d_obs)), fp(d_in_full["cover"]),
    "Derived ΔDOC", "Full model, PSIS-LOO", "out-of-sample", f2(loo_d["R2"]), f2(loo_d["RMSE"]), f2(sd(P$d_obs)), fp(loo_d["cover"]),
    "Derived ΔDOC", "Hold-out fit, training streams", "in-sample", f2(H$Od_in["R2"]), f2(H$Od_in["RMSE"]), f2(H$Od_in["RMSE_null"]), "—",
    "Derived ΔDOC", "Hold-out fit, unseen streams", "out-of-sample", f2(H$Od_out["R2"]), f2(H$Od_out["RMSE"]), f2(H$Od_out["RMSE_null"]), fp(mean(d_obs_h >= q_dh[1, ] & d_obs_h <= q_dh[2, ]))),
    caption = sprintf("Predictive performance of downstream_DOC_me, in and out of sample. In-sample predictions include the stream random effects (the model has seen the stream); out-of-sample predictions do not. Hold-out fit: max R-hat %.3f.", H$rhat["down"]))
} else { cat("Predictions not yet computed (notebook/fitting/me_predictions.R, notebook/fitting/me_holdout.R)."); val_msg <- "- pending" }
```

```{r fig-holdout}
#| fig-height: 4.5
#| fig-cap: "**Hold-out test of downstream_DOC_me.** The model was refitted on 80% of the streams and predicts the other 20%, which it has never seen. (a) Predicted (posterior mean, with 90% predictive interval) against observed downstream DOC at the unseen streams; dashed line: 1:1. (b) The derived local response, predicted downstream DOC minus upstream DOC, against the observed ΔDOC at the same samplings, with its own 90% predictive interval. $\\Delta\\text{DOC}_s$ is defined in @tbl-variables."
if (have_hold) {
  pa <- ggplot(tibble(y = y_h, mu = mu_h, lo = q_h[1, ], hi = q_h[2, ]), aes(y, mu)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
    geom_linerange(aes(ymin = lo, ymax = hi), colour = "grey70", alpha = .6) + geom_point(size = 1.6, colour = "#3B9AD1") +
    labs(x = expression("Observed downstream DOC (mg "*L^-1*")"), y = expression("Predicted downstream DOC (mg "*L^-1*")")) +
    theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank())
  pb <- ggplot(tibble(o = d_obs_h, p = d_pred_h, lo = q_dh[1, ], hi = q_dh[2, ]), aes(o, p)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
    geom_hline(yintercept = 0, colour = "grey80") + geom_vline(xintercept = 0, colour = "grey80") +
    geom_linerange(aes(ymin = lo, ymax = hi), colour = "grey70", alpha = .6) + geom_point(size = 1.6, colour = "#3B9AD1") +
    labs(x = expression("Observed "*Delta*"DOC (mg "*L^-1*")"), y = expression("Predicted "*Delta*"DOC (mg "*L^-1*")")) +
    theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank())
  figH <- pa + pb + plot_annotation(tag_levels = "a")
  ggsave(here("results", "figures", "plan", "fig_holdout.png"), figH, width = 9, height = 4.5, dpi = 300)
  figH
}
```

::: {.callout-important}
## Key message of @fig-holdout, and what the manuscript may claim

`r val_msg`
-   **Implication.** Almost all predictive skill is inheritance (downstream DOC is `r sprintf("%.0f", 100*gme(tO,"bsp_DOCout_miDOC_input")$mean)`% upstream DOC). The manuscript can claim that the model predicts downstream DOC at unsampled Swiss streams from the upstream concentration and the measured drivers, and that the beaver-mediated change will be small relative to that inherited concentration and of stream-specific sign — it cannot claim to forecast *where* a stream becomes a source or a sink from dams, residence time or solar radiation alone (Q2, Q4).
:::

# 9. Structural and causal sensitivity

The SCM of Q2 rests on causal assumptions that the data cannot fully verify: the arrows we drew, the arrows we left out, and the absence of unmeasured common causes. This section tests those assumptions in the tradition of @shipley2016 and Grace's causal-knowledge analysis: alternative structures (A), unmeasured confounding (B), relationships the DAG says must be null (C), robustness to modelling choices (D), and the qualitative causal-knowledge table (E). One fact conditions everything below: **the primary producers were measured at 16 streams, all in summer, and at none of them was the water residence time measured.** The producer paths therefore rest on 16 samplings, and the water-residence-time → producer paths are informed only by the priors and the imputation model. The sensitivity analyses say how much that matters.

**Manuscript placement: Appendix S1 (Figures S14–S17, Table S2), with one paragraph in the Methods.**

```{r sens-setup}
#| echo: false
library(brms); library(dagitty)
m6 <- read.csv(here("data", "processed", "m6.csv")) |> mutate(across(c(season, site), as.factor))
sens_path <- here("results", "dag_sensitivity_me.rds"); have_sens <- file.exists(sens_path)
if (have_sens) SENS <- readRDS(sens_path)
med_names <- c(bsp_DOCout_miDOC_input = "inheritance slope", bsp_DOCout_mimacrophy_abun_z = "macrophytes",
               bsp_DOCout_miplankton_abun_z = "phytoplankton", bsp_DOCout_micover_litter_z = "soil litter cover",
               bsp_DOCout_miwater_res_time_z = "water residence time")
```

## A. Alternative DAGs

Four structures, all keeping the mediator core of Figure 4, which is now DAG 0 itself:

-   **DAG 0** — the manuscript's model (@fig-q2): upstream DOC → macrophytes and → phytoplankton are included, alongside the upstream measurement error.
-   **DAG 1** — the two upstream → producer links **removed**: the structure used before this revision. Does keeping them change the mediator story, or only the picture of what drives the producers?
-   **DAG 2** — solar radiation acts directly on downstream DOC (photodegradation), not only through the producers.
-   **DAG 3** — season is a common cause of the producers and of downstream DOC, left unblocked in DAG 0.

Each DAG implies conditional independencies that can be checked on the data with `dagitty` [@shipley2016]. Each test is run on the rows complete for the variables it involves, and reported as the partial correlation with its 95% interval (the only frequentist quantity in this document, kept because the test is the standard one for this purpose). Structures that add an arrow *remove* an implied independence, so a DAG-0 independence that fails in the data, and that an alternative DAG does not imply, is evidence for that alternative.

```{r fig-dag-tests}
#| fig-height: 6
#| fig-cap: "**Conditional independencies implied by DAG 0, tested on the data.** Each row is one independence the manuscript's DAG implies (X ⊥ Y | Z); the point is the partial correlation on the rows complete for the variables involved, with its 95% interval and the number of samplings used. Filled points: the interval excludes zero (the independence fails). The right-hand label says which alternative DAG(s) would drop that independence, i.e. which alternative a failure supports. Tests with fewer than 10 complete rows are not run."
dag_txt <- list(
  dag0 = "dag { n_dams -> wrt; dam_height -> wrt; slope -> wrt; solar -> plankton; solar -> macro; solar -> litter;
           wrt -> plankton; wrt -> macro; wrt -> DOC_out; plankton -> DOC_out; macro -> DOC_out; litter -> DOC_out; DOC_in -> DOC_out;
           DOC_in -> plankton; DOC_in -> macro }",
  dag1 = "dag { n_dams -> wrt; dam_height -> wrt; slope -> wrt; solar -> plankton; solar -> macro; solar -> litter;
           wrt -> plankton; wrt -> macro; wrt -> DOC_out; plankton -> DOC_out; macro -> DOC_out; litter -> DOC_out; DOC_in -> DOC_out }",
  dag2 = "dag { n_dams -> wrt; dam_height -> wrt; slope -> wrt; solar -> plankton; solar -> macro; solar -> litter;
           wrt -> plankton; wrt -> macro; wrt -> DOC_out; plankton -> DOC_out; macro -> DOC_out; litter -> DOC_out; DOC_in -> DOC_out;
           DOC_in -> plankton; DOC_in -> macro; solar -> DOC_out }",
  dag3 = "dag { n_dams -> wrt; dam_height -> wrt; slope -> wrt; solar -> plankton; solar -> macro; solar -> litter;
           wrt -> plankton; wrt -> macro; wrt -> DOC_out; plankton -> DOC_out; macro -> DOC_out; litter -> DOC_out; DOC_in -> DOC_out;
           DOC_in -> plankton; DOC_in -> macro; season -> plankton; season -> macro; season -> litter; season -> DOC_out }")
gs <- lapply(dag_txt, dagitty)
dd <- m6 |> transmute(n_dams = n_dams_z, dam_height = dam_height_z, slope = slope_z, solar = solar_z,
                      plankton = plankton_abun_z, macro = macrophy_abun_z, litter = cover_litter_z, wrt = water_res_time_z,
                      DOC_out, DOC_in = DOC_input, season = as.numeric(season == "summer"))
ci_lab <- function(x) paste0(x$X, " ⊥ ", x$Y, if (length(x$Z)) paste0(" | ", paste(unlist(x$Z), collapse = ", ")) else "")
ci0 <- impliedConditionalIndependencies(gs$dag0)
implied_by <- function(g, lab) lab %in% sapply(impliedConditionalIndependencies(g), ci_lab)
tests <- lapply(ci0, function(x) {
  v <- unlist(c(x$X, x$Y, x$Z)); sub <- dd[complete.cases(dd[, v]), v, drop = FALSE]
  n <- nrow(sub); lab <- ci_lab(x)
  out <- tibble(test = lab, n = n, r = NA_real_, lo = NA_real_, hi = NA_real_)
  if (n >= 10 && all(apply(sub, 2, sd) > 0)) {
    t <- tryCatch(localTests(gs$dag0, sub, type = "cis", tests = list(x)), error = function(e) NULL)
    if (!is.null(t)) out <- tibble(test = lab, n = n, r = t$estimate, lo = t$`2.5%`, hi = t$`97.5%`) }
  out$dropped_by <- paste(names(gs)[-1][!sapply(gs[-1], implied_by, lab = lab)], collapse = ", ")
  out })
tests <- bind_rows(tests) |> filter(!is.na(r)) |>
  mutate(fails = lo > 0 | hi < 0, dropped_by = ifelse(dropped_by == "", "none", gsub("dag", "DAG ", dropped_by)),
         test = fct_reorder(paste0(test, "  (n = ", n, ")"), r))
ggplot(tests, aes(x = r, y = test)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
  geom_linerange(aes(xmin = lo, xmax = hi), colour = "grey60") +
  geom_point(aes(fill = fails), shape = 21, size = 2.6, colour = "black") +
  geom_text(aes(x = 1.05, label = dropped_by), hjust = 0, size = 3, colour = "grey30") +
  scale_fill_manual(values = c(`TRUE` = "black", `FALSE` = "white"), labels = c(`TRUE` = "fails", `FALSE` = "holds"), name = NULL) +
  coord_cartesian(xlim = c(-1, 1.6), clip = "off") +
  labs(x = "Partial correlation implied to be zero (95% interval)", y = NULL) +
  theme_bw(base_size = 11) + theme(panel.grid.minor = element_blank(), legend.position = "top") -> p_tests
ggsave(here("results", "figures", "plan", "fig_dag_tests.png"), p_tests, width = 9, height = 6, dpi = 300, bg = "white"); p_tests
```

```{r fig-dag-compare}
#| fig-height: 5
#| fig-cap: "**The four DAGs fitted as multivariate models.** (a) Out-of-sample fit for downstream DOC by PSIS-LOO: expected log predictive density relative to the best model (0 = best), with standard error; a difference smaller than about 4 is not meaningful. (b) The mediator effects on downstream DOC under each structure: posterior median with 66% and 95% credible intervals. Fits with 4 chains × 2500 post-warm-up draws; max R-hat in the caption of the callout."
if (have_sens) {
  cmp <- as.data.frame(SENS$loo_compare); if (!"model" %in% names(cmp)) cmp$model <- rownames(cmp)
  cmp <- cmp |> select(model, elpd_diff, se_diff) |> mutate(model = factor(gsub("dag", "DAG ", model), levels = paste("DAG", 0:3)))
  pa <- ggplot(cmp, aes(x = elpd_diff, y = fct_rev(model))) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
    geom_pointrange(aes(xmin = elpd_diff - se_diff, xmax = elpd_diff + se_diff), size = .5) +
    labs(x = "ELPD difference to the best DAG (± SE)", y = NULL) + theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank())
  eff <- bind_rows(lapply(c("dag0","dag1","dag2","dag3"), function(nm) as_tibble(SENS$models[[nm]]$draws) |>
    pivot_longer(everything(), names_to = "par", values_to = "v") |> mutate(model = gsub("dag", "DAG ", nm)))) |>
    filter(par != "bsp_DOCout_miDOC_input") |> mutate(par = factor(med_names[par], levels = med_names[-1]))
  pb <- ggplot(eff, aes(x = v, y = fct_rev(factor(model)), colour = model)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
    stat_pointinterval(.width = c(.66, .95), point_interval = "median_qi") +
    facet_wrap(~ par, ncol = 2, scales = "free_x") + scale_colour_brewer(palette = "Dark2", guide = "none") +
    labs(x = "Standardized effect on downstream DOC", y = NULL) + theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank())
  p_cmp <- pa + pb + plot_layout(widths = c(1, 1.6)) + plot_annotation(tag_levels = "a")
  ggsave(here("results", "figures", "plan", "fig_dag_compare.png"), p_cmp, width = 9, height = 5, dpi = 300, bg = "white"); p_cmp
} else cat("DAG fits not yet available (notebook/R/dag_sensitivity.R).")
```

```{r dag-new-paths}
#| echo: false
fl <- tests |> filter(fails); fx <- fl |> filter(dropped_by != "none")
strip <- function(x) sub("  \\(n.*", "", as.character(x))
a_msg <- sprintf("- **Conditional-independence tests** (@fig-dag-tests): %d of the %d testable independencies fail. Those that an alternative DAG would drop: %s. The remaining failures are among the exogenous drivers (e.g. %s): the dams, the gradient, solar radiation and the upstream DOC share landscape causes, which the SCM tolerates because none of them lies on a causal path to another, but which the Methods should state.",
  nrow(fl), nrow(tests),
  if (nrow(fx)) paste(sprintf("%s (r = %.2f [%.2f, %.2f], n = %d; supports %s)", strip(fx$test), fx$r, fx$lo, fx$hi, fx$n, fx$dropped_by), collapse = "; ") else "none",
  paste(head(strip(fl$test[fl$dropped_by == "none"]), 3), collapse = "; "))
if (have_sens) {
  np_tO <- function(par) { g <- gme(tO, par); sprintf("%.2f [%.2f, %.2f], P(> 0) = %.2f", g$mean, g$lo, g$hi, g$p_pos) }
  sup_tO <- function(par) { p <- gme(tO, par)$p_pos; if (p > 0.90) "supported (P > 0.90)" else if (p < 0.10) "supported, negative (P < 0.10)" else "no-evidence" }
  np <- function(nm, par) { x <- SENS$models[[nm]]$extra[, par]; sprintf("%.2f [%.2f, %.2f], P(> 0) = %.2f", median(x), quantile(x, .025), quantile(x, .975), mean(x > 0)) }
  pp <- function(nm, par) mean(SENS$models[[nm]]$extra[, par] > 0)
  e2 <- colnames(SENS$models$dag2$extra); p_sol <- e2[grepl("DOCout.*misolar", e2)][1]
  sup <- function(p) if (p > 0.90) "supported (P > 0.90)" else if (p < 0.10) "supported, negative (P < 0.10)" else "no-evidence"
  rh <- sapply(SENS$models[c("dag0","dag1","dag2","dag3")], `[[`, "rhat")
  cm <- as.data.frame(SENS$loo_compare); if (!"model" %in% names(cm)) cm$model <- rownames(cm)
  best <- gsub("dag", "DAG ", cm$model[1]); d0 <- cm[cm$model == "dag0", ]
  dag1_msg <- sprintf("- **DAG 1 (upstream → producer links removed):** in DAG 0 those links are %s (macrophytes) and %s (phytoplankton) per mg L⁻¹ of upstream DOC — %s and %s. Removing them (DAG 1) leaves the mediator effects on downstream DOC indistinguishable from DAG 0 (@fig-dag-compare b), and the LOO fit is no better without them (@fig-dag-compare a): the links matter for what predicts the producers, not for what the producers do to downstream DOC.",
                      np_tO("bsp_macrophyabunz_miDOC_input"), np_tO("bsp_planktonabunz_miDOC_input"),
                      sup_tO("bsp_macrophyabunz_miDOC_input"), sup_tO("bsp_planktonabunz_miDOC_input"))
  dag2_msg <- sprintf("- **DAG 2 (direct solar → DOC):** on top of DAG 0, the direct path is %s — %s. Under DAG 2 the producer effects shrink somewhat (part of the solar effect no longer has to travel through them) but keep their sign.",
                      np("dag2", p_sol), sup(pp("dag2", p_sol)))
  loo_msg <- sprintf("- **No alternative structure fits meaningfully better.** By PSIS-LOO the nominally best structure is %s; the manuscript's DAG 0 sits %.1f ELPD below it (SE %.1f), a difference of about %.1f standard errors and below the usual threshold of 4 ELPD for a meaningful preference. Max R-hat across the four fits: %.3f.",
                     best, -d0$elpd_diff, d0$se_diff, abs(d0$elpd_diff / d0$se_diff), max(rh))
}
```

::: {.callout-important}
## Conclusions — A. Alternative DAGs

`r if (have_sens) loo_msg else "- pending"`
`r if (have_sens) dag1_msg else ""`
`r if (have_sens) dag2_msg else ""`
`r if (have_sens) "- **DAG 3 (season as confounder):** because the producers were measured in summer only, the season term in the producer equations is informed by priors and imputation alone; the DAG-3 mediator effects should be read as 'unchanged under a prior-driven adjustment', not as a data-based test." else ""`
`r a_msg`
-   **Tests involving the producers rest on 16 rows** and are wide; the DAG-1 comparison above is real but weak, and it is the DAG-1 *fit* (@fig-dag-compare) that says whether removing the links changes anything that matters.
-   **What this settles:** the mediator effects are not an artefact of the particular arrows drawn, whether or not the two upstream → producer links are included. **What it cannot settle:** whether the producer paths, or the two new links themselves, are right at all — that rests on 16 summer samplings and on the priors, which B and D address.
:::

## B. Unmeasured confounding of the mediator → DOC paths

How strong would an unobserved common cause of a producer and downstream DOC have to be to erase the producer's effect? Following @cinelli2020 as implemented in `sensemakr`, the outcome equation is fitted as a linear model on the 16 producer streams (the only rows with the producers; water residence time is absent there and cannot be included), with solar radiation as the benchmark covariate: the contour plots show the adjusted effect for every combination of confounder strength (its partial R² with the treatment and with the outcome), and the marks show a confounder as strong as one, two and three times solar radiation. The **robustness value** is the strength (partial R² with both) at which the effect reaches zero.

```{r fig-sensemakr}
#| fig-height: 4
#| fig-cap: "**Robustness of each producer → downstream DOC path to an unmeasured confounder.** Contours: the adjusted effect of the producer on downstream DOC for a confounder with the given partial R² with the producer (x) and with downstream DOC (y); the triangle is the unadjusted estimate; the red marks are confounders 1×, 2× and 3× as strong as solar radiation. Linear outcome equation on the 16 producer streams (summer). Note that this linear fit does not carry the priors, the measurement error or the water-residence-time path of the SCM — it is the sensitivity of the raw 16-stream relationship."
library(sensemakr)
dl <- m6 |> filter(!is.na(plankton_abun_z)) |> select(DOC_out, DOC_input, macrophy_abun_z, plankton_abun_z, cover_litter_z, solar_z)
lmo <- lm(DOC_out ~ DOC_input + macrophy_abun_z + plankton_abun_z + cover_litter_z + solar_z, data = dl)
paths <- c(macrophy_abun_z = "macrophytes", plankton_abun_z = "phytoplankton", cover_litter_z = "soil litter cover")
rv <- sapply(names(paths), function(tr) { s <- sensemakr(lmo, treatment = tr, benchmark_covariates = "solar_z", kd = 1:3)
  c(est = s$sensitivity_stats$estimate, rv = s$sensitivity_stats$rv_q, r2 = s$sensitivity_stats$r2yd.x, r2_solar = s$bounds$r2dz.x[1]) })
draw_sm <- function() { par(mfrow = c(1, 3), mar = c(4, 4, 3, 1))
  for (tr in names(paths)) { s <- sensemakr(lmo, treatment = tr, benchmark_covariates = "solar_z", kd = 1:3)
    ovb_contour_plot(s, sensitivity.of = "estimate", main = paths[tr]) } }
png(here("results", "figures", "plan", "fig_sensemakr.png"), width = 9, height = 4, units = "in", res = 300); draw_sm(); invisible(dev.off())
draw_sm()
```

::: {.callout-important}
## Conclusions — B. Unmeasured confounding

-   **The raw 16-stream relationship is not the SCM's.** In the linear outcome equation the estimates are `r sprintf("%.2f", rv["est","macrophy_abun_z"])` (macrophytes), `r sprintf("%.2f", rv["est","plankton_abun_z"])` (phytoplankton) and `r sprintf("%.2f", rv["est","cover_litter_z"])` (litter), against `r sprintf("+%.2f", gme(tO,"bsp_DOCout_mimacrophy_abun_z")$mean)`, `r sprintf("+%.2f", gme(tO,"bsp_DOCout_miplankton_abun_z")$mean)` and `r sprintf("+%.2f", gme(tO,"bsp_DOCout_micover_litter_z")$mean)` in the SCM. The SCM's positive producer effects are therefore carried substantially by the informative priors (Normal(0.2, 0.075) on each producer path) and by the hierarchical structure, not by these 16 samplings alone. This is the single most important sensitivity result of the section, and it is confirmed by the prior-sensitivity fits in D.
-   **Robustness values** (strength a confounder needs to erase the raw effect): macrophytes `r sprintf("%.2f", rv["rv","macrophy_abun_z"])`, litter `r sprintf("%.2f", rv["rv","cover_litter_z"])`, phytoplankton `r sprintf("%.2f", rv["rv","plankton_abun_z"])`; solar radiation, the strongest measured covariate, explains only `r sprintf("%.0f%%", 100 * rv["r2_solar","macrophy_abun_z"])`–`r sprintf("%.0f%%", 100 * max(rv["r2_solar",]))` of the residual variance of any producer, so a confounder several times stronger than solar would be needed to remove the macrophyte or litter relationship, and almost none to remove the phytoplankton one.
-   **Water residence time cannot be assessed** this way: it was never measured at a producer stream.
-   **What the manuscript should say:** the producer → DOC paths are consistent with the literature-based priors and are not overturned by plausible confounding, but with 16 summer samplings they are prior-informed estimates rather than data-driven ones; the Methods should state the prior for each path and D should be cited.
:::

## C. Falsification tests and a placebo outcome

The DAG says some relationships must be absent. If they are present, either the DAG or the data are wrong somewhere. Three implied nulls among the exogenous drivers, tested as Bayesian regressions on all samplings, and one placebo outcome: upstream DOC, which lies above the dams and cannot be affected by beaver-engineering — if the number of dams, the dam height or the water residence time "predicted" upstream DOC, it would reveal a confounder (a landscape gradient that sets both the dams and the DOC).

```{r fig-negative-controls}
#| fig-height: 3.8
#| fig-cap: "**Relationships the DAG says must be null.** Posterior distributions of the standardized coefficients of four Bayesian Student-t regressions on all samplings (Normal(0, 1) priors on the slopes); point: median, bars: 66% and 95% credible intervals; dashed line: zero. The last three rows are the placebo outcome — upstream DOC regressed on the beaver-engineering variables, which cannot causally affect it."
nc_fit <- function(f, tag) brm(f, data = m6, family = student(), prior = c(prior(normal(0, 1), class = "b")),
  chains = 4, iter = 3000, warmup = 1000, seed = 7, backend = "cmdstanr", refresh = 0,
  file = here("results", "models_fit", paste0("negctrl_", tag)))
nc <- list(
  `dam height ~ solar radiation`      = nc_fit(dam_height_z ~ solar_z, "dh_solar"),
  `number of dams ~ upstream DOC`     = nc_fit(n_dams_z ~ DOC_input_z, "nd_docin"),
  `channel gradient ~ solar radiation`= nc_fit(slope_z ~ solar_z, "slope_solar"),
  `placebo: upstream DOC ~ dams, height, residence time` = nc_fit(DOC_input_z ~ n_dams_z + dam_height_z + water_res_time_z, "placebo"))
nc_draws <- bind_rows(lapply(names(nc), function(nm) { d <- as_draws_df(nc[[nm]]); b <- grep("^b_", names(d), value = TRUE); b <- b[b != "b_Intercept"]
  as_tibble(d[, b]) |> pivot_longer(everything(), names_to = "coef", values_to = "v") |>
    mutate(test = nm, coef = sub("^b_", "", coef), label = if (length(b) == 1) nm else paste0("placebo: upstream DOC ~ ", gsub("_z$", "", coef))) }))
nc_sum <- nc_draws |> group_by(label) |> summarise(med = median(v), lo = quantile(v, .025), hi = quantile(v, .975), p_pos = mean(v > 0))
ggplot(nc_draws, aes(x = v, y = fct_rev(factor(label, levels = unique(label))))) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
  stat_halfeye(.width = c(.66, .95), point_interval = "median_qi", fill = "grey80") +
  labs(x = "Standardized coefficient", y = NULL) + theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank()) -> p_nc
ggsave(here("results", "figures", "plan", "fig_negative_controls.png"), p_nc, width = 9, height = 3.8, dpi = 300, bg = "white"); p_nc
```

::: {.callout-important}
## Conclusions — C. Falsification tests

`r paste(sprintf("- **%s:** %.2f [%.2f, %.2f], P(> 0) = %.2f — %s.", nc_sum$label, nc_sum$med, nc_sum$lo, nc_sum$hi, nc_sum$p_pos, ifelse(nc_sum$lo > 0 | nc_sum$hi < 0, "**the implied null fails**", "consistent with the implied null")), collapse = "\n")`
`r pl <- nc_sum |> filter(grepl("placebo", label)); worst <- pl |> arrange(desc(abs(med))) |> slice(1); sprintf("- **Placebo outcome:** the largest association between a beaver-engineering variable and upstream DOC is %s (%.2f SD); %s. Any confounding of the dam → residence time → DOC chain by the catchment's DOC regime is therefore at most of that size, an order of magnitude below the inheritance slope.", sub("placebo: upstream DOC ~ ", "", worst$label), worst$med, if (worst$lo > 0 | worst$hi < 0) "credibly non-zero but small" else "not distinguishable from zero")`
-   **Reading:** a failed null among the exogenous drivers (dams, gradient, solar) means those drivers share a landscape cause — which the SCM tolerates, because they are not on a causal path to each other, but which a reader should know. A failed placebo would mean that the beaver-engineering variables are confounded with the DOC regime of the catchment; the result above says whether that is the case.
:::

## D. Robustness to modelling choices

Two checks are already in section 4 (outcome parameterization; inheritance slope fixed at 1). Two remain: the prior scale, and spatial autocorrelation of the residuals.

```{r fig-prior-sens}
#| fig-height: 4
#| fig-cap: "**Prior sensitivity.** Mediator effects on downstream DOC when the standard deviation of every informative prior is halved, kept as in the manuscript, or doubled; posterior median with 66% and 95% credible intervals. A path whose estimate follows the prior scale is prior-driven; a path that stays put is data-driven."
if (have_sens) {
  ps <- bind_rows(lapply(c(prior_half = "prior_half", manuscript = "dag0", prior_double = "prior_double"), function(nm) as_tibble(SENS$models[[nm]]$draws) |>
    pivot_longer(everything(), names_to = "par", values_to = "v")), .id = "prior") |>
    mutate(prior = factor(prior, levels = c("prior_half", "manuscript", "prior_double"), labels = c("sd halved", "manuscript", "sd doubled")),
           par = factor(med_names[par], levels = med_names))
  ggplot(ps, aes(x = v, y = fct_rev(prior), colour = prior)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
    stat_pointinterval(.width = c(.66, .95), point_interval = "median_qi") +
    facet_wrap(~ par, scales = "free_x", ncol = 5) + scale_colour_manual(values = c("#E8603C", "black", "#3B9AD1"), guide = "none") +
    labs(x = "Effect on downstream DOC", y = NULL) + theme_bw(base_size = 11) + theme(panel.grid.minor = element_blank()) -> p_ps
  ggsave(here("results", "figures", "plan", "fig_prior_sens.png"), p_ps, width = 10, height = 4, dpi = 300, bg = "white"); p_ps
} else cat("Prior-sensitivity fits not yet available.")
```

```{r fig-spatial-resid}
#| fig-height: 2.8
#| fig-cap: "**Spatial autocorrelation of the residuals of downstream DOC (model downstream_DOC_me).** For each posterior draw, the Spearman correlation between each sampling's residual (observed minus expected downstream DOC) and the mean residual of its 8 nearest neighbouring samplings, using the stream coordinates; point: median, bars: 66% and 95% credible intervals; dashed line: no spatial autocorrelation."
PME <- readRDS(here("results", "me_predictions.rds")); obs2 <- PME$obsO
lon <- m6$x_upstream[obs2]; lat <- m6$y_upstream[obs2]; okc <- !is.na(lon)
kmx2 <- (lon - mean(lon, na.rm = TRUE)) * 111.32 * cos(mean(lat, na.rm = TRUE) * pi / 180); kmy2 <- (lat - mean(lat, na.rm = TRUE)) * 110.57
D2 <- as.matrix(dist(cbind(kmx2[okc], kmy2[okc]))); diag(D2) <- Inf
nb2 <- t(apply(D2, 1, function(r) order(r)[1:8]))
resid <- sweep(-PME$O$ep_sub[, obs2][, okc], 2, -m6$DOC_out[obs2][okc])
rho_res <- apply(resid, 1, function(r) cor(r, rowMeans(matrix(r[as.vector(nb2)], nrow = length(r))), method = "spearman"))
p_spatial_resid <- ggplot(tibble(rho = rho_res), aes(x = rho, y = 1)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
  stat_halfeye(.width = c(.66, .95), point_interval = "median_qi", fill = "grey80") +
  labs(x = "Spearman correlation of residuals with the mean of the 8 nearest neighbours", y = NULL) +
  theme_bw(base_size = 12) + theme(panel.grid.minor = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank())
ggsave(here("results", "figures", "plan", "fig_spatial_resid.png"), p_spatial_resid, width = 8, height = 2.8, dpi = 300, bg = "white")
p_spatial_resid
```

```{r prior-sens-numbers}
#| echo: false
if (have_sens) {
  pm <- ps |> group_by(par, prior) |> summarise(med = median(v), .groups = "drop") |> pivot_wider(names_from = prior, values_from = med)
  shift <- pm |> mutate(range = `sd doubled` - `sd halved`)
  most <- shift |> arrange(desc(abs(range))) |> slice(1)
  d_msg <- sprintf("- **Prior scale:** the inheritance slope does not move (%.2f, %.2f, %.2f from halved to doubled sd). The producer paths move with the prior: the largest shift is %s, from %.2f (sd halved) to %.2f (sd doubled). The path that keeps its sign when the prior is loosened is phytoplankton (%.2f with sd doubled); macrophytes and litter follow the prior. With 16 summer samplings, the producer effects are prior-informed estimates, the same conclusion as B, reached from the other side.",
                   pm$`sd halved`[1], pm$manuscript[1], pm$`sd doubled`[1], as.character(most$par), most$`sd halved`, most$`sd doubled`, pm$`sd doubled`[pm$par == "phytoplankton"])
} else d_msg <- "- pending"
```

::: {.callout-important}
## Conclusions — D. Modelling choices

`r d_msg`
-   **Spatial residuals:** the residual neighbour correlation is `r sprintf("%.2f [%.2f, %.2f]", median(rho_res), quantile(rho_res, .025), quantile(rho_res, .975))` — no spatial autocorrelation is left in downstream DOC once upstream DOC and the drivers are in the model, so the stream random effect is doing its job and a spatial term is not needed.
-   Together with section 4: outcome parameterization, inheritance slope, prior scale and space — the inheritance and the *absence* of a consistent local effect are robust; the *sign* of each producer path is where the model leans on prior knowledge.
:::

## E. Causal knowledge analysis

The qualitative frame the quantitative checks sit in (after Grace's causal-knowledge analysis; **the wording of the eight questions must be checked against the paper before this goes into Appendix S1**).

```{r}
#| label: tbl-cka
#| tbl-cap: "Causal knowledge analysis of the beaver–DOC system."
#| echo: false
kable(tribble(
  ~`Question`, ~`Answer for the beaver–DOC system`, ~`Where addressed`,
  "1. What causal quantity is sought?", "The local modification that beaver-engineering adds to downstream DOC, decomposed into its hydrological and biological pathways; and its size relative to the inherited seasonal contrast.", "Q2 (SCM), Q3 ($R_s$, $G$)",
  "2. What is the causal system and its boundary?", "The beaver-engineered stream from the first to the last dam; inputs are upstream DOC, solar radiation, channel gradient, dams; the floodplain and the catchment are outside, represented by upstream DOC and $S$.", "Figure 3b, DAG (Figure 4)",
  "3. What is assumed about the structure (arrows present)?", "Dams → residence time → producers → DOC; solar → producers → DOC; litter → DOC; upstream DOC inherited with slope near 1; upstream DOC → macrophytes (and, more weakly, → phytoplankton).", "Table S1 (13 hypotheses), section 2",
  "4. What is assumed absent (arrows omitted, no confounding)?", "No direct solar → DOC, no season confounding, no unmeasured common cause of producers and DOC.", "Section 9 A–C: none of the omitted arrows is supported; placebo and implied nulls in C",
  "5. How are the effects identified?", "By the DAG's adjustment sets, with measurement error on every concentration and imputation of missing predictors; the inheritance slope is estimated, not assumed.", "Methods; section 4 (robustness to slope = 1)",
  "6. Is the data adequate for the structure?", "158 streams for the inheritance and the direction; 16 summer streams for the producers, with no water residence time at those streams. The producer paths are the weak point.", "Section 9 B and D: prior-informed estimates",
  "7. How sensitive are the conclusions?", "Inheritance, the null average and the heterogeneity are robust to structure, priors, outcome parameterization and space; the sign of each producer path follows the prior scale.", "Sections 4, 8, 9",
  "8. To what does the result transport?", "To Swiss low-order single-thread streams with 1–19 dams; downstream DOC at an unseen stream is predictable (hold-out R² given in section 8), the sign of the local change is not.", "Section 8"))
```

::: {.callout-important}
## Conclusions — E. Causal knowledge analysis

-   The structure of Q2 is well supported where the data are dense (inheritance, direction, dams → residence time) and prior-dependent where they are thin (producers).
-   The manuscript should say so in one Methods paragraph: which arrows were tested and how (A, C), how strong a confounder would have to be (B), and that the producer effects follow the prior scale (D) — with Appendix S1 Figures S10–S15 and Table S2.
-   This is not a weakness to hide: it is the honest statement of what a national survey with a 16-stream mechanistic subset can and cannot identify, and it points to the next study — residence time and producers measured at the same streams, in both seasons.
:::

# 10. Q5 — Pathway decomposition of the local modification

::: {.callout-note appearance="simple"}
**Question:** Why do some streams enrich and others deplete DOC, and why does the same stream change role between seasons?
:::

The regional analysis (section 7) says an inherited DOC signal exists; the SCM (section 4) says which local processes modify it. What neither says is *why* a given stream ends up enriching or depleting, and why the same stream can change role between winter and summer. The approach is to use the fitted SCM itself: for every sampling, season and posterior draw, split the predicted local modification (predicted downstream DOC minus upstream DOC) into the contribution of each modelled pathway, and compare the balance between enriching and depleting streams and between seasons.

From @eq-derived-plan the local modification of sampling $i$ in draw $d$ is a sum of named parts:

$$
\begin{aligned}
L^{(d)}_i \;=\; \mu^{(d)}_i - \text{DOC}_{up,i}
&= \underbrace{(\beta_{inh}^{(d)} - 1)\,\text{DOC}_{up,i}}_{\text{inheritance shortfall}}
 + \underbrace{\beta^{(d)}_{pla} M_{pla,i}}_{\text{phytoplankton}}
 + \underbrace{\beta^{(d)}_{mac} M_{mac,i}}_{\text{macrophytes}} \\
&\quad + \underbrace{\beta^{(d)}_{lit} M_{lit,i}}_{\text{litter}}
 + \underbrace{\beta^{(d)}_{wrt} M_{wrt,i}}_{\text{hydrology}}
 + \underbrace{\alpha^{(d)} + u^{(d)}_{site(i)}}_{\text{unresolved}}
\end{aligned}
$$ {#eq-decomp}

The last term is what the model cannot attribute to a measured pathway: the intercept and the stream's own effect. It is reported as an **unresolved net removal/transformation component**, not assigned to any process. Two limits are stated before the results, because they decide how much this analysis can say:

-   **The producers were measured at 16 streams, in summer only, and water residence time at none of them.** At every other sampling the mediator values in @eq-decomp are the draw's *imputed* values, generated by the mediator sub-models from solar radiation and residence time — and, because the model is joint, partly from the outcome itself. The pathway contributions at imputed samplings are therefore model-implied, not measured, and can partly re-label the residual. The decomposition is shown for all samplings and, separately, for the 16 measured streams.
-   **The decomposition is only as good as the SCM's paths**, and section 9 showed the producer effects to be prior-informed.

**Manuscript placement: Appendix S1 (Figure S18), with two sentences in the Discussion.**

```{r decomp-load}
#| echo: false
dec_path <- here("results", "pathway_decomposition.rds"); have_dec <- file.exists(dec_path)
if (have_dec) DEC <- readRDS(dec_path)
comp_lv <- c("phytoplankton", "macrophytes", "litter", "hydrology", "inheritance", "unresolved")
comp_lab <- c(phytoplankton = "phytoplankton", macrophytes = "macrophytes", litter = "litter / organic inputs",
              hydrology = "hydrology (residence time)", inheritance = "inheritance shortfall", unresolved = "unresolved (intercept + stream effect)")
comp_pal <- c(phytoplankton = "#3B9AD1", macrophytes = "#41AB5D", litter = "#B5651D", hydrology = "#8856A7", inheritance = "grey40", unresolved = "grey75")
```

```{r fig-decomp}
#| fig-height: 7
#| fig-cap: "**What the modelled pathways contribute to the local modification, in enriching and depleting samplings.** Posterior mean contribution (mg L⁻¹) of each component of @eq-decomp to the predicted local modification, averaged over the samplings classified as source (P(L > 0) > 0.90), sink (P(L < 0) > 0.90) or no-evidence; point: posterior median, bars: 66% and 95% credible intervals. (a) All samplings (mediators imputed where not measured). (b) The 16 streams where the producers were measured. (c) Source-minus-sink difference per component, all samplings; a component whose interval excludes zero is one that distinguishes enriching from depleting streams. Variables defined in @tbl-variables; $L$ is the derived local response of the SCM."
if (have_dec) {
  mk <- function(df) df |> mutate(component = factor(component, levels = rev(comp_lv)), class = factor(class, levels = c("source", "no-evidence", "sink")))
  pl <- function(df, ttl) ggplot(mk(df), aes(x = med, y = component, colour = class)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
    geom_pointrange(aes(xmin = lo, xmax = hi), position = position_dodge(width = .6), size = .35) +
    scale_colour_manual(values = c(source = "#3B9AD1", `no-evidence` = "grey60", sink = "#E8603C"), name = NULL) +
    scale_y_discrete(labels = comp_lab) +
    labs(x = expression("Contribution to the local modification (mg "*L^-1*")"), y = NULL, subtitle = ttl) +
    theme_bw(base_size = 11) + theme(panel.grid.minor = element_blank(), legend.position = "top")
  pa <- pl(DEC$summ_all, sprintf("All samplings (source n = %d, sink n = %d, no-evidence n = %d)", DEC$n_class["source"], DEC$n_class["sink"], DEC$n_class["no-evidence"]))
  pb <- pl(DEC$summ_meas, "The 16 measured streams (summer)")
  pc <- ggplot(DEC$diff_all |> mutate(component = factor(component, levels = rev(comp_lv))), aes(x = med, y = component)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
    geom_pointrange(aes(xmin = lo, xmax = hi), size = .4) + scale_y_discrete(labels = comp_lab) +
    labs(x = expression("Source minus sink (mg "*L^-1*")"), y = NULL, subtitle = "What distinguishes enriching from depleting samplings") +
    theme_bw(base_size = 11) + theme(panel.grid.minor = element_blank())
  fig_dec <- (pa + pb) / pc + plot_layout(heights = c(1.2, 1)) + plot_annotation(tag_levels = "a")
  ggsave(here("results", "figures", "plan", "fig_decomp.png"), fig_dec, width = 11, height = 7, dpi = 300, bg = "white"); fig_dec
} else cat("Decomposition not yet computed (notebook/R/pathway_decomposition.R).")
```

```{r fig-decomp-season}
#| fig-height: 4.5
#| fig-cap: "**The seasonal change of each pathway contribution, by the sign of the beaver seasonal modification B.** For every stream with both seasons, the summer-minus-winter change of each component of @eq-decomp, averaged over the streams whose B (derived from the SCM) is credibly positive, credibly negative, or no-evidence; point: posterior median, bars: 66% and 95% credible intervals. Right: the share of B carried by each component — the posterior slope of the component's seasonal change on B across streams (the shares sum to one)."
if (have_dec) {
  lvB <- c("B > 0 (summer > winter)", "no-evidence", "B < 0 (winter > summer)")
  pa <- ggplot(DEC$seas |> mutate(component = factor(component, levels = rev(comp_lv)), classB = factor(classB, levels = lvB)),
               aes(x = med, y = component, colour = classB)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
    geom_pointrange(aes(xmin = lo, xmax = hi), position = position_dodge(width = .6), size = .35) +
    scale_colour_manual(values = c(`B > 0 (summer > winter)` = "#8856A7", `no-evidence` = "grey60", `B < 0 (winter > summer)` = "#41AB5D"), name = NULL) +
    scale_y_discrete(labels = comp_lab) +
    labs(x = expression("Summer minus winter contribution (mg "*L^-1*")"), y = NULL) +
    theme_bw(base_size = 11) + theme(panel.grid.minor = element_blank(), legend.position = "top")
  pb <- ggplot(DEC$share |> mutate(component = factor(component, levels = rev(comp_lv))), aes(x = med, y = component)) +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
    geom_pointrange(aes(xmin = lo, xmax = hi), size = .4) + scale_y_discrete(labels = comp_lab) +
    labs(x = "Share of B carried by the component", y = NULL) + theme_bw(base_size = 11) + theme(panel.grid.minor = element_blank())
  fig_ds <- pa + pb + plot_layout(widths = c(1.4, 1)) + plot_annotation(tag_levels = "a")
  ggsave(here("results", "figures", "plan", "fig_decomp_season.png"), fig_ds, width = 11, height = 4.5, dpi = 300, bg = "white"); fig_ds
}
```

```{r decomp-numbers}
#| echo: false
if (have_dec) {
  vs <- DEC$vshare |> mutate(txt = sprintf("%s %.0f%% [%.0f–%.0f%%]", comp_lab[component], 100*med, 100*lo, 100*hi))
  df <- DEC$diff_all |> mutate(txt = sprintf("%s %.2f [%.2f, %.2f]", comp_lab[component], med, lo, hi))
  sh <- DEC$share |> mutate(txt = sprintf("%s %.0f%% [%.0f–%.0f%%]", comp_lab[component], 100*med, 100*lo, 100*hi))
  top_v <- vs |> arrange(desc(med)) |> slice(1); top_d <- df |> filter(component != "unresolved") |> arrange(desc(abs(med))) |> slice(1)
}
```

::: {.callout-important}
## Key message of @fig-decomp and @fig-decomp-season

`r if (have_dec) sprintf("- **Where the variation in the local modification comes from (share of variance across samplings):** %s.", paste(vs$txt, collapse = "; ")) else "- pending"`
`r if (have_dec) sprintf("- **What distinguishes sources from sinks (source minus sink, mg L⁻¹):** %s.", paste(df$txt, collapse = "; ")) else ""`
`r if (have_dec) sprintf("- **What carries the seasonal change B (share of B):** %s; %d samplings-pairs with both seasons in the SCM data (m6), of which %d have B credibly positive and %d credibly negative.", paste(sh$txt, collapse = "; "), DEC$n_pairs, DEC$n_classB["B > 0 (summer > winter)"], DEC$n_classB["B < 0 (winter > summer)"]) else ""`
`r if (have_dec) { gd <- function(k) df$med[df$component == k]; gs <- function(k) 100*sh$med[sh$component == k]; sprintf("- **Implication: the sign is set first by what comes in, then by phytoplankton.** The component that separates sources from sinks is the inheritance shortfall (%.2f mg L⁻¹ more negative in sinks): the streams that deplete are the ones receiving the most DOC, where a 5%% shortfall on a large inherited concentration outweighs the biological additions. Phytoplankton is the only measured pathway that also separates them (%.2f mg L⁻¹, credible), and at the 16 measured streams its contribution is clearly positive in sources and negative in sinks (@fig-decomp b). Macrophytes, litter and residence time do not distinguish the two.", gd("inheritance"), gd("phytoplankton")) } else ""`
`r if (have_dec) sprintf("- **Implication for the seasons.** The summer-minus-winter change B is carried by the producers (phytoplankton %.0f%%, macrophytes %.0f%%) and by the seasonal change of the inherited state itself (%.0f%%): streams whose B is positive gain phytoplankton and macrophyte contributions in summer, streams whose B is negative lose through the inheritance term (their upstream DOC rises more than the producers can add). The unresolved component carries none of the seasonal change **by construction** — the SCM has one stream effect for both seasons — so seasonal switching is forced onto the measured and imputed pathways; a season-specific stream effect would be needed to test whether unmeasured removal also changes with season.", gs("phytoplankton"), gs("macrophytes"), gs("inheritance")) else ""`
`r if (have_dec) sprintf("- **What can and cannot be claimed.** With the producers imputed at all but 16 streams, the phytoplankton and macrophyte shares (%.0f%% and %.0f%% of the variance) are model-implied outside those 16; the inheritance result rests on all 158 streams and is solid. The decomposition supports the framing — an inherited signal, locally modified, with enrichment where phytoplankton adds to a modest inflow and depletion where a rich inflow is not matched by production — but the removal side (mineralization, photodegradation) is unmeasured and appears here only as the 5%% shortfall, which is why it must be described as unresolved net removal rather than attributed.", 100*vs$med[vs$component == "phytoplankton"], 100*vs$med[vs$component == "macrophytes"]) else ""`
:::

# 11. Methods figure

**Manuscript placement: main text, Figure 3 (both panels redrawn: panel a shows the 158 analysed streams, panel b shows $S$, $\Delta\text{DOC}_s$, $R_s$ and $F$ in place of the amplification-factor quantities).**

The chunk below draws panel (a) directly from the analysed data — the red dots are exactly the 158 streams, so map and analysis can never disagree — plus the 16 producer-abundance streams (black triangles; coordinates in `data/processed/producer_sites_coordinates.csv`, taken from the field-campaign site list). Panel (b) is the schematic of the derived variables.

```{r fig-map}
#| fig-height: 11
#| fig-cap: "**Sampling design and derived variables.** (a) Sampling locations across Switzerland (elevation background): red dots are the 158 analysed streams; black triangles are the 16 streams where primary producer abundance was sampled. Water was sampled upstream of the first and downstream of the last beaver dam of each stream, once in winter and once in summer. (b) Schematic of the derived variables: the inherited seasonal contrast $S$ is generated over the floodplain area $A_\\text{floodplain}$; the local response $\\Delta\\text{DOC}_s$ is measured across the beaver-engineered stream of ponded area $A_\\text{beaver}$; the relative response is $R_s = \\Delta\\text{DOC}_s / \\lvert S \\rvert$ and the footprint contrast is $F = A_\\text{floodplain} / A_\\text{beaver}$. All variables are defined in @tbl-variables."
# panel (a): the 158 analysed streams (d, from the Q1 chunk) with coordinates,
# and the 16 producer-abundance streams from the field-campaign site list
pts  <- d |> distinct(site) |>
  left_join(read.csv(here("data", "processed", "m6.csv")) |> distinct(site, .keep_all = TRUE) |>
              select(site, lon = x_upstream, lat = y_upstream), by = "site")
prod <- read.csv(here("data", "processed", "producer_sites_coordinates.csv"))
# elevation background: Swiss DEM (30 arc-second, data/elevation/) with a
# hillshade underlay and a hypsometric tint, as in the original map
library(terra); library(tidyterra)
dem    <- rast(here("data", "elevation", "CHE_elv_msk.tif"))
hill   <- shade(terrain(dem, "slope", unit = "radians"),
                terrain(dem, "aspect", unit = "radians"), 40, 270)
greys  <- hcl.colors(1000, "Grays")
hillcol <- greys[round(scales::rescale(values(hill), to = c(1, 1000)))]
pmap <- ggplot() +
  geom_spatraster(data = hill, fill = hillcol, maxcell = Inf) +
  geom_spatraster(data = dem, maxcell = Inf) +
  scale_fill_hypso_tint_c(palette = "dem_poster", alpha = 0.5, na.value = NA, name = "Elevation (m)") +
  geom_point(data = pts, aes(lon, lat), colour = "red", size = 1.7, alpha = 0.9) +
  geom_point(data = prod, aes(longitude, latitude), colour = "black", size = 2.2, shape = 17) +
  annotate("text", x = 5.99, y = 47.75, label = "(a)", size = 6.5, hjust = 0) +
  theme_void() + theme(legend.position = "right")
# panel (b): the schematic, included as an image below the map
schem <- png::readPNG(here("results", "figures", "plan", "schematic_S_R_F.png"))
psch  <- ggplot() + annotation_custom(grid::rasterGrob(schem)) +
  annotate("text", x = 0.02, y = 0.97, label = "(b)", size = 6.5, hjust = 0) +
  xlim(0, 1) + ylim(0, 1) + theme_void()
fig3 <- pmap / psch + plot_layout(heights = c(1, 1.15))
ggsave(here("results", "figures", "plan", "fig_map_schematic.png"), fig3, width = 9, height = 11, dpi = 300, bg = "white")
fig3
```

# 12. Figures for the manuscript

```{r inventory}
#| echo: false
kable(tribble(
  ~`Manuscript figure`, ~Content, ~Answers, ~`This document`, ~Status,
  "Figure 1a", "Posterior predictive distribution of $\\Delta\\text{DOC}_s$ by season", "Q1", "@fig-q1", "existing",
  "Figure 2", "Direction figure: now single-panel (panel b, the SCM, promoted to its own Figure 4)", "Q1", "@fig-q1", "**panel b removed**",
  "Figure 4", "Structural causal model, outcome downstream DOC, with estimated effects (new causal links, upstream measurement error)", "Q2", "@fig-q2", "**new — standalone, was Figure 2b**",
  "Figure 2",  "Relative response $R_s$: across streams (a), by stream (b); seasonal gain $G$ by stream (c)", "Q3", "@fig-R", "**new — replaces the amplification-factor figure**",
  "Figure 3",  "Map (158 analysed streams), sampling design and schematic of $S$, $\\Delta\\text{DOC}_s$, $R_s$, $F$", "Methods", "@fig-map", "**both panels redrawn**",
  "Figure S1–S9", "Hypothesised DAG, priors, checks, stream-level classification, total effects, connectivity model", "Q2 support", "—", "existing",
  "Figure S10", "Derivation of the variables", "all", "@fig-dag", "**new**",
  "Figure S11", "Spatial coherence of $S$, $B$ and $T$", "Discussion support", "@fig-spatial", "**new**",
  "Figure S12", "$R_{summer}$ and $G$ against the number of dams and the water residence time", "Q4", "@fig-intensity", "**new**",
  "Figure S13", "Hold-out prediction of downstream DOC and $\\Delta\\text{DOC}$", "Methods support", "@fig-holdout", "**new**",
  "Figure S18", "Pathway decomposition of the local modification: sources vs sinks, and the seasonal change", "Q5", "@fig-decomp, @fig-decomp-season", "**new**",
  "Figures S10–S15, Table S2", "Structural and causal sensitivity: alternative DAGs, unmeasured confounding, falsification tests, prior and spatial checks, causal knowledge analysis", "Q2 support", "@fig-dag-tests, @fig-dag-compare, @fig-sensemakr, @fig-negative-controls, @fig-prior-sens, @fig-spatial-resid, @tbl-cka", "**new**"))
```

::: {.callout-warning}
## What changes in the manuscript

1.  **Abstract.** The sentence *"Compared to floodplain-scale DOC variation, beaver-engineering amplified DOC concentration changes 141 times in summer and 33 times in winter."* is replaced by: *"Relative to the inherited seasonal contrast of upstream DOC, beaver-engineering added DOC equal to `r sprintf("%.0f%%", 100*median(medR(R_s)))` of that contrast in summer (95% credible interval `r sprintf("%.0f–%.0f%%", 100*quantile(medR(R_s), .025), 100*quantile(medR(R_s), .975))`) and had no net effect in winter, on a footprint of about `r sprintf("%.1f%%", 100 * median(1 / d$F))` of the floodplain area."*
2.  **Results.** The amplification-factor section and its figure are removed; Q3 is answered by $R_s$ (Figure 2a, b) and by the seasonal gain $G$ (Figure 2c: share of streams that reinforce / dampen / reverse).
3.  **Methods.** The definitions of $S$, $B$, $G$, $R_s$ and $F$ and the measurement-error model on the concentrations replace the amplification-factor equations; one sentence states that the structural causal model is parameterized with downstream DOC as the outcome (inheritance slope 0.95) and $\Delta\text{DOC}_s$ is derived from the posterior; Figure 3b is redrawn.
4.  **Discussion.** One sentence on spatial coherence: the inherited seasonal contrast is regional, the beaver seasonal modification is local (Figure S11).
5.  **Whole manuscript.** Every mention of 176 streams is replaced by **158**, the streams actually analysed (abstract, introduction, methods, Figure 3a and its caption); the count appears with the same value everywhere.
6.  **Appendix S1.** Figure S10 (derivation of the variables), Figure S11 (spatial coherence), Figure S12 (response against intensity of beaver-engineering) and Figure S13 (hold-out validation) are added.
7.  **Methods and Discussion.** One sentence each on validation: the SCM predicts downstream DOC at unseen streams (hold-out R², coverage), but not the sign of the local change; the "predictive basis" claim of the Discussion is reworded accordingly (section 8).
:::

# 13. Reproducibility

```{r session}
#| echo: false
cat("Data: data/processed/m12.csv | concentration posteriors: notebook/concentrations_fit.qmd -> results/concentrations.rds (thinned copy concentrations_thin.rds is tracked)\n")
sessionInfo()$R.version$version.string
```

# References

::: {#refs}
:::
